---
title: "R Functions Reference"
---

## Overview

The R source files in `R/` provide data loading, schema management, caching,
and plot theming for the analysis notebooks.

| File | Purpose |
|------|---------|
| `data_loading.R` | All data loading functions with caching and validation |
| `schemas.R` | Arrow schema registry, factor levels, validation rules |
| `cache.R` | Parquet caching system with zstd compression |
| `plot_themes.R` | Plot theming and visualization helpers |

---

## data_loading.R

```{r}
#| echo: false
#| output: asis
cat("```r\n")
cat(readLines(here::here("R", "data_loading.R")), sep = "\n")
cat("\n```\n")
```

---

## schemas.R

```{r}
#| echo: false
#| output: asis
cat("```r\n")
cat(readLines(here::here("R", "schemas.R")), sep = "\n")
cat("\n```\n")
```

---

## cache.R

```{r}
#| echo: false
#| output: asis
cat("```r\n")
cat(readLines(here::here("R", "cache.R")), sep = "\n")
cat("\n```\n")
```

---

## plot_themes.R

```{r}
#| echo: false
#| output: asis
cat("```r\n")
cat(readLines(here::here("R", "plot_themes.R")), sep = "\n")
cat("\n```\n")
```
