---
title: "v5q Benchmark Set Variant and Region Characterization"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(
  load_plot_themes = TRUE,
  load_sessioninfo = TRUE
)
```

## Data Loading

This analysis uses **primary aggregated metrics** rather than loading full variant tables directly. This approach is faster, more memory-efficient, and sufficient for most analyses. See the [Pipeline Outputs Reference](../docs/pipeline-outputs.md) for details on when to load detailed variant tables.

### Primary Metrics

```{r load_primary}
analysis_data <- load_primary_analysis_data(
  include_variants = TRUE,
  include_reference_sizes = TRUE,
  include_hg002q100_size = TRUE,
  include_benchmark_regions = TRUE
)
list2env(analysis_data, envir = environment())

## Filtering variants_df to benchmark regions for the characterization
variants_df <- variants_df %>%
  filter(region_ids == "BMKREGIONS", is_pass)
```

## Benchmark Set Summary Table

Consolidated summary of variant counts and benchmark region sizes across
benchmark versions. GRCh38 is the primary reference for the manuscript;
GRCh37 and CHM13v2.0 are included in supplemental tables.

```{r benchmark_summary_data}
## Compute benchmark region sizes
bench_region_stats_df <- bench_regions_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1", "v0.6"),
    chrom %in% CHROM_LEVELS
  ) %>%
  group_by(bench_type, ref, bench_version) %>%
  summarise(total_bp = sum(interval_size), .groups = "drop") %>%
  left_join(total_ref_asm_bp, by = "ref") %>%
  mutate(pct_ref_cvg = 100 * (total_bp / total_asm_bp)) %>%
  select(-total_asm_bp)

## Compute variant counts by type
bench_var_cnt_df <- variants_df %>%
  count(bench_type, ref, bench_version, var_type)

## Build unified summary: variant counts + region stats
bench_summary_df <- bench_var_cnt_df %>%
  pivot_wider(names_from = var_type, values_from = n, values_fill = 0L) %>%
  mutate(Total = rowSums(across(any_of(c("SNV", "INDEL", "INS", "DEL"))))) %>%
  full_join(bench_region_stats_df, by = c("bench_type", "ref", "bench_version")) %>%
  mutate(
    ref = std_references(ref),
    bench_version = std_bench_versions(bench_version),
    bench_mb = round(total_bp / 1e6, 0)
  ) %>%
  arrange(bench_type, ref, bench_version)
```

```{r format_bench_table_fn}
## Shared formatting function for benchmark summary tables
format_bench_table <- function(df, var_cols) {
  df %>%
    select(ref, bench_version, all_of(var_cols), Total, bench_mb, pct_ref_cvg) %>%
    as_flextable(show_coltype = FALSE) %>%
    fmt_integer_flextable(columns = c(var_cols, "Total", "bench_mb")) %>%
    fmt_number_flextable(columns = "pct_ref_cvg", decimals = 1) %>%
    cols_label_flextable(
      ref = "Reference",
      bench_version = "Version",
      Total = "Total Variants",
      bench_mb = "Bench (Mb)",
      pct_ref_cvg = "% Ref"
    ) %>%
    flextable::align(j = "ref", align = "right", part = "body") %>%
    theme_flextable_manuscript()
}
```

### Main Text: Summary (Table 1)

Small variants on GRCh38 (v5.0q vs v4.2.1). Structural variants on GRCh37
(v5.0q vs v0.6) since v0.6 was only generated for GRCh37.

```{r table1_smvar}
bench_summary_df %>%
  filter(bench_type == "smvar", ref == "GRCh38") %>%
  format_bench_table(var_cols = c("SNV", "INDEL"))
```

```{r table1_stvar}
bench_summary_df %>%
  filter(bench_type == "stvar", ref == "GRCh37") %>%
  format_bench_table(var_cols = c("DEL", "INS"))
```

### Supplemental: All References

```{r table_s_smvar}
bench_summary_df %>%
  filter(bench_type == "smvar") %>%
  format_bench_table(var_cols = c("SNV", "INDEL"))
```

```{r table_s_stvar}
bench_summary_df %>%
  filter(bench_type == "stvar") %>%
  format_bench_table(var_cols = c("DEL", "INS"))
```


## Variant Size Distributions

Size distributions for v5.0q compared to legacy benchmarks (GRCh38 only).
Overlaid histograms show the expansion in variant content.

```{r indel_length_distribution}
#| message: false
#| warning: false
smvar_len_df <- variants_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1"),
    bench_type == "smvar",
    var_type == "INDEL",
    ref == "GRCh38"
  ) %>%
  select(bench_version, var_size) %>%
  mutate(var_size = if_else(var_size > 0, var_size + 0.5, var_size - 0.5))

(smvar_len_dist_plt <- ggplot(smvar_len_df, aes(x = var_size, fill = bench_version)) +
  geom_histogram(
    binwidth = 1,
    position = "identity",
    alpha = 0.7,
    color = NA
  ) +
  scale_fill_manual(
    values = c("v5.0q" = get_color_palettes()$bench_version["v5.0q"],
               "v4.2.1" = get_color_palettes()$bench_version["v4.2.1"])
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "INDEL Length (bp)", y = "Count", fill = "Version") +
  theme_manuscript() +
  theme(legend.position = "none"))
```

```{r sv_length_distribution}
#| message: false
## Use GRCh37 for SV comparison since v0.6 is GRCh37-only
stvar_len_df <- variants_df %>%
  filter(
    bench_version %in% c("v5.0q", "v0.6"),
    bench_type == "stvar",
    ref == "GRCh37"
  ) %>%
  select(bench_version, var_size) %>%
  filter(abs(var_size) < 20000)

(stvar_len_dist_plt <- ggplot(stvar_len_df, aes(x = var_size, fill = bench_version)) +
  geom_histogram(
    bins = 50,
    position = "identity",
    alpha = 0.7,
    color = NA
  ) +
  scale_fill_manual(
    values = c("v5.0q" = get_color_palettes()$bench_version["v5.0q"],
               "v0.6" = get_color_palettes()$bench_version["v0.6"])
  ) +
  scale_x_continuous(
    breaks = c(-10000, -5000, -1000, -500, -100, -50,
               0, 50, 100, 500, 1000, 5000, 10000),
    trans = scales::pseudo_log_trans(sigma = 100, base = 10),
    labels = scales::comma
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "SV Length (bp)", y = "Count", fill = "Version") +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0),
    legend.position = "none"
  ))
```

```{r combined_var_len_dist}
(combined_var_len_dist_plt <- smvar_len_dist_plt + stvar_len_dist_plt +
  plot_layout(ncol = 2, guides = "collect", widths = c(1, 2)) +
  plot_annotation(tag_levels = "A") &
  theme_manuscript() +
    theme(
      axis.text.x = element_text(angle = -45, hjust = 0),
      legend.position = "bottom",
      plot.margin = unit(c(5, 5, 5, 5), "pt")
    ))
```

```{r var_length_dist_save}
ggsave(
  filename = here("manuscript", "figs", "var_len_dist.png"),
  plot = combined_var_len_dist_plt,
  width = 8,
  height = 4,
  dpi = 300
)
```


## Benchmark Region Coverage Change by Chromosome

Relative change in benchmark region size (v5.0q vs legacy) per autosome.
Small variants use GRCh38 (v5.0q vs v4.2.1); structural variants use
GRCh37 (v5.0q vs v0.6) since v0.6 is GRCh37-only.

```{r coverage_change_plot, fig.width = 8, fig.height = 4}
coverage_change_df <- bench_regions_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1", "v0.6"),
    (bench_type == "smvar" & ref == "GRCh38") |
      (bench_type == "stvar" & ref == "GRCh37"),
    chrom %in% AUTOSOME_CHROM_LEVELS
  ) %>%
  group_by(bench_type, chrom, bench_version) %>%
  summarise(total_size = sum(interval_size), .groups = "drop") %>%
  mutate(
    vers = if_else(bench_version == "v5.0q", "new", "old"),
    chrom = factor(chrom, levels = AUTOSOME_CHROM_LEVELS)
  ) %>%
  select(-bench_version) %>%
  pivot_wider(names_from = vers, values_from = total_size) %>%
  mutate(
    pct_change = 100 * (new - old) / old,
    total_change = new - old
  )

(coverage_change_plt <- coverage_change_df %>%
  ggplot() +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(
    aes(
      x = chrom,
      y = pct_change,
      fill = total_change / 1e6,
      shape = bench_type
    ),
    size = 3,
    color = "black",
    position = position_jitterdodge(jitter.width = 0.1)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    n.breaks = 8
  ) +
  scale_fill_gradient2(
    low = "red", mid = "white", high = "blue",
    midpoint = 0, labels = scales::comma
  ) +
  scale_shape_manual(
    values = c("smvar" = 21, "stvar" = 24),
    labels = c("smvar" = "Small Variants", "stvar" = "Structural Variants")
  ) +
  labs(
    x = "Chromosome",
    y = "Coverage Change (v5.0q vs legacy)",
    fill = "Change (Mb)",
    shape = "Benchmark Type"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0),
    legend.position = "bottom"
  ))

ggsave(
  filename = here("manuscript", "figs", "benchmark_region_size_change.png"),
  plot = coverage_change_plt,
  width = 8,
  height = 4,
  dpi = 300
)
```
