---
title: "v5q Benchmark Set Exclusion Regions"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(
  load_plot_themes = TRUE
)

library(ggupset)
library(glue)
```

```{r}
# Map pipeline exclusion names to manuscript table names
exclusion_name_map <- c(
  "segdups" = "Segmental Duplications",
  "tandem-repeats" = "Tandem Repeats",
  "satellites" = "Satellites",
  "pav-inversions" = "PAV-inversions",
  "flanks" = "Flanks",
  "VDJ" = "VDJ",
  "HG002Q100-errors" = "HG002Q100v1.1 errors",
  "TSPY2-segdups" = "TSPY2-segdups",
  "consecutive-svs" = "Consecutive SVs",
  "gaps" = "Gaps",
  "dipcall-bugs-T2TACE" = "Dipcall-bugs T2TACE",
  "svs-and-simple-repeats" = "SV regions",
  "self-discrep" = "Self-Discrepancies",
  "HG002-mosaic" = "HG002-mosaic",
  "dipcall-pav_discrep-smvar" = "dipcall-PAV Discrep. (smvar)",
  "dipcall-pav_discrep-stvar" = "dipcall-PAV Discrep. (stvar)"
)

rename_exclusion <- function(x) {
  dplyr::if_else(x %in% names(exclusion_name_map), exclusion_name_map[x], x)
}

# Rename pipe-separated exclusion combinations
rename_exclusion_combination <- function(x) {
  vapply(x, function(combo) {
    parts <- strsplit(combo, "\\|")[[1]]
    paste(rename_exclusion(parts), collapse = "|")
  }, character(1), USE.NAMES = FALSE)
}
```

## Data Loading

```{r}
exclusion_metrics_tbl <- load_exclusion_metrics() %>%
  dplyr::mutate(exclusions = rename_exclusion(exclusions))

if (nrow(exclusion_metrics_tbl) == 0) {
  stop("No v5.0q exclusion metrics found in results/exclusions.")
}
```

```{r}
exclusion_interaction_tbl <- load_exclusion_interactions() %>%
  dplyr::mutate(
    exclusion_combination = rename_exclusion_combination(exclusion_combination)
  )

if (nrow(exclusion_interaction_tbl) == 0) {
  stop("No v5.0q exclusion interaction data found in results/exclusions.")
}
```


## Bases Removed by Exclusion

### GRCh38 (Main)

Exclusion names in *italics* apply only to the small variant benchmark.

```{r}
grch38_exclusion_tbl <- exclusion_metrics_tbl %>%
  dplyr::filter(ref == "GRCh38") %>%
  dplyr::select(exclusions, bench_type, intersect_bp, pct_of_dip) %>%
  dplyr::group_by(exclusions) %>%
  dplyr::mutate(
    is_smvar_only = all(bench_type == "smvar")
  ) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(exclusions, .keep_all = TRUE) %>%
  dplyr::arrange(desc(intersect_bp)) %>%
  dplyr::select(exclusions, intersect_bp, pct_of_dip, is_smvar_only)

# Identify rows to italicize
smvar_only_rows <- which(grch38_exclusion_tbl$is_smvar_only)

grch38_exclusion_tbl %>%
  dplyr::select(-is_smvar_only) %>%
  flextable::flextable() %>%
  fmt_integer_flextable(columns = "intersect_bp") %>%
  fmt_number_flextable(columns = "pct_of_dip", decimals = 3) %>%
  flextable::align(
    j = c("intersect_bp", "pct_of_dip"), align = "right", part = "all"
  ) %>%
  flextable::italic(i = smvar_only_rows, j = "exclusions", part = "body") %>%
  flextable::color(
    i = smvar_only_rows, j = "exclusions", color = "#555555", part = "body"
  ) %>%
  cols_label_flextable(
    exclusions = "Exclusion",
    intersect_bp = "Bases Excluded",
    pct_of_dip = "% of Dip.bed"
  ) %>%
  theme_flextable_manuscript()
```

### All References (Supplemental)

Exclusion names in *italics* apply only to the small variant benchmark.

```{r}
supp_exclusion_tbl <- exclusion_metrics_tbl %>%
  dplyr::select(ref, exclusions, bench_type, intersect_bp, pct_of_dip) %>%
  dplyr::group_by(exclusions) %>%
  dplyr::mutate(
    is_smvar_only = all(bench_type == "smvar")
  ) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(exclusions, ref, .keep_all = TRUE) %>%
  dplyr::select(-bench_type) %>%
  tidyr::pivot_wider(
    names_from = ref,
    values_from = c(intersect_bp, pct_of_dip)
  ) %>%
  dplyr::arrange(desc(intersect_bp_GRCh38))

supp_smvar_only_rows <- which(supp_exclusion_tbl$is_smvar_only)

bp_cols <- c("intersect_bp_GRCh37", "intersect_bp_GRCh38", "intersect_bp_CHM13v2.0")
pct_cols <- c("pct_of_dip_GRCh37", "pct_of_dip_GRCh38", "pct_of_dip_CHM13v2.0")

supp_exclusion_tbl %>%
  dplyr::select(-is_smvar_only) %>%
  flextable::flextable() %>%
  fmt_integer_flextable(columns = bp_cols) %>%
  fmt_number_flextable(columns = pct_cols, decimals = 3) %>%
  flextable::align(j = c(bp_cols, pct_cols), align = "right", part = "all") %>%
  flextable::italic(
    i = supp_smvar_only_rows, j = "exclusions", part = "body"
  ) %>%
  flextable::color(
    i = supp_smvar_only_rows, j = "exclusions", color = "#555555", part = "body"
  ) %>%
  cols_label_flextable(
    exclusions = "Exclusion",
    intersect_bp_GRCh37 = "Bases",
    intersect_bp_GRCh38 = "Bases",
    intersect_bp_CHM13v2.0 = "Bases",
    pct_of_dip_GRCh37 = "% Dip",
    pct_of_dip_GRCh38 = "% Dip",
    pct_of_dip_CHM13v2.0 = "% Dip"
  ) %>%
  flextable::add_header_row(
    values = c("", "GRCh37", "GRCh38", "CHM13v2.0", "GRCh37", "GRCh38", "CHM13v2.0"),
    top = TRUE
  ) %>%
  flextable::merge_h(part = "header") %>%
  theme_flextable_manuscript()
```


## Exclusion Interactions

Exclusion regions can overlap — a single genomic region may be excluded by multiple
criteria simultaneously. The upset plots below show how much sequence is removed by each
exclusion combination, highlighting where multiple exclusions co-occur.

```{r}
# Compute summary metrics per ref/bench_type
interaction_summary_tbl <- exclusion_interaction_tbl %>%
  dplyr::group_by(ref, bench_type) %>%
  dplyr::mutate(
    total_bases = sum(bases_bp),
    total_variants = sum(variant_count),
    pct_bases = 100 * bases_bp / total_bases,
    pct_variants = 100 * variant_count / total_variants
  ) %>%
  dplyr::ungroup()
```

```{r}
bench_type_labels <- get_bench_type_labels()

# Internal separator used by ggupset for combination matrix labels
UPSET_SEP <- "-__-__-_-"

# Upset plot for excluded bases.
# Uses scale_x_discrete + axis_combmatrix instead of scale_x_upset
# to control bar ordering by bases_bp (scale_x_upset orders by row
# frequency which doesn't work with stat="identity").
# Exclusions to drop from intersection plots (no overlaps with others)
EXCLUDE_FROM_UPSET <- c("VDJ")

create_base_upset <- function(data, ref_name, bench_type_name,
                              n_top = 15, title_suffix = "") {
  plot_data <- data %>%
    dplyr::filter(ref == ref_name, bench_type == bench_type_name) %>%
    # Drop combinations that contain excluded exclusions
    dplyr::filter(!grepl(
      paste(EXCLUDE_FROM_UPSET, collapse = "|"),
      exclusion_combination
    )) %>%
    dplyr::arrange(desc(bases_bp)) %>%
    dplyr::slice_head(n = n_top) %>%
    dplyr::mutate(
      exclusions_list = strsplit(exclusion_combination, "\\|"),
      # Collapse to string with ggupset's internal separator
      combo_str = vapply(exclusions_list, function(x) {
        paste(sort(x), collapse = UPSET_SEP)
      }, character(1))
    ) %>%
    dplyr::ungroup()

  if (nrow(plot_data) == 0) {
    return(NULL)
  }

  # Factor levels ordered by descending bases_bp
  ordered_levels <- plot_data %>%
    dplyr::arrange(desc(bases_bp)) %>%
    dplyr::pull(combo_str)

  # Order y-axis exclusion labels by bases from single-exclusion rows
  single_exclusion_bases <- plot_data %>%
    dplyr::filter(lengths(exclusions_list) == 1) %>%
    dplyr::mutate(excl_name = vapply(exclusions_list, `[`, character(1), 1)) %>%
    dplyr::arrange(desc(bases_bp)) %>%
    dplyr::pull(excl_name)

  # Add any exclusions that only appear in multi-combos (not as singles)
  all_excl_names <- unique(unlist(plot_data$exclusions_list))
  remaining <- setdiff(all_excl_names, single_exclusion_bases)
  set_levels <- c(single_exclusion_bases, remaining)

  bt_label <- bench_type_labels[bench_type_name]

  ggplot(plot_data, aes(x = factor(combo_str, levels = ordered_levels), y = bases_bp)) +
    geom_bar(stat = "identity", fill = "#2171b5") +
    scale_x_discrete(limits = ordered_levels) +
    ggupset::axis_combmatrix(sep = UPSET_SEP, levels = set_levels) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = glue::glue("{bt_label}: {ref_name}{title_suffix}"),
      x = NULL,
      y = "Bases (bp)"
    ) +
    theme_manuscript() +
    theme(
      panel.grid.major.x = element_blank()
    )
}
```


```{r}
# Upset plot for excluded variants (same structure as bases)
create_variant_upset <- function(data, ref_name, bench_type_name,
                                 n_top = 15, title_suffix = "") {
  plot_data <- data %>%
    dplyr::filter(ref == ref_name, bench_type == bench_type_name) %>%
    dplyr::filter(!grepl(
      paste(EXCLUDE_FROM_UPSET, collapse = "|"),
      exclusion_combination
    )) %>%
    dplyr::arrange(desc(variant_count)) %>%
    dplyr::slice_head(n = n_top) %>%
    dplyr::mutate(
      exclusions_list = strsplit(exclusion_combination, "\\|"),
      combo_str = vapply(exclusions_list, function(x) {
        paste(sort(x), collapse = UPSET_SEP)
      }, character(1))
    ) %>%
    dplyr::ungroup()

  if (nrow(plot_data) == 0) {
    return(NULL)
  }

  ordered_levels <- plot_data %>%
    dplyr::arrange(desc(variant_count)) %>%
    dplyr::pull(combo_str)

  single_exclusion_vars <- plot_data %>%
    dplyr::filter(lengths(exclusions_list) == 1) %>%
    dplyr::mutate(excl_name = vapply(exclusions_list, `[`, character(1), 1)) %>%
    dplyr::arrange(desc(variant_count)) %>%
    dplyr::pull(excl_name)

  all_excl_names <- unique(unlist(plot_data$exclusions_list))
  remaining <- setdiff(all_excl_names, single_exclusion_vars)
  set_levels <- c(single_exclusion_vars, remaining)

  bt_label <- bench_type_labels[bench_type_name]

  ggplot(plot_data, aes(x = factor(combo_str, levels = ordered_levels), y = variant_count)) +
    geom_bar(stat = "identity", fill = "#d94801") +
    scale_x_discrete(limits = ordered_levels) +
    ggupset::axis_combmatrix(sep = UPSET_SEP, levels = set_levels) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = glue::glue("{bt_label}: {ref_name}{title_suffix}"),
      x = NULL,
      y = "Variant Count"
    ) +
    theme_manuscript() +
    theme(
      panel.grid.major.x = element_blank()
    )
}
```


### GRCh38 Bases Excluded (Main Figure)

```{r grch38-upset}
#| fig-height: 8
#| fig-width: 12
#| fig-cap: "Bases excluded by each exclusion combination for GRCh38 small variant (top) and structural variant (bottom) benchmarks. Bars are ordered by total bases excluded."

p_bases_smvar <- create_base_upset(interaction_summary_tbl, "GRCh38", "smvar")
p_bases_stvar <- create_base_upset(interaction_summary_tbl, "GRCh38", "stvar")

(grch38_upset_plt <- p_bases_smvar / p_bases_stvar +
  plot_annotation(tag_levels = "A"))
```

```{r}
ggsave(
  filename = here::here("manuscript", "figs", "grch38_exclusion_upset.png"),
  plot = grch38_upset_plt,
  width = 12,
  height = 8,
  dpi = 300
)
```

### All References Exclusion Patterns (Supplemental)

```{r all-refs-upset}
#| fig-height: 8
#| fig-width: 12
#| results: asis

refs <- c("GRCh37", "GRCh38", "CHM13v2.0")

for (ref in refs) {
  p_smvar <- create_base_upset(interaction_summary_tbl, ref, "smvar")
  p_stvar <- create_base_upset(interaction_summary_tbl, ref, "stvar")

  if (!is.null(p_smvar) || !is.null(p_stvar)) {
    combined <- p_smvar / p_stvar +
      plot_annotation(
        title = glue::glue("{ref} Exclusion Patterns"),
        tag_levels = "A",
        theme = theme(
          plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
        )
      )
    print(combined)
    cat("\n\n---\n\n")
  }
}
```


## Variants in Excluded Regions

### GRCh38 Variant Exclusion Patterns

```{r grch38-variant-upset}
#| fig-height: 8
#| fig-width: 12
#| fig-cap: "Variants excluded by each exclusion combination for GRCh38 small variant (top) and structural variant (bottom) benchmarks. Bars are ordered by variant count."

p_vars_smvar <- create_variant_upset(interaction_summary_tbl, "GRCh38", "smvar")
p_vars_stvar <- create_variant_upset(interaction_summary_tbl, "GRCh38", "stvar")

p_vars_smvar / p_vars_stvar +
  plot_annotation(tag_levels = "A")
```

### All References Variant Exclusion Patterns

```{r all-refs-variant-upset}
#| fig-height: 8
#| fig-width: 12
#| results: asis

for (ref in refs) {
  p_smvar <- create_variant_upset(interaction_summary_tbl, ref, "smvar")
  p_stvar <- create_variant_upset(interaction_summary_tbl, ref, "stvar")

  if (!is.null(p_smvar) || !is.null(p_stvar)) {
    combined <- p_smvar / p_stvar +
      plot_annotation(
        title = glue::glue("{ref} Variant Exclusion Patterns"),
        tag_levels = "A",
        theme = theme(
          plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
        )
      )
    print(combined)
    cat("\n\n---\n\n")
  }
}
```

### Excluded Variant Counts by Exclusion

```{r}
# Single-exclusion variant counts for GRCh38
interaction_summary_tbl %>%
  dplyr::filter(n_exclusions == 1, ref == "GRCh38") %>%
  dplyr::select(exclusion_combination, bench_type, variant_count, bases_bp) %>%
  dplyr::mutate(
    bench_type = bench_type_labels[bench_type],
    variants_per_mb = round(variant_count / (bases_bp / 1e6), 1)
  ) %>%
  dplyr::arrange(desc(variant_count)) %>%
  flextable::flextable() %>%
  fmt_integer_flextable(columns = c("variant_count", "bases_bp")) %>%
  fmt_number_flextable(columns = "variants_per_mb", decimals = 1) %>%
  flextable::align(
    j = c("variant_count", "bases_bp", "variants_per_mb"),
    align = "right", part = "all"
  ) %>%
  cols_label_flextable(
    exclusion_combination = "Exclusion",
    bench_type = "Type",
    variant_count = "Variants",
    bases_bp = "Bases",
    variants_per_mb = "Variants/Mb"
  ) %>%
  theme_flextable_manuscript()
```

### Excluded Variant Counts — All References

```{r}
interaction_summary_tbl %>%
  dplyr::filter(n_exclusions == 1) %>%
  dplyr::select(ref, exclusion_combination, bench_type, variant_count) %>%
  dplyr::mutate(bench_type = bench_type_labels[bench_type]) %>%
  tidyr::pivot_wider(
    names_from = ref,
    values_from = variant_count,
    values_fill = 0L
  ) %>%
  dplyr::arrange(desc(GRCh38)) %>%
  flextable::flextable() %>%
  fmt_integer_flextable(columns = c("GRCh37", "GRCh38", "CHM13v2.0")) %>%
  flextable::align(
    j = c("GRCh37", "GRCh38", "CHM13v2.0"),
    align = "right", part = "all"
  ) %>%
  cols_label_flextable(
    exclusion_combination = "Exclusion",
    bench_type = "Type"
  ) %>%
  theme_flextable_manuscript()
```


### Summary Statistics

```{r}
interaction_summary_tbl %>%
  dplyr::group_by(ref, bench_type) %>%
  dplyr::summarise(
    total_combinations = dplyr::n(),
    single_exclusions = sum(n_exclusions == 1),
    multi_exclusions = sum(n_exclusions >= 2),
    max_overlap = max(n_exclusions),
    total_bases = sum(bases_bp),
    total_variants = sum(variant_count),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    bench_type = bench_type_labels[bench_type]
  ) %>%
  flextable::flextable() %>%
  fmt_integer_flextable(
    columns = c(
      "total_combinations", "single_exclusions", "multi_exclusions",
      "total_bases", "total_variants"
    )
  ) %>%
  flextable::align(
    j = c(
      "total_combinations", "single_exclusions", "multi_exclusions",
      "max_overlap", "total_bases", "total_variants"
    ),
    align = "right", part = "all"
  ) %>%
  cols_label_flextable(
    ref = "Reference",
    bench_type = "Type",
    total_combinations = "Total Combos",
    single_exclusions = "Single Only",
    multi_exclusions = "Multi-Overlap",
    max_overlap = "Max Overlap",
    total_bases = "Total Bases",
    total_variants = "Total Variants"
  ) %>%
  theme_flextable_manuscript()
```
