---
title: "Variant Counts in Difficult Genomic Contexts"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Overview

This notebook focuses on variant counts within difficult genomic contexts
(HP, MAP, SD, TR) across benchmark versions. It produces Figure 3
(variant counts by genomic context) and fold change analysis.

For coverage-level analysis of genomic contexts (% context covered, % bench
in context), see `genomic_context_analysis.qmd`.

## Data Loading

```{r}
analysis_bench_order <- rev(BENCH_VERSION_LEVELS)
variants_df <- load_all_variant_tables() %>%
  filter(region_ids == "BMKREGIONS")
```

## Variants in Difficult Genomic Contexts

This section visualizes variant counts by genomic context region using loaded variant tables.

```{r variant_counts_by_context}
# Create variant count breakdown by genomic context from aggregated metrics
genomic_context_var_df <- variants_df %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  count(bench_version, ref, bench_type, var_type, context_ids, name = "count")
```

```{r}
genomic_context_var_df
```

```{r variant_counts_by_context_plot}
(genomic_context_smvar_plt <- genomic_context_var_df %>%
  filter(bench_type == "smvar", ref == "GRCh38") %>%
  mutate(
    bench_version = factor(
      bench_version,
      levels = analysis_bench_order
    )
  ) %>%
  ggplot() +
  geom_bar(
    aes(
      x = var_type,
      y = count,
      fill = bench_version
    ),
    color = "grey20",
    position = "dodge",
    stat = "identity",
    show.legend = TRUE
  ) +
  facet_wrap(. ~ context_ids, nrow = 1) +
  scale_y_continuous(label = scales::label_comma()) +
  scale_benchmark_version(aesthetic = "fill") +
  labs(x = "Variant Type", y = "Variant Count") +
  theme_manuscript() +
  theme(
    legend.position = "bottom"
  ))
```

```{r structural_variant_counts_by_context_plot}
(genomic_context_stvar_plt <- genomic_context_var_df %>%
  filter(bench_type == "stvar", ref == "GRCh37", .preserve = TRUE) %>%
  mutate(
    bench_version = factor(
      bench_version,
      levels = analysis_bench_order
    )
  ) %>%
  ggplot() +
  geom_bar(
    aes(x = var_type, y = count, fill = bench_version),
    color = "grey20",
    position = "dodge",
    stat = "identity",
    show.legend = TRUE
  ) +
  facet_wrap(. ~ context_ids, nrow = 1) +
  scale_y_continuous(label = scales::label_comma()) +
  scale_benchmark_version(aesthetic = "fill") +
  labs(x = "Variant Type", y = "Variant Count") +
  theme_manuscript())
```

```{r  combined_variant_counts_by_context_plot}
(combined_context_plt <- genomic_context_smvar_plt +
  genomic_context_stvar_plt +
  plot_layout(guides = "collect", nrow = 2, ncol = 1) +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "bottom"
  ))

ggsave(
  filename = here("manuscript", "figs", "genomic_context_variant_counts.png"),
  plot = combined_context_plt,
  width = 8,
  height = 6,
  dpi = 300
)
```


```{r variant_counts_by_context_fold_change}
genomic_context_var_fld_chg <- variants_df %>%
  mutate(
    var_size_bin = case_when(
      abs(var_size) < 15 ~ "<15bp",
      abs(var_size) < 50 ~ "15-49bp",
      abs(var_size) < 1001 ~ "50 - 1,000bp",
      abs(var_size) < 10001 ~ "1,001 - 10,000bp",
      .default = ">10kb"
    )
  ) %>%
  count(bench_version, ref, bench_type, var_type, var_size_bin, context_ids) %>%
  pivot_wider(
    names_from = bench_version,
    values_from = n,
    id_cols = c(ref, bench_type, var_type, var_size_bin, context_ids)
  ) %>%
  filter(
    !is.na(v0.6) | !is.na(v4.2.1),
    !(ref == "GRCh37" & bench_type == "smvar")
  ) %>%
  mutate(
    fold_chg = case_when(
      !is.na(v0.6) ~ v5.0q / v0.6,
      !is.na(v4.2.1) ~ v5.0q / v4.2.1,
      .default = NA
    )
  )
```


```{r variant_counts_by_context_fold_change_plot_combined}
genomic_context_var_fld_chg %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  mutate(
    bench_type = factor(
      bench_type,
      levels = BENCH_TYPE_LEVELS,
      labels = c("Small Variants", "Structural Variants")
    )
  ) %>%
  ggplot() +
  geom_point(
    aes(
      x = context_ids,
      y = fold_chg,
      fill = var_size_bin,
      shape = var_type
    ),
    size = 4,
    color = "black",
    position = "jitterdodge"
  ) +
  facet_wrap(~bench_type, scales = "free", ncol = 1) +
  scale_shape_manual(
    values = c(21, 22, 23, 24),
    guide = guide_legend(override.aes = list(fill = "grey80"))
  ) +
  scale_fill_viridis_d(
    labels = c("<15bp", "15-49bp", "50bp-1kb", "1kb-10kb", ">10kb"),
    guide = guide_legend(override.aes = list(shape = 21))
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    x = "Difficult Genomic Context",
    y = "Fold Change Variant Count",
    fill = "Variant Size Bin",
    shape = "Variant Class"
  ) +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical"
  )
```




