---
title: "v5q Benchmark Set Variant and Region Characterization"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(
  load_plot_themes = TRUE,
  load_sessioninfo = TRUE
)
```

## Data Loading

This analysis uses **primary aggregated metrics** rather than loading full variant tables directly. This approach is faster, more memory-efficient, and sufficient for most analyses. See the [Pipeline Outputs Reference](../docs/pipeline-outputs.md) for details on when to load detailed variant tables.

### Primary Metrics

```{r load_primary}
analysis_data <- load_primary_analysis_data(
  include_variants = TRUE,
  include_reference_sizes = TRUE,
  include_hg002q100_size = TRUE,
  include_benchmark_regions = TRUE
)
list2env(analysis_data, envir = environment())

## Filtering variants_df to benchmark regions for the characterization
variants_df <- variants_df %>%
  filter(region_ids == "BMKREGIONS", is_pass)
```

## Benchmark Set Summary Table

### Benchmark Regions

```{r}
glimpse(bench_regions_df)
```

```{r}
grch37_dip_bp <- 2831481872
grch38_dip_bp <- 2841316100
chm13_dip_bp <- 2938461012

dip_df <- tibble(
  ref = c("GRCh37", "GRCh38", "CHM13v2.0"),
  dip_bp = c(grch37_dip_bp, grch38_dip_bp, chm13_dip_bp)
)

bench_region_stats_df <- bench_regions_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1", "v0.6"),
    chrom %in% CHROM_LEVELS
  ) %>%
  group_by(bench_type, ref, bench_version) %>%
  summarise(total_bp = sum(interval_size), .groups = "drop") %>%
  left_join(total_ref_asm_bp) %>%
  left_join(dip_df) %>% 
  mutate(
    pct_ref_cvg = round(100 * (total_bp / total_asm_bp), 0),
    pct_dip_cvg = if_else(bench_version == "v5.0q", round(100 * (total_bp / dip_bp), 0), NA_real_),
    pct_q100_cvg = round(100 * (total_bp / hg002q100_size), 0),
    pct_q100_dip_cvg = if_else(bench_version == "v5.0q", round(100 * (dip_bp / hg002q100_size), 0), NA_real_),
    pct_ref_dip_cvg = if_else(bench_version == "v5.0q", round(100 * (dip_bp/ total_asm_bp), 0), NA_real_)
  ) %>%
  select(-total_asm_bp, -dip_bp)
```
```{r}
bench_region_stats_df %>%
  filter(bench_version == "v5.0q") %>%
  select(ref, bench_type, pct_dip_cvg, pct_ref_dip_cvg, pct_ref_cvg, pct_q100_dip_cvg, pct_q100_cvg, total_bp) %>% 
  mutate(ref = std_references(ref)) %>% arrange(ref) 
```
```{r}
bench_region_stats_df %>%
  select(ref, bench_type, bench_version, pct_ref_cvg, total_bp) %>% 
  mutate(ref = std_references(ref)) %>% arrange(ref) 
```


### Bechmark Variants

```{r}
bench_var_cnt_df <- variants_df %>%
  # filter(bench_version == "v5.0q") %>%
  group_by(bench_type, ref) %>%
  count(bench_type, ref, bench_version, var_type) %>%
  ungroup() #%>%
  # pivot_wider(
  #   names_from = var_type,
  #   values_from = n
  # )
```

```{r}
bench_var_cnt_df %>% 
  group_by(bench_type, ref, bench_version) %>% 
  summarise(n = sum(n), .groups = "drop") %>% 
  add_column(var_type = "Total") %>% 
  bind_rows(bench_var_cnt_df)
```


```{r}
## Join variant counts with region stats
bench_summary_df <- bench_var_cnt_df %>%
  pivot_wider(names_from = var_type, values_from = n, values_fill = 0L) %>%
  mutate(Total = rowSums(across(any_of(c("SNV", "INDEL", "INS", "DEL"))))) %>%
  full_join(bench_region_stats_df, by = c("bench_type", "ref", "bench_version")) %>%
  mutate(
    ref = std_references(ref)
  ) %>%
  arrange(bench_type, ref)

## Small variants table
smvar_tbl <- bench_summary_df %>%
  filter(bench_type == "smvar") %>%
  select(bench_type, ref, bench_version, SNV, INDEL, Total, total_bp, pct_dip_cvg, pct_ref_dip_cvg, pct_ref_cvg, pct_q100_dip_cvg, pct_q100_cvg)

## Structural variants table
stvar_tbl <- bench_summary_df %>%
  filter(bench_type == "stvar") %>%
  select(bench_type, ref, bench_version, DEL, INS, Total, total_bp, pct_dip_cvg, pct_ref_dip_cvg, pct_ref_cvg, pct_q100_dip_cvg, pct_q100_cvg)

## Shared formatting function
format_bench_table <- function(df) {
  df %>% select(-bench_type) %>% as_flextable(show_coltype = FALSE) %>%
    # as_grouped_flextable(groupname_col = "bench_type") %>%
    fmt_integer_flextable(
      columns = intersect(c("SNV", "INDEL", "DEL", "INS", "Total", "total_bp"), names(df))
    ) %>%
    fmt_percent_flextable(
      columns = starts_with("pct_"),
      decimals = 0
    ) %>%
    cols_label_flextable(
      ref = "Reference",
      bench_version = "Version",
      Total = "Total",
      total_bp = "Benchmark Bases",
      pct_dip_cvg = "% Dip. Cov.",
      pct_ref_cvg = "% Ref Cov.",
      pct_ref_dip_cvg = "% Ref. Dip. Cov.",
      pct_q100_cvg = "% Asm Cov.",
      pct_q100_dip_cvg = "% Asm. Dip. Cov."
    ) %>%
    flextable::align(j = "ref", align = "right", part = "body") %>%
    theme_flextable_manuscript()
}
```
```{r}
format_bench_table(smvar_tbl)
```


## Variant Size Distributions


```{r sv_length_distribution}
#| message: true
# Load structural variant data for selected benchmarks
# This loads full variant tables for size distribution analysis

stvar_len_df <- variants_df %>%
  filter(
    bench_version %in% c("v5.0q", "v0.6"),
    bench_type == "stvar"
  ) %>%
  mutate(ref = std_references(ref)) %>%
  select(bench_version, bench_type, ref, chrom, pos, var_size) %>%
  filter(abs(var_size) < 20000)

(stvar_len_dist_plt <- ggplot(
  stvar_len_df,
  aes(x = var_size)
) +
  geom_histogram(
    data = filter(stvar_len_df, bench_version != "v0.6", ref == "GRCh38"),
    fill = "grey80",
    color = "grey20",
    bins = 50,
    drop = "all"
  ) +
  # geom_histogram(
  #   data = filter(stvar_len_df, bench_version == "v0.6"),
  #   fill = "grey30",
  #   color = "grey20",
  #   bins = 50,
  #   drop = "all"
  # ) +
  scale_x_continuous(
    breaks = c(
      -10000,
      -5000,
      -1000,
      -500,
      -100,
      -50,
      0,
      50,
      100,
      500,
      1000,
      5000,
      10000
    ),
    trans = scales::pseudo_log_trans(sigma = 100, base = 10),
    labels = scales::comma
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "SV Length (bp)",
    y = "Count"
  ) +
  # facet_wrap(~ref, ncol = 1) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0)
  ))
```


```{r indel_length_distribution}
#| message: true
#| warning: false
# Load small variant data for selected benchmarks

smvar_len_df <- variants_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1"),
    bench_type == "smvar",
    var_type == "INDEL"
  ) %>%
  mutate(ref = std_references(ref)) %>%
  select(bench_version, bench_type, ref, chrom, pos, var_size) %>% 
  mutate(var_size = if_else(var_size > 0, var_size + 0.5, var_size - 0.5))

(smvar_len_dist_plt <- ggplot(
  smvar_len_df,
  aes(x = var_size)
) +
  geom_histogram(
    data = filter(smvar_len_df, bench_version != "v4.2.1", ref == "GRCh38"),
    fill = "grey80",
    color = "grey20",
    binwidth = 1,
    # closed = "right",
    drop = "all"
  ) +
  # geom_histogram(
  #   data = filter(smvar_len_df, bench_version == "v4.2.1"),
  #   fill = "grey30",
  #   color = "grey20",
  #   binwidth = 2,
  #   drop = "all"
  # ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "INDEL Length (bp)",
    y = "Count"
  ) +
  # facet_wrap(~ref, ncol = 1) +
  theme_manuscript())
```

```{r}
(combined_var_len_dist_plt <- (smvar_len_dist_plt) +
  (stvar_len_dist_plt) +
  plot_layout(
    ncol = 2,
    guides = "collect",
    widths = c(1, 2)
  ) +
  plot_annotation(tag_levels = "A") &
  theme_manuscript() +
    theme(
      axis.text.x = element_text(angle = -45, hjust = 0),
      plot.margin = unit(c(5, 5, 5, 5), "pt")
    ))
```

```{r var_length_dist_save}
ggsave(
  filename = here("manuscript", "figs", "var_len_dist.png"),
  plot = combined_var_len_dist_plt,
  width = 8,
  height = 4,
  dpi = 300
)
```



## Benchmark region characterization

- Number of bases in benchmark regions in v5 compared to v4 and v0.6

### Relative changes between v5q and v4.2.1/v0.6 by chromosome
```{r fig.width = 8, fig.height = 4}
(coverage_change_plt <- bench_regions_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1", "v0.6"),
    (bench_type == "smvar" & ref == "GRCh38") |
      (bench_type == "stvar" & ref == "GRCh37"),
    chrom %in% AUTOSOME_CHROM_LEVELS
  ) %>%
  select(-starts_with("X")) %>%
  group_by(bench_type, chrom, bench_version) %>%
  summarise(total_size = sum(interval_size), .groups = "drop") %>%
  mutate(
    vers = if_else(bench_version == "v5.0q", "new", "old"),
    chrom = factor(chrom, levels = AUTOSOME_CHROM_LEVELS)
  ) %>%
  select(-bench_version) %>%
  pivot_wider(names_from = vers, values_from = total_size) %>%
  mutate(
    pct_change = 100 * (new - old) / old,
    total_change = new - old
  ) %>%
  ggplot() +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(
    aes(
      x = chrom,
      y = pct_change,
      fill = total_change / 1000000,
      shape = bench_type
    ),
    size = 4,
    color = "black",
    position = "jitterdodge"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    n.breaks = 10
  ) +
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0,
    labels = scales::comma
  ) +
  scale_shape_manual(values = c(21, 24)) +
  labs(
    x = "Chromosome",
    y = "Coverage Change",
    fill = "Coverage change (Mb)",
    shape = "Variant Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0),
    legend.position = "bottom"
  ))
```

```{r}
(ggsave(
  filename = here("manuscript", "figs", "benchmark_region_size_change.png"),
  plot = coverage_change_plt,
  width = 8,
  height = 4,
  dpi = 300
))
```
