---
title: "Variant Counts in Difficult Genomic Contexts"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Overview

This notebook focuses on variant counts within difficult genomic contexts
(HP, MAP, SD, TR) across benchmark versions. It produces Figure 3
(variant counts by genomic context) and fold change analysis.

For coverage-level analysis of genomic contexts (% context covered, % bench
in context), see `genomic_context_analysis.qmd`.

## Data Loading

```{r}
analysis_bench_order <- rev(BENCH_VERSION_LEVELS)
variants_df <- load_all_variant_tables() %>%
  filter(region_ids == "BMKREGIONS", is_pass)
```

## Variant Fold Change by Genomic Context

Compute fold change in total variant counts (v5.0q vs legacy) per context.
Small variants: GRCh38 v5.0q / v4.2.1. Structural variants: GRCh37 v5.0q / v0.6.

Complex variants are defined following the GIAB genome stratifications
(v3.6 GenomeSpecific): regions with two or more variants within 10 bp of
each other on the same chromosome.

```{r complex_variants}
## Identify complex variants: >=2 variants within 10bp on the same chromosome.
## A variant is "complex" if it is within 10bp of at least one other variant
## in the same benchmark set.
complex_counts <- variants_df %>%
  arrange(bench_version, ref, bench_type, chrom, pos) %>%
  group_by(bench_version, ref, bench_type, chrom) %>%
  mutate(
    dist_to_next = lead(pos) - pos,
    dist_to_prev = pos - lag(pos)
  ) %>%
  ungroup() %>%
  filter(
    (!is.na(dist_to_next) & dist_to_next <= 10) |
      (!is.na(dist_to_prev) & dist_to_prev <= 10)
  ) %>%
  count(bench_version, ref, bench_type, name = "count") %>%
  mutate(context_ids = "COMPLEX")
```

```{r fold_change_data}
## Total variant counts per context (no size bin breakdown)
## Explode comma-separated context_ids so variants in multiple contexts
## (e.g., "SD,TR") are counted once per context
genomic_context_counts <- variants_df %>%
  filter(!is.na(context_ids), context_ids != "") %>%
  separate_rows(context_ids, sep = ",") %>%
  mutate(context_ids = str_trim(context_ids)) %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  count(bench_version, ref, bench_type, context_ids, name = "count")

## Combine genomic context counts with complex variant counts
## Keep long format for grouped bar plot (one row per version x context)
## Drop contexts with only one version (no paired comparison possible)
context_counts_long <- bind_rows(genomic_context_counts, complex_counts) %>%
  filter(
    (bench_type == "smvar" & ref == "GRCh38" &
      bench_version %in% c("v5.0q", "v4.2.1")) |
      (bench_type == "stvar" & ref == "GRCh37" &
        bench_version %in% c("v5.0q", "v0.6"))
  ) %>%
  group_by(bench_type, context_ids) %>%
  filter(n_distinct(bench_version) >= 2) %>%
  ungroup() %>%
  mutate(
    context_label = get_context_labels()[as.character(context_ids)],
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  )

## Compute fold change for text labels (one row per context x bench_type)
context_fold_chg <- context_counts_long %>%
  pivot_wider(
    names_from = bench_version,
    values_from = count,
    id_cols = c(ref, bench_type, context_ids, context_label, bench_type_label)
  ) %>%
  mutate(
    legacy_count = coalesce(v4.2.1, v0.6),
    fold_chg = v5.0q / legacy_count
  ) %>%
  filter(!is.na(fold_chg)) %>%
  ## Position fold change label above the taller bar (v5.0q)
  mutate(label_y = v5.0q)
```

### Figure 3: Variant Counts by Genomic Context

```{r variant_count_main_plot, fig.width = 8, fig.height = 4.5}
(variant_count_plt <- context_counts_long %>%
   mutate(context_id = if_else(context_ids == "COMPLEX", "CMPLX", context_ids)) %>% 
  ggplot(aes(
    x = context_ids,
    y = count,
    fill = bench_version
  )) +
  geom_col(
    position = position_dodge(width = 0.8),
    color = "grey30",
    linewidth = 0.3,
    width = 0.7
  ) +
  geom_text(
    data = context_fold_chg,
    aes(
      x = context_ids,
      y = label_y,
      label = sprintf("%.1fx", fold_chg)
    ),
    inherit.aes = FALSE,
    vjust = -1,
    size = 3,
    fontface = "bold"
  ) +
  facet_wrap(~bench_type_label, scales = "free") +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.2)),
    labels = scales::comma,
    breaks = scales::pretty_breaks(n = 5)
  ) +
  scale_benchmark_version(aesthetic = "fill") +
  labs(
    x = "Genomic Context",
    y = "Variant Count",
    fill = "Benchmark Version"
  ) +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    plot.margin = margin(1, 5, 1, 1)
  ))

ggsave(
  filename = here("figures", "genomic_context_variant_counts.png"),
  plot = variant_count_plt,
  width = 6,
  height = 3,
  dpi = 300
)
ggsave(
  filename = here("figures", "genomic_context_variant_counts.pdf"),
  plot = variant_count_plt,
  width = 6,
  height = 3
)
```

### Supplemental: Fold Change by Variant Size Bin

```{r fold_change_by_size_data}
size_bin_levels <- c("<15bp", "15-49bp", "50-1,000bp", "1,001-10,000bp", ">10kb")

add_size_bin <- function(df) {
  df %>%
    mutate(
      var_size_bin = factor(case_when(
        abs(var_size) < 15 ~ "<15bp",
        abs(var_size) < 50 ~ "15-49bp",
        abs(var_size) < 1001 ~ "50-1,000bp",
        abs(var_size) < 10001 ~ "1,001-10,000bp",
        .default = ">10kb"
      ), levels = size_bin_levels)
    )
}

## Genomic context counts by size bin
gc_size_counts <- variants_df %>%
  filter(!is.na(context_ids), context_ids != "") %>%
  separate_rows(context_ids, sep = ",") %>%
  mutate(context_ids = str_trim(context_ids)) %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  add_size_bin() %>%
  count(bench_version, ref, bench_type, var_type, var_size_bin, context_ids)

## Complex variant counts by size bin
complex_size_counts <- variants_df %>%
  arrange(bench_version, ref, bench_type, chrom, pos) %>%
  group_by(bench_version, ref, bench_type, chrom) %>%
  mutate(
    dist_to_next = lead(pos) - pos,
    dist_to_prev = pos - lag(pos)
  ) %>%
  ungroup() %>%
  filter(
    (!is.na(dist_to_next) & dist_to_next <= 10) |
      (!is.na(dist_to_prev) & dist_to_prev <= 10)
  ) %>%
  add_size_bin() %>%
  count(bench_version, ref, bench_type, var_type, var_size_bin) %>%
  mutate(context_ids = "COMPLEX")

genomic_context_var_fld_chg <- bind_rows(gc_size_counts, complex_size_counts) %>%
  pivot_wider(
    names_from = bench_version,
    values_from = n,
    id_cols = c(ref, bench_type, var_type, var_size_bin, context_ids)
  ) %>%
  filter(
    (bench_type == "smvar" & ref == "GRCh38" & !is.na(v4.2.1)) |
      (bench_type == "stvar" & ref == "GRCh37" & !is.na(v0.6))
  ) %>%
  mutate(
    fold_chg = coalesce(v5.0q / v4.2.1, v5.0q / v0.6),
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  )
```

```{r fold_change_by_size_plot, fig.width = 8, fig.height = 6}
(fold_change_size_plt <- genomic_context_var_fld_chg %>%
  ggplot() +
  geom_point(
    aes(
      x = var_size_bin,
      y = fold_chg,
      shape = var_type,
      fill = var_size_bin
    ),
    size = 3,
    color = "black",
    position = position_jitterdodge(jitter.width = 0.15)
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  facet_grid(bench_type_label~context_ids, scales = "free") +
  scale_shape_manual(
    values = c(21, 22, 23, 24),
    guide = guide_legend(override.aes = list(fill = "grey80"))
  ) +
  scale_fill_viridis_d(
    guide = guide_legend(override.aes = list(shape = 21))
  ) +
  scale_y_continuous(
    trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4, 8, 16),
    labels = scales::label_number(accuracy = 0.01, drop0trailing = TRUE)
  ) +
  labs(
    x = "Variant Size",
    y = "log₂ Fold Change",
    # fill = "Variant Size",
    shape = "Variant Type"
  ) +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical"
  ))

ggsave(
  filename = here("figures", "genomic_context_fold_change_by_size.png"),
  plot = fold_change_size_plt,
  width = 8,
  height = 6,
  dpi = 300
)
ggsave(
  filename = here("figures", "genomic_context_fold_change_by_size.pdf"),
  plot = fold_change_size_plt,
  width = 8,
  height = 6
)
```

### Supplemental: Small Variant Count by Size Bin

Raw variant counts for GRCh38 small variants (v4.2.1 and v5.0q), broken
down by variant type (SNP, INS, DEL) and size bin. Only the two small-variant
size bins (<15bp and 15-49bp) are shown; v0.6 is excluded because it uses
GRCh37 for smvar.

```{r smvar_size_bin_table}
smvar_size_bin_counts <- variants_df %>%
  filter(
    bench_type == "smvar",
    ref == "GRCh38",
    bench_version %in% c("v4.2.1", "v5.0q"),
    var_type %in% c("SNP", "INS", "DEL")
  ) %>%
  add_size_bin() %>%
  filter(var_size_bin %in% c("<15bp", "15-49bp")) %>%
  count(var_type, var_size_bin, bench_version, name = "count") %>%
  pivot_wider(
    names_from = bench_version,
    values_from = count,
    values_fill = 0L
  ) %>%
  arrange(var_type, var_size_bin) %>%
  select(var_type, var_size_bin, `v4.2.1`, `v5.0q`)

smvar_size_bin_counts %>%
  flextable() %>%
  fmt_integer_flextable(columns = c("v4.2.1", "v5.0q")) %>%
  cols_label_flextable(
    var_type = "Variant Type",
    var_size_bin = "Size Bin"
  ) %>%
  theme_flextable_manuscript()
```
