---
title: "Variant Counts in Difficult Genomic Contexts"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Overview

This notebook focuses on variant counts within difficult genomic contexts
(HP, MAP, SD, TR) across benchmark versions. It produces Figure 3
(variant counts by genomic context) and fold change analysis.

For coverage-level analysis of genomic contexts (% context covered, % bench
in context), see `genomic_context_analysis.qmd`.

## Data Loading

```{r}
analysis_bench_order <- rev(BENCH_VERSION_LEVELS)
variants_df <- load_all_variant_tables() %>%
  filter(region_ids == "BMKREGIONS")
```

## Variant Fold Change by Genomic Context

Compute fold change in total variant counts (v5.0q vs legacy) per context.
Small variants: GRCh38 v5.0q / v4.2.1. Structural variants: GRCh37 v5.0q / v0.6.

```{r fold_change_data}
## Total variant counts per context (no size bin breakdown)
context_totals <- variants_df %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  count(bench_version, ref, bench_type, context_ids, name = "count") %>%
  pivot_wider(
    names_from = bench_version,
    values_from = count,
    id_cols = c(ref, bench_type, context_ids)
  ) %>%
  filter(
    (bench_type == "smvar" & ref == "GRCh38" & !is.na(v4.2.1)) |
      (bench_type == "stvar" & ref == "GRCh37" & !is.na(v0.6))
  ) %>%
  mutate(
    legacy_count = coalesce(v4.2.1, v0.6),
    fold_chg = v5.0q / legacy_count,
    context_label = get_context_labels()[as.character(context_ids)],
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  )
```

### Figure 3: Fold Change in Variant Counts by Genomic Context

```{r fold_change_main_plot, fig.width = 7, fig.height = 4}
(fold_change_plt <- context_totals %>%
  ggplot(aes(
    x = context_label,
    y = fold_chg,
    fill = context_ids
  )) +
  geom_col(color = "grey30", linewidth = 0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  geom_text(
    aes(label = sprintf("%.1fx", fold_chg)),
    vjust = -0.5,
    size = 3
  ) +
  facet_wrap(~bench_type_label, scales = "free_x") +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    breaks = scales::pretty_breaks(n = 5)
  ) +
  scale_genomic_context(aesthetic = "fill", guide = "none") +
  labs(
    x = "Genomic Context",
    y = "Fold Change (v5.0q / legacy)"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -30, hjust = 0)
  ))

ggsave(
  filename = here("manuscript", "figs", "genomic_context_variant_counts.png"),
  plot = fold_change_plt,
  width = 7,
  height = 4,
  dpi = 300
)
```

### Supplemental: Fold Change by Variant Size Bin

```{r fold_change_by_size_data}
genomic_context_var_fld_chg <- variants_df %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  mutate(
    var_size_bin = factor(case_when(
      abs(var_size) < 15 ~ "<15bp",
      abs(var_size) < 50 ~ "15-49bp",
      abs(var_size) < 1001 ~ "50-1,000bp",
      abs(var_size) < 10001 ~ "1,001-10,000bp",
      .default = ">10kb"
    ), levels = c("<15bp", "15-49bp", "50-1,000bp", "1,001-10,000bp", ">10kb"))
  ) %>%
  count(bench_version, ref, bench_type, var_type, var_size_bin, context_ids) %>%
  pivot_wider(
    names_from = bench_version,
    values_from = n,
    id_cols = c(ref, bench_type, var_type, var_size_bin, context_ids)
  ) %>%
  filter(
    (bench_type == "smvar" & ref == "GRCh38" & !is.na(v4.2.1)) |
      (bench_type == "stvar" & ref == "GRCh37" & !is.na(v0.6))
  ) %>%
  mutate(
    fold_chg = coalesce(v5.0q / v4.2.1, v5.0q / v0.6),
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  )
```

```{r fold_change_by_size_plot, fig.width = 8, fig.height = 6}
(fold_change_size_plt <- genomic_context_var_fld_chg %>%
  ggplot() +
  geom_point(
    aes(
      x = context_ids,
      y = fold_chg,
      fill = var_size_bin,
      shape = var_type
    ),
    size = 3,
    color = "black",
    position = position_jitterdodge(jitter.width = 0.15)
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  facet_wrap(~bench_type_label, scales = "free", ncol = 1) +
  scale_shape_manual(
    values = c(21, 22, 23, 24),
    guide = guide_legend(override.aes = list(fill = "grey80"))
  ) +
  scale_fill_viridis_d(
    guide = guide_legend(override.aes = list(shape = 21))
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Genomic Context",
    y = "Fold Change (v5.0q / legacy)",
    fill = "Variant Size",
    shape = "Variant Type"
  ) +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical"
  ))

ggsave(
  filename = here("manuscript", "figs", "genomic_context_fold_change_by_size.png"),
  plot = fold_change_size_plt,
  width = 8,
  height = 6,
  dpi = 300
)
```

