---
title: "v5q Benchmark Set Variant and Region Characterization"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(
  load_plot_themes = TRUE,
  load_sessioninfo = TRUE
)
```

## Data Loading

This analysis uses **primary aggregated metrics** rather than loading full variant tables directly. This approach is faster, more memory-efficient, and sufficient for most analyses. See the [Pipeline Outputs Reference](../docs/pipeline-outputs.md) for details on when to load detailed variant tables.

### Primary Metrics

```{r load_primary}
analysis_data <- load_primary_analysis_data(
  include_variants = TRUE,
  include_reference_sizes = TRUE,
  include_hg002q100_size = TRUE,
  include_benchmark_regions = TRUE
)
list2env(analysis_data, envir = environment())

## Filtering variants_df to benchmark regions for the characterization
variants_df <- variants_df %>%
  filter(region_ids == "BMKREGIONS", is_pass)
```

## Benchmark Set Summary Table

Consolidated summary of variant counts and benchmark region sizes across
benchmark versions. GRCh38 is the primary reference for the manuscript;
GRCh37 and CHM13v2.0 are included in supplemental tables.


```{r benchmark_summary_data}

## Compute benchmark region sizes
bench_region_stats_df <- bench_regions_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1", "v0.6"),
    chrom %in% CHROM_LEVELS
  ) %>%
  group_by(bench_type, ref, bench_version) %>%
  summarise(total_bp = sum(interval_size), .groups = "drop") %>%
  left_join(total_ref_asm_bp, by = "ref") %>%
  mutate(pct_ref_cvg = 100 * (total_bp / total_asm_bp),
         pct_asm_cvg = 100* (total_bp / hg002q100_size)) %>%
  select(-total_asm_bp)

## Compute variant counts by type
bench_var_cnt_df <- variants_df %>%
  count(bench_type, ref, bench_version, var_type)

## Build unified summary: variant counts + region stats
bench_summary_df <- bench_var_cnt_df %>%
  pivot_wider(names_from = var_type, values_from = n, values_fill = 0L) %>%
  mutate(Total = rowSums(across(any_of(c("SNV", "INDEL", "INS", "DEL"))))) %>%
  full_join(bench_region_stats_df, by = c("bench_type", "ref", "bench_version")) %>%
  mutate(
    ref = std_references(ref),
    bench_version = std_bench_versions(bench_version),
    bench_mb = round(total_bp / 1e6, 0)
  ) %>%
  arrange(bench_type, ref, bench_version)
```

```{r format_bench_table_fn}
## Shared formatting function for benchmark summary tables
format_bench_table <- function(df, var_cols) {
  df %>%
    select(ref, bench_version, all_of(var_cols), Total, bench_mb, pct_ref_cvg, pct_asm_cvg) %>%
    as_flextable(show_coltype = FALSE) %>%
    fmt_integer_flextable(columns = c(var_cols, "Total", "bench_mb")) %>%
    fmt_number_flextable(columns = c("pct_ref_cvg", "pct_asm_cvg"), decimals = 1) %>%
    cols_label_flextable(
      ref = "Reference",
      bench_version = "Version",
      Total = "Total Variants",
      bench_mb = "Bench (Mb)",
      pct_ref_cvg = "% Ref",
      pct_asm_cvg = "% Asm"
    ) %>%
    flextable::align(j = "ref", align = "right", part = "body") %>%
    theme_flextable_manuscript() %>%
    knitr::knit_print()
}
```

### Main Text: Summary (Table 1)

Small variants use GRCh38 (v5.0q vs v4.2.1); structural variants use GRCh37
(v5.0q vs v0.6) since v0.6 was not generated for GRCh38.

```{r table1_combined}
## Compute Total Variation (sum of absolute variant lengths, in Mb)
total_var_df <- variants_df %>%
  group_by(bench_type, ref, bench_version) %>%
  mutate(var_len = if_else(var_size == 0, 1, abs(var_size))) %>% 
  summarise(total_var_bp = sum(var_len), .groups = "drop") %>%
  mutate(
    ref = std_references(ref),
    bench_version = std_bench_versions(bench_version)
  )

bench_full_df <- bench_summary_df %>%
  left_join(total_var_df, by = c("bench_type", "ref", "bench_version"))

## Build 6 metric rows for one bench type
build_metric_rows <- function(bt, type_label, var1_col, var1_label,
                               var2_col, var2_label, delta_ref, legacy_ver) {
  gv5 <- function(r) bench_full_df %>%
    filter(bench_type == bt, bench_version == "v5.0q", ref == r)

  v5_37 <- gv5("GRCh37")
  v5_38 <- gv5("GRCh38")
  v5_c  <- gv5("CHM13v2.0")
  v5_d  <- gv5(delta_ref)
  leg   <- bench_full_df %>%
    filter(bench_type == bt, bench_version == legacy_ver, ref == delta_ref)

  p1 <- function(df, col) {
    v <- df[[col]]
    if (length(v) == 0 || is.na(v[1])) NA_real_ else as.numeric(v[1])
  }
  fk   <- function(x) if (is.na(x)) "—" else prettyNum(as.integer(round(x)), big.mark = ",")
  fmb  <- function(x) if (is.na(x)) "—" else sprintf("%.1f", x)
  fr   <- function(x) if (is.na(x)) "—" else prettyNum(round(x), big.mark = ",")
  fp   <- function(x) if (is.na(x)) "—" else sprintf("%.1f%%", x)
  dn_k  <- function(v, l) {
    if (anyNA(c(v, l)) || l == 0) return("—")
    diff <- as.integer(round(v - l))
    paste0(if (diff >= 0) "+" else "", prettyNum(diff, big.mark = ","))
  }
  dn_mb <- function(v, l) if (anyNA(c(v, l)) || l == 0) "—" else sprintf("%+.1f Mb", v - l)
  dn_r  <- function(v, l) if (anyNA(c(v, l)) || l == 0) "—" else sprintf("%+.0f Mb", v - l)
  dp    <- function(v, l) if (anyNA(c(v, l)) || l == 0) "—" else sprintf("%+.1f%%", 100 * (v - l) / l)
  dpp   <- function(v, l) if (anyNA(c(v, l))) "—" else sprintf("%+.1f pp", v - l)

  v1 <- list(r37 = p1(v5_37, var1_col), r38 = p1(v5_38, var1_col), rc = p1(v5_c, var1_col),
             d = p1(v5_d, var1_col), l = p1(leg, var1_col))
  v2 <- list(r37 = p1(v5_37, var2_col), r38 = p1(v5_38, var2_col), rc = p1(v5_c, var2_col),
             d = p1(v5_d, var2_col), l = p1(leg, var2_col))
  vt <- list(r37 = p1(v5_37, "Total"), r38 = p1(v5_38, "Total"), rc = p1(v5_c, "Total"),
             d = p1(v5_d, "Total"), l = p1(leg, "Total"))
  vbp <- list(r37 = p1(v5_37, "total_var_bp"), r38 = p1(v5_38, "total_var_bp"),
             rc = p1(v5_c, "total_var_bp"), d = p1(v5_d, "total_var_bp"), l = p1(leg, "total_var_bp"))
  rg <- list(r37 = p1(v5_37, "bench_mb"), r38 = p1(v5_38, "bench_mb"), rc = p1(v5_c, "bench_mb"),
             d = p1(v5_d, "bench_mb"), l = p1(leg, "bench_mb"))
  pr <- list(r37 = p1(v5_37, "pct_ref_cvg"), r38 = p1(v5_38, "pct_ref_cvg"),
             rc = p1(v5_c, "pct_ref_cvg"), d = p1(v5_d, "pct_ref_cvg"), l = p1(leg, "pct_ref_cvg"))

  tibble(
    bench_type_label = type_label,
    col_a = c(var1_label, var2_label, "Total", "Total", "Region (Mb)", "% Reference"),
    col_b = c("", "", "Variants", "Variation (bp)", "", ""),
    grch37 = c(fk(v1$r37), fk(v2$r37), fk(vt$r37), fmb(vbp$r37), fr(rg$r37), fp(pr$r37)),
    grch38 = c(fk(v1$r38), fk(v2$r38), fk(vt$r38), fmb(vbp$r38), fr(rg$r38), fp(pr$r38)),
    chm13  = c(fk(v1$rc),  fk(v2$rc),  fk(vt$rc),  fmb(vbp$rc),  fr(rg$rc),  fp(pr$rc)),
    delta_n = c(
      dn_k(v1$d, v1$l), dn_k(v2$d, v2$l), dn_k(vt$d, vt$l),
      dn_mb(vbp$d, vbp$l), dn_r(rg$d, rg$l), "—"
    ),
    delta_pct = c(
      dp(v1$d, v1$l), dp(v2$d, v2$l), dp(vt$d, vt$l),
      dp(vbp$d, vbp$l), dp(rg$d, rg$l), dpp(pr$d, pr$l)
    )
  )
}

smvar_rows <- build_metric_rows(
  bt = "smvar", type_label = "Small\nVariants",
  var1_col = "SNV", var1_label = "SNV",
  var2_col = "INDEL", var2_label = "INDEL",
  delta_ref = "GRCh38", legacy_ver = "v4.2.1"
)

stvar_rows <- build_metric_rows(
  bt = "stvar", type_label = "Structural\nVariants",
  var1_col = "DEL", var1_label = "DEL",
  var2_col = "INS", var2_label = "INS",
  delta_ref = "GRCh37", legacy_ver = "v0.6"
)

table1_df <- bind_rows(smvar_rows, stvar_rows)

n_sm <- nrow(smvar_rows)
smvar_total_rows <- which(smvar_rows$col_b != "")
stvar_total_rows <- which(stvar_rows$col_b != "") + n_sm
non_total_rows   <- setdiff(seq_len(nrow(table1_df)), c(smvar_total_rows, stvar_total_rows))

ft <- flextable::flextable(table1_df) %>%
  flextable::set_header_labels(
    bench_type_label = "Bench Type",
    col_a = "Metric",
    col_b = "",
    grch37 = "GRCh37",
    grch38 = "GRCh38",
    chm13 = "CHM13v2.0",
    delta_n = "n",
    delta_pct = "%"
  ) %>%
  flextable::add_header_row(
    values = c("", "", "v5.0q", "Δ vs Legacy"),
    colwidths = c(1, 2, 3, 2),
    top = TRUE
  ) %>%
  flextable::merge_at(i = 2, j = 2:3, part = "header") %>%
  flextable::merge_v(j = "bench_type_label") %>%
  flextable::merge_at(i = smvar_total_rows, j = 2) %>%
  flextable::merge_at(i = stvar_total_rows, j = 2) %>%
  flextable::align(j = 1, align = "left", part = "body") %>%
  flextable::align(j = 2:3, align = "right", part = "body") %>%
  flextable::align(j = 4:8, align = "right", part = "body") %>%
  flextable::align(j = 4:8, align = "right", part = "header") %>%
  flextable::valign(j = 1:2, valign = "center", part = "body") %>%
  flextable::add_footer_lines(
    paste0(
      "Δ compares v5.0q (GRCh38) vs v4.2.1 (GRCh38) for small variants; ",
      "v5.0q (GRCh37) vs v0.6 (GRCh37) for structural variants. ",
      "pp = percentage points."
    )
  ) %>%
  theme_flextable_manuscript()

for (i in non_total_rows) {
  ft <- flextable::merge_at(ft, i = i, j = 2:3)
}

ft %>% knitr::knit_print()
```

### Supplemental: All References

```{r table_s_smvar}
bench_summary_df %>%
  filter(bench_type == "smvar") %>%
  format_bench_table(var_cols = c("SNV", "INDEL"))
```

```{r table_s_stvar}
bench_summary_df %>%
  filter(bench_type == "stvar") %>%
  format_bench_table(var_cols = c("DEL", "INS"))
```


## Variant Size Distributions

Size distributions for v5.0q compared to legacy benchmarks (GRCh38 only).
Overlaid histograms show the expansion in variant content.

```{r indel_length_distribution}
#| message: false
#| warning: false
smvar_len_df <- variants_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1"),
    bench_type == "smvar",
    var_type == "INDEL",
    ref == "GRCh38"
  ) %>%
  select(bench_version, var_size) %>% 
  mutate(abs_var_size = abs(var_size),
         var_type = if_else(var_size > 0, "Insertion", "Deletion"))

(smvar_len_dist_plt <- ggplot(smvar_len_df, aes(x = abs_var_size, fill = bench_version)) +
  geom_histogram(
    bins = 15,
    position = "dodge",
    color = "grey20"
  ) +
  scale_benchmark_version(aesthetic = "fill", name = "Version") +
  scale_x_continuous(
    breaks = c(1, 5, 10, 15, 20, 30, 40, 50)) +
  labs(x = "Absolute INDEL Length (bp)", y = "Count", fill = "Version") +
  facet_wrap(~var_type, ncol = 2) + 
  theme_manuscript(legend.position = c(0.85, 0.85)))
```

```{r sv_length_distribution}
#| message: false
## Use GRCh37 for SV comparison since v0.6 is GRCh37-only
stvar_len_df <- variants_df %>%
  filter(
    bench_version %in% c("v5.0q", "v0.6"),
    bench_type == "stvar",
    ref == "GRCh37"
  ) %>%
  select(bench_version, var_size) %>%
  filter(abs(var_size) < 10000) %>%  
  mutate(abs_var_size = abs(var_size),
         var_type = if_else(var_size > 0, "Insertion", "Deletion"))

(stvar_len_dist_plt <- ggplot(stvar_len_df, aes(x = abs_var_size, fill = bench_version)) +
  geom_histogram(
    bins = 15,
    position = "dodge",
    color = "grey20"
  ) +
  scale_benchmark_version(aesthetic = "fill", name = "Version") +
  scale_x_log10() +
  labs(x = "Absolute SV Length (bp)", y = "Count", fill = "Version") +
  facet_wrap(~var_type, ncol = 2) + 
  theme_manuscript(legend.position = c(0.85, 0.85)))
```

```{r combined_var_len_dist, width = 4, height = 4}
(combined_var_len_dist_plt <- smvar_len_dist_plt +
  stvar_len_dist_plt +
   plot_layout(ncol = 1, nrow = 2) + 
  plot_annotation(tag_levels = "A") &
  # theme_manuscript(
  #   strip.background = element_rect(fill = "grey92", color = NA),
  #   strip.text = element_text(color = "grey20", face = "bold"),
  #   panel.grid.major = element_line(color = "grey88"),
  #   panel.grid.minor = element_blank()
  # ) +
    theme(
      # axis.text.x = element_text(angle = -45, hjust = 0),
      # legend.position = "bottom",
      plot.margin = unit(c(1, 1, 1, 1), "pt")
    ))
```

```{r var_length_dist_save}
ggsave(
  filename = here("manuscript", "figs", "var_len_dist.png"),
  plot = combined_var_len_dist_plt,
  width = 4,
  height = 4,
  units = "in",
  dpi = 300
)
ggsave(
  filename = here("manuscript", "figs", "var_len_dist.pdf"),
  plot = combined_var_len_dist_plt,
  width = 4,
  height = 4,
  units = "in"
)
```


## Benchmark Region Coverage Change by Chromosome

Relative change in benchmark region size (v5.0q vs legacy) per autosome.
Small variants use GRCh38 (v5.0q vs v4.2.1); structural variants use
GRCh37 (v5.0q vs v0.6) since v0.6 is GRCh37-only.

```{r coverage_change_plot, fig.width = 8, fig.height = 4}
coverage_change_df <- bench_regions_df %>%
  filter(
    bench_version %in% c("v5.0q", "v4.2.1", "v0.6"),
    (bench_type == "smvar" & ref == "GRCh38") |
      (bench_type == "stvar" & ref == "GRCh37"),
    chrom %in% AUTOSOME_CHROM_LEVELS
  ) %>%
  group_by(bench_type, chrom, bench_version) %>%
  summarise(total_size = sum(interval_size), .groups = "drop") %>%
  mutate(
    vers = if_else(bench_version == "v5.0q", "new", "old"),
    chrom = factor(chrom, levels = AUTOSOME_CHROM_LEVELS)
  ) %>%
  select(-bench_version) %>%
  pivot_wider(names_from = vers, values_from = total_size) %>%
  mutate(
    pct_change = 100 * (new - old) / old,
    total_change = new - old
  )

(coverage_change_plt <- coverage_change_df %>%
  ggplot() +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(
    aes(
      x = chrom,
      y = pct_change,
      fill = total_change / 1e6,
      shape = bench_type
    ),
    size = 3,
    color = "black",
    position = position_jitterdodge(jitter.width = 0.1)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    n.breaks = 8
  ) +
  scale_fill_gradient2(
    low = "red", mid = "white", high = "blue",
    midpoint = 0, labels = scales::comma
  ) +
  scale_shape_manual(
    values = c("smvar" = 21, "stvar" = 24),
    labels = c("smvar" = "Small Variants", "stvar" = "Structural Variants")
  ) +
  labs(
    x = "Chromosome",
    y = "Coverage Change (v5.0q vs legacy)",
    fill = "Change (Mb)",
    shape = "Benchmark Type"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0),
    legend.position = "bottom"
  ))

ggsave(
  filename = here("manuscript", "figs", "benchmark_region_size_change.png"),
  plot = coverage_change_plt,
  width = 8,
  height = 4,
  dpi = 300
)
ggsave(
  filename = here("manuscript", "figs", "benchmark_region_size_change.pdf"),
  plot = coverage_change_plt,
  width = 8,
  height = 4
)
```

## Variant Fold Change by Genomic Context

Fold change in variant counts (v5.0q vs legacy) per difficult genomic context.

```{r fold_change_data}
smvar_szbin_order <- c(
  "SNP",
  "[1,5)",
  "[5,10)",
  "[10,15)",
  "[15,20)",
  "[20,30)",
  "[30,40)",
  "[40,50)"
)
stvar_szbin_order <- c(
  "[50,100)",
  "[100,200)",
  "[200,300)",
  "[300,400)",
  "[400,600)",
  "[600,800)",
  "[800,1k)",
  "[1k,2.5k)",
  "[2.5k,5k)",
  ">=5k"
)

context_totals_bytype <- variants_df %>%
  filter(!is.na(context_ids), context_ids != "") %>%
  separate_rows(context_ids, sep = ",") %>%
  mutate(context_ids = str_trim(context_ids)) %>%
  filter(context_ids %in% c("HP", "MAP", "SD", "TR")) %>%
  count(bench_version, ref, bench_type, context_ids, var_type, szbin, name = "count") %>%
  pivot_wider(
    names_from = bench_version,
    values_from = count,
    id_cols = c(ref, bench_type, context_ids, var_type, szbin)
  ) %>%
  filter(
    (bench_type == "smvar" & ref == "GRCh38") |
      (bench_type == "stvar" & ref == "GRCh37")
  ) %>%
  mutate(
    legacy_count = coalesce(v4.2.1, v0.6),
    fold_chg = v5.0q / legacy_count
  ) %>%
  filter(!is.na(fold_chg), is.finite(fold_chg), fold_chg > 0) %>%
  mutate(
    context_label = get_context_labels()[as.character(context_ids)],
    szbin = factor(szbin, levels = c(smvar_szbin_order, stvar_szbin_order)),
    var_type = factor(var_type)
  )
```

```{r fold_change_plot, fig.width = 10, fig.height = 8}
make_fc_panel <- function(df, y_label) {
  ggplot(df, aes(x = szbin, y = fold_chg, color = var_type, shape = var_type)) +
    geom_line(aes(group = var_type), linewidth = 0.4, alpha = 0.6) +
    geom_point(size = 2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
    facet_wrap(~context_label, ncol = 2) +
    scale_y_continuous(
      trans = "log2",
      breaks = c(0.25, 0.5, 1, 2, 4, 8),
      labels = scales::label_number(accuracy = 0.01, drop0trailing = TRUE)
    ) +
    scale_color_brewer(palette = "Set1", name = "Variant Type") +
    scale_shape_discrete(name = "Variant Type") +
    labs(x = "Size Bin", y = y_label) +
    theme_manuscript() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

smvar_fc_plt <- context_totals_bytype %>%
  filter(bench_type == "smvar") %>%
  make_fc_panel("Fold Change (v5.0q / v4.2.1)")

stvar_fc_plt <- context_totals_bytype %>%
  filter(bench_type == "stvar") %>%
  make_fc_panel("Fold Change (v5.0q / v0.6)")

(fold_change_plt <- (smvar_fc_plt + stvar_fc_plt) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom"))
```

## Combined Coverage and Variant Change Figure (Figure 4)

```{r combined_coverage_variant, fig.width = 10, fig.height = 10}
(combined_coverage_variant_plt <- coverage_change_plt /
  fold_change_plt +
  plot_layout(heights = c(1, 2)) &
  theme(plot.margin = unit(c(5, 5, 5, 5), "pt")))

ggsave(
  filename = here("manuscript", "figs", "combined_coverage_variant_change.png"),
  plot = combined_coverage_variant_plt,
  width = 10,
  height = 10,
  dpi = 300
)
ggsave(
  filename = here("manuscript", "figs", "combined_coverage_variant_change.pdf"),
  plot = combined_coverage_variant_plt,
  width = 10,
  height = 10
)
```
