---
title: "Unique Region and Variant Analysis: v5 vs Previous Benchmarks"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(
  load_plot_themes = TRUE
)

library(data.table)
library(yaml)
library(arrow)

source(here::here("R/bed_helpers.R"))
```

## Scope

This notebook compares:

- `v5.0q_GRCh38_smvar` vs `v4.2.1_GRCh38_smvar`
- `v5.0q_GRCh37_stvar` vs `v0.6_GRCh37_stvar`

It summarizes:

- benchmark region bases unique to each version
- variants in version-specific regions, grouped by variant type and size
- for old-only regions, exclusion-overlap reasons and genomic context
- for v5-only regions, variant and genomic-context breakdowns

```{r run-analysis}
data_root <- resolve_data_root()
results_dir <- file.path(data_root, "results")
resources_dir <- file.path(data_root, "resources")
config_path <- here::here("config", "config.yaml")

comparisons <- tribble(
  ~comp_id, ~comparison_label, ~bench_type, ~ref, ~old_benchmark, ~new_benchmark,
  "v5_vs_v4_smvar", "GRCh38 Small Variants (v5 vs v4.2.1)", "smvar", "GRCh38", "v4.2.1_GRCh38_smvar", "v5.0q_GRCh38_smvar",
  "v5_vs_v0.6_stvar", "GRCh37 Structural Variants (v5 vs v0.6)", "stvar", "GRCh37", "v0.6_GRCh37_stvar", "v5.0q_GRCh37_stvar"
)

context_cache <- list(
  GRCh38 = read_context_beds("GRCh38", resources_dir),
  GRCh37 = read_context_beds("GRCh37", resources_dir)
)

comparison_results <- comparisons %>%
  split(.$comp_id) %>%
  purrr::map(~ analyze_comparison(
    .x,
    results_dir = results_dir,
    resources_dir = resources_dir,
    config_path = config_path,
    context_cache = context_cache
  ))

unique_summary_tbl <- bind_rows(purrr::map(comparison_results, "summary"))
old_region_reason_tbl <- bind_rows(purrr::map(comparison_results, "old_region_reason"))
old_exclusion_base_tbl <- bind_rows(purrr::map(comparison_results, "old_exclusion_base"))
old_variant_status_tbl <- bind_rows(purrr::map(comparison_results, "old_variant_status"))
old_variant_exclusion_tbl <- bind_rows(purrr::map(comparison_results, "old_variant_exclusion"))
old_variant_type_size_tbl <- bind_rows(purrr::map(comparison_results, "old_variant_type_size"))
old_excluded_region_context_tbl <- bind_rows(purrr::map(comparison_results, "old_excluded_region_context"))
new_variant_type_size_tbl <- bind_rows(purrr::map(comparison_results, "new_variant_type_size"))
new_variant_context_tbl <- bind_rows(purrr::map(comparison_results, "new_variant_context"))
new_region_context_tbl <- bind_rows(purrr::map(comparison_results, "new_region_context"))
```

## Unique Bases and Variants

```{r}
unique_summary_tbl %>%
  select(
    comparison_label,
    old_benchmark,
    new_benchmark,
    old_only_bp,
    new_only_bp,
    old_only_variants,
    new_only_variants,
    pct_old_only_of_old,
    pct_new_only_of_v5
  ) %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = c("old_only_bp", "new_only_bp", "old_only_variants", "new_only_variants")) %>%
  fmt_number_flextable(columns = c("pct_old_only_of_old", "pct_new_only_of_v5"), decimals = 2) %>%
  cols_label_flextable(
    old_benchmark = "Previous Benchmark",
    new_benchmark = "v5 Benchmark",
    old_only_bp = "Previous-Only Bases",
    new_only_bp = "v5-Only Bases",
    old_only_variants = "Previous-Only Variants",
    new_only_variants = "v5-Only Variants",
    pct_old_only_of_old = "% Previous Benchmark Bases",
    pct_new_only_of_v5 = "% v5 Benchmark Bases"
  ) %>%
  theme_flextable_manuscript()
```

## Previous Benchmark-Specific Regions

### Region Status Relative to v5 Dip and Exclusions

```{r}
old_region_reason_tbl %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = "bases_bp") %>%
  fmt_number_flextable(columns = "pct_of_old_only", decimals = 2) %>%
  cols_label_flextable(
    category = "Category",
    bases_bp = "Bases",
    pct_of_old_only = "% of Previous-Only Bases"
  ) %>%
  theme_flextable_manuscript()
```

### Exclusion Overlap Reasons (Excluded Old-Only Regions)

```{r}
old_exclusion_reason_tbl <- old_exclusion_base_tbl %>%
  full_join(
    old_variant_exclusion_tbl,
    by = c("comparison_id", "comparison_label", "exclusion", "description")
  ) %>%
  mutate(
    bases_bp = coalesce(bases_bp, 0),
    variant_count = coalesce(variant_count, 0L),
    pct_of_excluded_bases = coalesce(pct_of_excluded_bases, 0)
  ) %>%
  arrange(comparison_label, desc(bases_bp), desc(variant_count))

old_exclusion_reason_tbl %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = c("bases_bp", "variant_count")) %>%
  fmt_number_flextable(columns = "pct_of_excluded_bases", decimals = 2) %>%
  cols_label_flextable(
    exclusion = "Exclusion",
    description = "Description",
    bases_bp = "Excluded Bases Overlap",
    variant_count = "Excluded Variants Overlap",
    pct_of_excluded_bases = "% of Excluded Old-Only Bases"
  ) %>%
  theme_flextable_manuscript()
```

### Genomic Context of Excluded Old-Only Regions

```{r}
old_excluded_region_context_tbl %>%
  select(comparison_label, context_name, overlap_bp, pct_of_region) %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = "overlap_bp") %>%
  fmt_number_flextable(columns = "pct_of_region", decimals = 2) %>%
  cols_label_flextable(
    context_name = "Genomic Context",
    overlap_bp = "Overlap Bases",
    pct_of_region = "% of Excluded Old-Only Region Bases"
  ) %>%
  theme_flextable_manuscript()
```

### Previous-Only Variant Breakdown by Type and Size

```{r}
old_variant_type_size_tbl %>%
  select(comparison_label, variant_type, size_bin, variant_count) %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = "variant_count") %>%
  cols_label_flextable(
    variant_type = "Variant Type",
    size_bin = "Size Bin",
    variant_count = "Variant Count"
  ) %>%
  theme_flextable_manuscript()
```

## v5-Specific Regions

### v5-Only Variant Breakdown by Type and Size

```{r}
new_variant_type_size_tbl %>%
  select(comparison_label, variant_type, size_bin, variant_count) %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = "variant_count") %>%
  cols_label_flextable(
    variant_type = "Variant Type",
    size_bin = "Size Bin",
    variant_count = "Variant Count"
  ) %>%
  theme_flextable_manuscript()
```

### Genomic Context of v5-Only Regions

```{r}
new_region_context_tbl %>%
  select(comparison_label, context_name, overlap_bp, pct_of_region) %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = "overlap_bp") %>%
  fmt_number_flextable(columns = "pct_of_region", decimals = 2) %>%
  cols_label_flextable(
    context_name = "Genomic Context",
    overlap_bp = "Overlap Bases",
    pct_of_region = "% of v5-Only Region Bases"
  ) %>%
  theme_flextable_manuscript()
```

### Genomic Context of v5-Only Variants

```{r}
new_variant_context_tbl %>%
  as_grouped_flextable(groupname_col = "comparison_label") %>%
  fmt_integer_flextable(columns = "variant_count") %>%
  cols_label_flextable(
    context_name = "Genomic Context",
    variant_count = "Variant Count"
  ) %>%
  theme_flextable_manuscript()
```
