---
title: "v5.0q Small and Structural Variant External Evaluation"
date: '`r Sys.Date()`'
author: "N.D. Olson"
format: html
editor: source
embed-resources: true
---

Update curations with JZ notes https://docs.google.com/spreadsheets/d/1evd6q8jqF51IM_uX6XrTlc-3f7dDUvurlGI-_vcV6t0/edit?usp=sharing

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Data Loading

### External Evaluation Data

The external evaluation data with JZ curations is loaded from local TSV files
(previously sourced from Google Sheets).

```{r message=FALSE}
## Loading collaborator curations
eval_csvs <- list.files(
  path = here("data/external-evaluations"),
  pattern = "Miqa.csv",
  full.names = TRUE,
  recursive = TRUE,
  include.dirs = TRUE
) %>%
  {
    set_names(., str_extract(., "(?<=evaluations/).*(?= On)"))
  }

## Subsetting small and structural variant tables based on number of rows
## - small variants expect 71, structural variants 41
## a few files have less as less than 5 variants per strata to subset
nlines <- eval_csvs %>% map_int(~ length(read_lines(.x)))

cols_to_keep <- c("callset","correct.in.callset", "correct.in.benchmark", "chrom", "chromStart", "var_type","strata")
smvar_evals <- eval_csvs[nlines > 50] %>%
  map_dfr(read_csv, name_repair = "universal", show_col_types = FALSE, .id = "callset") %>%
  select(all_of(cols_to_keep))

stvar_evals <- eval_csvs[nlines < 50] %>%
  map_dfr(read_csv,
    name_repair = "universal",
    show_col_types = FALSE,
    .id = "callset")


## Load pre-combined evaluation data with JZ curations from local TSV files
smvar_evals_jz <- read_tsv(
  here("data/external-evaluations/Q100-ext-evals-2025-04-11-smvars.tsv"),
  show_col_types = FALSE,
  name_repair = "universal"
)

stvar_evals_jz <- read_tsv(
  here("data/external-evaluations/Q100-ext-evals-2025-04-11-stvars.tsv"),
  show_col_types = FALSE,
  name_repair = "universal"
)

## Manually Revising callset names for figures
stvar_callsets <- c(
  "Baylor-Hifi" = "sniffles2\nHiFi",
  "Baylor-Ont" = "sniffles2\nONT",
  "Comenius-Svdss2" = "svdss2",
  "Illumina-Dragen" = "DRAGEN\nIll",
  "Iter-Dragen" = "DRAGEN\nITER",
  "Pacbio-Sawfish" = "sawfish",
  "Ucla-Delly-Newvcf" = "Delly",
  "Ucla-Manta-Newvcf" = "Manta",
  "Ucla-Ensamble" = "Ensamble"
)

smvar_callsets <- c(
  "Illumina-Dragen" = "DRAGEN",
  "Jhu-Imputefirst" = "Impute",
  "Pacbio-Deepvariant" = "PB-DV",
  "Roche-Neusomatic" = "roche",
  "Ultima-Ug" = "ultima"
)


## Cleaned up tables with updated curations
## Note: TSV files already contain JZ curations, no join needed
smvar_curations <- full_join(smvar_evals, smvar_evals_jz) %>%
  rename(unique_to = label) %>%
  select(
    callset,
    GRCh38_chr, GRCh38_start, GRCh38_end,
    correct.in.benchmark, JZcorrect.in.benchmark,
    correct.in.callset, JZcorrect.in.callset,
    var_type, strata, unique_to
  ) %>%
  mutate(
    curation_benchmark = if_else(!is.na(JZcorrect.in.benchmark),
      JZcorrect.in.benchmark,
      correct.in.benchmark
    ),
    curation_benchmark = factor(curation_benchmark,
      levels = c("no", "unsure", "yes")
    ),
    curation_callset = if_else(!is.na(JZcorrect.in.callset),
      JZcorrect.in.callset,
      correct.in.callset
    ),
    curation_callset = factor(curation_callset,
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  mutate(callset = smvar_callsets[callset])

stvar_curations <- full_join(stvar_evals, stvar_evals_jz) %>%
  rename(var_type = SVTYPE) %>%
  mutate(
    unique_to = if_else(unique_to == "benchmark", "bench", "query"),
    strata = str_glue("{TR_anno}\n{sv_cat}\n{unique_to}")
  ) %>%
  select(
    callset,
    GRCh38_chr, GRCh38_start, GRCh38_end,
    correct.in.benchmark, JZcorrect.in.benchmark,
    correct.in.callset, JZcorrect.in.callset,
    var_type, strata, unique_to, TR_anno, sv_cat, unique_to
  ) %>%
  mutate(
    curation_benchmark = if_else(!is.na(JZcorrect.in.benchmark),
      JZcorrect.in.benchmark,
      correct.in.benchmark
    ),
    curation_benchmark = factor(curation_benchmark,
      levels = c("no", "unsure", "yes")
    ),
    curation_callset = if_else(!is.na(JZcorrect.in.callset),
      JZcorrect.in.callset,
      correct.in.callset
    ),
    curation_callset = factor(curation_callset,
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  mutate(callset = stvar_callsets[callset]) %>% 
  filter(callset != "Manta", callset != "Delly")
```



## Strata Label Mapping

Map S-codes to readable labels for the figures.

```{r strata_labels}
smvar_strata_labels <- c(
  "S01" = "SNV: TR",
  "S02" = "SNV: Low Map",
  "S03" = "SNV: XY nonPAR",
  "S04" = "SNV: Autosome",
  "S05" = "INDEL: HP >6bp",
  "S06" = "INDEL: TR",
  "S07" = "INDEL: Low Map",
  "S08" = "INDEL: XY nonPAR",
  "S09" = "INDEL: Autosome",
  "S10" = "SNV: TR",
  "S11" = "SNV: Low Map",
  "S12" = "SNV: XY nonPAR",
  "S13" = "SNV: Autosome",
  "S14" = "INDEL: HP >6bp",
  "S15" = "INDEL: TR",
  "S16" = "INDEL: Low Map",
  "S17" = "INDEL: XY nonPAR",
  "S18" = "INDEL: Autosome"
)

## Add readable labels to smvar curations
smvar_curations <- smvar_curations %>%
  mutate(
    unique_to = if_else(strata %in% paste0("S0", 1:9), "Query", "Benchmark"),
    strata_label = smvar_strata_labels[strata]
  )
```

## Small Variants External Evaluations

```{r}
total_smvs <- smvar_evals %>% nrow()
smvs_to_review <- smvar_evals %>%
  filter(correct.in.benchmark != "yes") %>%
  nrow()
```

A total of `r total_smvs` small variants were evaluated, of which
`r smvs_to_review` were categorized as not correct or unsure in the benchmark.

```{r smvar_by_callset, fig.width = 7, fig.height = 3.5}
(smvar_eval_callset_plt <- smvar_curations %>%
  count(callset, curation_benchmark) %>%
  ggplot(aes(y = callset, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Callset", y = "# of Variants", fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

```{r smvar_by_callset_cs, fig.width = 7, fig.height = 3.5}
(smvar_eval_callset_cs_plt <- smvar_curations %>%
  count(callset, curation_callset) %>%
  ggplot(aes(y = callset, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Callset", y = "# of Variants", fill = "Correct in\nCallset") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

```{r smvar_by_strata, fig.width = 7, fig.height = 4}
(smvar_eval_strata_plt <- smvar_curations %>%
  count(strata_label, unique_to, curation_benchmark) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  facet_grid(
    rows = vars(unique_to),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = NULL, fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0)
  ))
```

```{r smvar_by_strata_cs, fig.width = 7, fig.height = 4}
(smvar_eval_strata_cs_plt <- smvar_curations %>%
  count(strata_label, unique_to, curation_callset) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  facet_grid(
    rows = vars(unique_to),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = NULL, fill = "Correct in\nCallset") +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0)
  ))
```

### Stratification Flow Diagram (Supplemental)

```{r smvar_flow, fig.width = 8, fig.height = 8}
library(ggflowchart)

edges <- tribble(
  ~from, ~to,
  "Start", "SNV_Branch",
  "Start", "INDEL_Branch",
  "SNV_Branch", "SNV_IsTR",
  "SNV_IsTR", "S01_S10",
  "SNV_IsTR", "SNV_IsDiff",
  "SNV_IsDiff", "S02_S11",
  "SNV_IsDiff", "SNV_IsXY",
  "SNV_IsXY", "S03_S12",
  "SNV_IsXY", "S04_S13",
  "INDEL_Branch", "IND_IsHomo",
  "IND_IsHomo", "S05_S14",
  "IND_IsHomo", "IND_IsTR",
  "IND_IsTR", "S06_S15",
  "IND_IsTR", "IND_IsDiff",
  "IND_IsDiff", "S07_S16",
  "IND_IsDiff", "IND_IsXY",
  "IND_IsXY", "S08_S17",
  "IND_IsXY", "S09_S18"
)

nodes <- tribble(
  ~name, ~label, ~type,
  "Start", "Small Variants", "Root",
  "SNV_Branch", "SNVs", "Branch",
  "INDEL_Branch", "INDELs", "Branch",
  "SNV_IsTR", "in Tandem\nRepeat", "Decision",
  "SNV_IsDiff", "in Low Map\nRegion", "Decision",
  "SNV_IsXY", "XY\nnonPAR", "Decision",
  "IND_IsHomo", "in HP\n(>6bp)", "Decision",
  "IND_IsTR", "in Tandem\nRepeat", "Decision",
  "IND_IsDiff", "in Low Map\nRegion", "Decision",
  "IND_IsXY", "XY\nnonPAR", "Decision",
  "S01_S10", "S01/S10\n(TR)", "Leaf",
  "S02_S11", "S02/S11\n(Low Map)", "Leaf",
  "S03_S12", "S03/S12\n(XY)", "Leaf",
  "S04_S13", "S04/S13\n(Auto)", "Leaf",
  "S05_S14", "S05/S14\n(HP)", "Leaf",
  "S06_S15", "S06/S15\n(TR)", "Leaf",
  "S07_S16", "S07/S16\n(Low Map)", "Leaf",
  "S08_S17", "S08/S17\n(XY)", "Leaf",
  "S09_S18", "S09/S18\n(Auto)", "Leaf"
)

(smvar_flow <- ggflowchart(edges, node_data = nodes, fill = type, text_size = 2.5) +
  scale_fill_manual(values = c(
    "Root" = "#333333", "Branch" = "#777777",
    "Decision" = "#56B4E9", "Leaf" = "#E69F00"
  )) +
  labs(
    title = "Small Variant Stratification Logic",
    caption = "Left: Yes | Right/Down: No. S01-S09 unique to query; S10-S18 unique to benchmark."
  ) +
  theme(legend.title = element_blank()))

ggsave(
  filename = here("manuscript", "figs", "smvar_strat_flow.png"),
  plot = smvar_flow,
  width = 8, height = 8, dpi = 300
)
```


### Structural Variants 

```{r stvar_by_strata, fig.width = 7, fig.height = 4}
## Create readable strata labels for SV
(stvar_eval_strata_plt <- stvar_curations %>%
  mutate(
    strata_label = str_replace_all(strata, "\n", " "),
    strata_label = str_replace(strata_label, "nonTR", "Non-TR"),
    strata_label = str_replace(strata_label, "TR", "TR"),
    curation_benchmark = factor(
      coalesce(as.character(curation_benchmark), correct.in.benchmark),
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  count(strata_label, curation_benchmark) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "# of Variants", y = NULL,
    fill = "Correct in\nBenchmark",
  ) +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r stvar_by_strata_cs, fig.width = 7, fig.height = 4}
(stvar_eval_strata_cs_plt <- stvar_curations %>%
  mutate(
    strata_label = str_replace_all(strata, "\n", " "),
    strata_label = str_replace(strata_label, "nonTR", "Non-TR"),
    strata_label = str_replace(strata_label, "TR", "TR"),
    curation_callset = factor(
      coalesce(as.character(curation_callset), correct.in.callset),
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  count(strata_label, curation_callset) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "# of Variants", y = NULL,
    fill = "Correct in\nCallset",
  ) +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r stvar_by_callset, fig.width = 7, fig.height = 3.5}
(stvar_eval_callset_plt <- stvar_curations %>%
  count(callset, curation_benchmark) %>%
  ggplot(aes(y = callset, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Callset", y = "# of Variants", fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r stvar_by_callset_cs, fig.width = 7, fig.height = 3.5}
(stvar_eval_callset_cs_plt <- stvar_curations %>%
  count(callset, curation_callset) %>%
  ggplot(aes(y = callset, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Callset", y = "# of Variants", fill = "Correct in\nCallset") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

## Combined Evaluation Figure

```{r combined_eval, fig.width = 8, fig.height = 8}
(combined_eval_plt <- (
  smvar_eval_callset_plt + smvar_eval_strata_plt +
    stvar_eval_callset_plt + stvar_eval_strata_plt
) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect", ncol = 2, nrow = 2) &
  theme(legend.position = "bottom"))

ggsave(
  filename = here("manuscript", "figs", "combined_eval.png"),
  plot = combined_eval_plt,
  width = 8, height = 8, dpi = 300
)
```

## Combined Evaluation Figure (Correct in Callset)

```{r combined_eval_cs, fig.width = 8, fig.height = 8}
(combined_eval_cs_plt <- (
  smvar_eval_callset_cs_plt + smvar_eval_strata_cs_plt +
    stvar_eval_callset_cs_plt + stvar_eval_strata_cs_plt
) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect", ncol = 2, nrow = 2) &
  theme(legend.position = "bottom"))

ggsave(
  filename = here("manuscript", "figs", "combined_eval_callset.png"),
  plot = combined_eval_cs_plt,
  width = 8, height = 8, dpi = 300
)
```

## Binomial Model: Probability Discrepancy is Error in Comparison Callset

A binomial GLM estimates the probability that a discrepancy between the
benchmark and a comparison callset is an error in the comparison callset
(i.e., the benchmark is correct). The outcome is binarized from the curation:
`yes` = correct in benchmark vs. `no`/`unsure` = not confirmed correct.
Separate models are fit for small and structural variants. Because the goal
is to establish a lower bound on benchmark reliability, one-sided 95% lower
confidence bounds are reported.

A sensitivity analysis treats "unsure" curations as correct in the benchmark
(combining with "yes") rather than incorrect (combining with "no"), providing
an upper range on the estimated probabilities.


```{r ci_data_prep}
library(emmeans)
library(brglm2)

## Shared feature preparation
prepare_smvar_features <- function(df) {
  df |>
    mutate(
      genomic_context = str_extract(strata_label, "(?<=: ).*"),
      genomic_context = factor(genomic_context,
                               levels = c("Autosome", "TR", "HP >6bp",
                                          "Low Map", "XY nonPAR")),
      genomic_context = droplevels(genomic_context),
      var_type = factor(var_type, levels = c("SNP", "INDEL")),
      unique_to = factor(unique_to, levels = c("Query", "Benchmark"))
    )
}

prepare_stvar_features <- function(df) {
  df |>
    mutate(
      TR_anno = factor(TR_anno, levels = c("nonTR", "TR")),
      sv_cat = factor(sv_cat, levels = c("DEL", "INS")),
      unique_to = factor(unique_to, levels = c("query", "bench"))
    )
}

## Primary analysis: unsure = incorrect (conservative)
smvar_model_df <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark == "yes")) |>
  prepare_smvar_features()

stvar_model_df <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark == "yes")) |>
  prepare_stvar_features()

## Sensitivity analysis: unsure = correct
smvar_sens_df <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark %in% c("yes", "unsure"))) |>
  prepare_smvar_features()

stvar_sens_df <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark %in% c("yes", "unsure"))) |>
  prepare_stvar_features()
```

```{r ci_models}
## Primary models (standard GLM)
smvar_glm <- glm(correct_binary ~ var_type + genomic_context + unique_to,
                 family = binomial, data = smvar_model_df)
stvar_glm <- glm(correct_binary ~ TR_anno + sv_cat + unique_to,
                 family = binomial, data = stvar_model_df)

## Sensitivity models
## smvar sensitivity has quasi-complete separation (only 5 "no" curations
## with 0 in the Low Map context), so bias-reduced estimation is used.
smvar_sens_glm <- glm(correct_binary ~ var_type + genomic_context + unique_to,
                      family = binomial, data = smvar_sens_df,
                      method = "brglmFit")
stvar_sens_glm <- glm(correct_binary ~ TR_anno + sv_cat + unique_to,
                      family = binomial, data = stvar_sens_df)
```

```{r ci_emmeans}
## Helper to extract emmeans with one-sided 95% lower bound
extract_emmeans <- function(em, benchmarkset, feature_name) {
  as_tibble(summary(em, side = ">")) |>
    mutate(
      benchmarkset = benchmarkset,
      feature = feature_name,
      level = if ("1" %in% names(pick(everything()))) "Overall"
              else as.character(pick(1)[[1]])
    ) |>
    select(benchmarkset, feature, level, prob, SE, asymp.LCL)
}

## Compute emmeans for all four models
compute_all_emmeans <- function(smvar_fit, stvar_fit) {
  bind_rows(
    extract_emmeans(
      emmeans(smvar_fit, ~ 1, type = "response", weights = "proportional"),
      "Small Variant", "Overall"),
    extract_emmeans(
      emmeans(smvar_fit, ~ var_type, type = "response", weights = "proportional"),
      "Small Variant", "Variant Type"),
    extract_emmeans(
      emmeans(smvar_fit, ~ genomic_context, type = "response", weights = "proportional"),
      "Small Variant", "Genomic Context"),
    extract_emmeans(
      emmeans(smvar_fit, ~ unique_to, type = "response", weights = "proportional"),
      "Small Variant", "Unique To"),
    extract_emmeans(
      emmeans(stvar_fit, ~ 1, type = "response", weights = "proportional"),
      "Structural Variant", "Overall"),
    extract_emmeans(
      emmeans(stvar_fit, ~ TR_anno, type = "response", weights = "proportional"),
      "Structural Variant", "TR Annotation"),
    extract_emmeans(
      emmeans(stvar_fit, ~ sv_cat, type = "response", weights = "proportional"),
      "Structural Variant", "SV Category"),
    extract_emmeans(
      emmeans(stvar_fit, ~ unique_to, type = "response", weights = "proportional"),
      "Structural Variant", "Unique To")
  )
}

primary_results <- compute_all_emmeans(smvar_glm, stvar_glm) |>
  mutate(analysis = "Conservative (unsure = incorrect)")
sens_results <- compute_all_emmeans(smvar_sens_glm, stvar_sens_glm) |>
  mutate(analysis = "Sensitive (unsure = correct)")

ci_results <- bind_rows(primary_results, sens_results)
```

```{r ci_forest_plot, fig.width = 7, fig.height = 5.5}
## Prepare display labels
ci_plot_df <- ci_results |>
  mutate(
    label = if_else(feature == "Overall", "Overall", level),
    label = case_when(
      label == "nonTR" ~ "Non-TR",
      label == "bench" ~ "Benchmark",
      label == "query" ~ "Query",
      .default = label
    ),
    feature = factor(feature,
      levels = rev(c("Overall", "Variant Type", "Genomic Context",
                      "Unique To", "TR Annotation", "SV Category"))
    ),
    benchmarkset = factor(benchmarkset,
      levels = c("Small Variant", "Structural Variant"))
  )

(ci_forest_plt <- ggplot(ci_plot_df,
       aes(x = prob, y = reorder(label, prob),
           xmin = asymp.LCL, xmax = prob,
           color = analysis, shape = analysis)) +
  geom_pointrange(size = 0.4,
                  position = position_dodge(width = 0.5)) +
  facet_grid(
    rows = vars(feature),
    cols = vars(benchmarkset),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0.6, 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Probability Discrepancy is Error in Comparison Callset",
    y = NULL,
    color = "Unsure Classification",
    shape = "Unsure Classification"
  ) +
  theme_manuscript() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "bottom"
  ))
```


```{r}
ggsave(
  filename = here("manuscript", "figs", "ride_estimates.png"),
  plot = ci_forest_plt,
  width = 8, height = 8, dpi = 300
)
```

```{r ci_summary_table}
## Per-level sample sizes
smvar_n <- smvar_model_df |>
  summarise(
    Overall = n(), SNP = sum(var_type == "SNP"),
    INDEL = sum(var_type == "INDEL"),
    Autosome = sum(genomic_context == "Autosome"),
    TR = sum(genomic_context == "TR"),
    `HP >6bp` = sum(genomic_context == "HP >6bp"),
    `Low Map` = sum(genomic_context == "Low Map"),
    Query = sum(unique_to == "Query"),
    Benchmark = sum(unique_to == "Benchmark")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Small Variant")

stvar_n <- stvar_model_df |>
  summarise(
    Overall = n(), `Non-TR` = sum(TR_anno == "nonTR"),
    TR = sum(TR_anno == "TR"), DEL = sum(sv_cat == "DEL"),
    INS = sum(sv_cat == "INS"), Query = sum(unique_to == "query"),
    Benchmark = sum(unique_to == "bench")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Structural Variant")

n_lookup <- bind_rows(smvar_n, stvar_n)

## Build table with both analyses side by side
ci_table_wide <- ci_results |>
  mutate(
    level = case_when(
      feature == "Overall" ~ "",
      level == "nonTR" ~ "Non-TR",
      level == "bench" ~ "Benchmark",
      level == "query" ~ "Query",
      .default = level
    )
  ) |>
  select(benchmarkset, feature, level, analysis, prob, asymp.LCL) |>
  pivot_wider(
    names_from = analysis,
    values_from = c(prob, asymp.LCL),
    names_glue = "{analysis}|{.value}"
  ) |>
  left_join(n_lookup, by = c("benchmarkset", "level"))

ci_table_wide |>
  select(
    benchmarkset, feature, level, n,
    `Conservative (unsure = incorrect)|prob`,
    `Conservative (unsure = incorrect)|asymp.LCL`,
    `Sensitive (unsure = correct)|prob`,
    `Sensitive (unsure = correct)|asymp.LCL`
  ) |>
  gt(groupname_col = "benchmarkset", rowname_col = "level") |>
  cols_label(
    feature = "Feature", n = "N",
    `Conservative (unsure = incorrect)|prob` = "Est. (%)",
    `Conservative (unsure = incorrect)|asymp.LCL` = "LCB (%)",
    `Sensitive (unsure = correct)|prob` = "Est. (%)",
    `Sensitive (unsure = correct)|asymp.LCL` = "LCB (%)"
  ) |>
  cols_align(align = "center",
             columns = -c(feature, level)) |>
  cols_align(align = "left", columns = feature) |>
  fmt_percent(
    columns = c(
      `Conservative (unsure = incorrect)|prob`,
      `Conservative (unsure = incorrect)|asymp.LCL`,
      `Sensitive (unsure = correct)|prob`,
      `Sensitive (unsure = correct)|asymp.LCL`
    ),
    decimals = 1
  ) |>
  tab_spanner(
    label = "Conservative (unsure = incorrect)",
    columns = c(`Conservative (unsure = incorrect)|prob`,
                `Conservative (unsure = incorrect)|asymp.LCL`)
  ) |>
  tab_spanner(
    label = "Sensitive (unsure = correct)",
    columns = c(`Sensitive (unsure = correct)|prob`,
                `Sensitive (unsure = correct)|asymp.LCL`)
  ) |>
  tab_header(
    title = "Estimated Probability That a Discrepancy is an Error in the Comparison Callset",
    subtitle = "Marginal estimates from binomial GLMs with one-sided 95% lower confidence bounds (LCB)"
  ) |>
  tab_footnote(
    footnote = "Marginal probabilities via emmeans with proportional weighting, averaging over other model terms.",
    locations = cells_column_spanners()
  ) |>
  tab_footnote(
    footnote = "HP >6bp stratum applies only to INDELs.",
    locations = cells_body(columns = feature, rows = level == "HP >6bp")
  ) |>
  tab_footnote(
    footnote = "Small variant sensitivity model fit with bias-reduced logistic regression (brglm2) due to quasi-complete separation.",
    locations = cells_column_spanners(
      spanners = "Sensitive (unsure = correct)")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = everything(), rows = feature == "Overall")
  )
```

## Binomial Model: Benchmark Identifies Errors in Comparison Callset (Callset-Based)

This analysis uses the `correct.in.callset` curation to estimate the
probability that a discrepancy reflects an error in the comparison callset.
When the comparison callset is judged "not correct" (`no`), the benchmark has
identified a real error — even if the benchmark's own call at that site is
also imperfect. This captures the broader utility of the benchmark for
flagging comparison callset errors.

The outcome is inverted relative to the curation values: `no` = the
comparison is wrong (success), `yes` = the comparison is correct (failure).
The conservative analysis treats `unsure` as the comparison being correct
(fewer successes); the sensitivity analysis treats `unsure` as the comparison
being wrong (more successes).

```{r cs_data_prep}
## Primary analysis: unsure = callset correct (conservative)
smvar_model_df_cs <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_callset == "no")) |>
  prepare_smvar_features()

stvar_model_df_cs <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_callset == "no")) |>
  prepare_stvar_features()

## Sensitivity analysis: unsure = callset incorrect
smvar_sens_df_cs <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_callset %in% c("no", "unsure"))) |>
  prepare_smvar_features()

stvar_sens_df_cs <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_callset %in% c("no", "unsure"))) |>
  prepare_stvar_features()
```

```{r cs_models}
## Primary models
smvar_glm_cs <- glm(correct_binary ~ var_type + genomic_context + unique_to,
                    family = binomial, data = smvar_model_df_cs)
stvar_glm_cs <- glm(correct_binary ~ TR_anno + sv_cat + unique_to,
                    family = binomial, data = stvar_model_df_cs)

## Sensitivity models — use brglmFit if quasi-complete separation is detected
smvar_sens_glm_cs <- tryCatch(
  glm(correct_binary ~ var_type + genomic_context + unique_to,
      family = binomial, data = smvar_sens_df_cs),
  warning = function(w) {
    glm(correct_binary ~ var_type + genomic_context + unique_to,
        family = binomial, data = smvar_sens_df_cs,
        method = "brglmFit")
  }
)
stvar_sens_glm_cs <- tryCatch(
  glm(correct_binary ~ TR_anno + sv_cat + unique_to,
      family = binomial, data = stvar_sens_df_cs),
  warning = function(w) {
    glm(correct_binary ~ TR_anno + sv_cat + unique_to,
        family = binomial, data = stvar_sens_df_cs,
        method = "brglmFit")
  }
)
```

```{r cs_emmeans}
primary_results_cs <- compute_all_emmeans(smvar_glm_cs, stvar_glm_cs) |>
  mutate(analysis = "Conservative (unsure = callset correct)")
sens_results_cs <- compute_all_emmeans(smvar_sens_glm_cs, stvar_sens_glm_cs) |>
  mutate(analysis = "Sensitive (unsure = callset incorrect)")

ci_results_cs <- bind_rows(primary_results_cs, sens_results_cs)
```

```{r cs_forest_plot, fig.width = 7, fig.height = 5.5}
ci_plot_df_cs <- ci_results_cs |>
  mutate(
    label = if_else(feature == "Overall", "Overall", level),
    label = case_when(
      label == "nonTR" ~ "Non-TR",
      label == "bench" ~ "Benchmark",
      label == "query" ~ "Query",
      .default = label
    ),
    feature = factor(feature,
      levels = rev(c("Overall", "Variant Type", "Genomic Context",
                      "Unique To", "TR Annotation", "SV Category"))
    ),
    benchmarkset = factor(benchmarkset,
      levels = c("Small Variant", "Structural Variant"))
  )

(ci_forest_cs_plt <- ggplot(ci_plot_df_cs,
       aes(x = prob, y = reorder(label, prob),
           xmin = asymp.LCL, xmax = prob,
           color = analysis, shape = analysis)) +
  geom_pointrange(size = 0.4,
                  position = position_dodge(width = 0.5)) +
  facet_grid(
    rows = vars(feature),
    cols = vars(benchmarkset),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0.6, 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Probability Benchmark Identifies Error in Comparison Callset",
    y = NULL,
    color = "Unsure Classification",
    shape = "Unsure Classification"
  ) +
  theme_manuscript() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "bottom"
  ))
```

```{r}
ggsave(
  filename = here("manuscript", "figs", "ride_estimates_callset.png"),
  plot = ci_forest_cs_plt,
  width = 8, height = 8, dpi = 300
)
```

```{r cs_summary_table}
## Per-level sample sizes (same as benchmark analysis)
smvar_n_cs <- smvar_model_df_cs |>
  summarise(
    Overall = n(), SNP = sum(var_type == "SNP"),
    INDEL = sum(var_type == "INDEL"),
    Autosome = sum(genomic_context == "Autosome"),
    TR = sum(genomic_context == "TR"),
    `HP >6bp` = sum(genomic_context == "HP >6bp"),
    `Low Map` = sum(genomic_context == "Low Map"),
    Query = sum(unique_to == "Query"),
    Benchmark = sum(unique_to == "Benchmark")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Small Variant")

stvar_n_cs <- stvar_model_df_cs |>
  summarise(
    Overall = n(), `Non-TR` = sum(TR_anno == "nonTR"),
    TR = sum(TR_anno == "TR"), DEL = sum(sv_cat == "DEL"),
    INS = sum(sv_cat == "INS"), Query = sum(unique_to == "query"),
    Benchmark = sum(unique_to == "bench")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Structural Variant")

n_lookup_cs <- bind_rows(smvar_n_cs, stvar_n_cs)

ci_table_wide_cs <- ci_results_cs |>
  mutate(
    level = case_when(
      feature == "Overall" ~ "",
      level == "nonTR" ~ "Non-TR",
      level == "bench" ~ "Benchmark",
      level == "query" ~ "Query",
      .default = level
    )
  ) |>
  select(benchmarkset, feature, level, analysis, prob, asymp.LCL) |>
  pivot_wider(
    names_from = analysis,
    values_from = c(prob, asymp.LCL),
    names_glue = "{analysis}|{.value}"
  ) |>
  left_join(n_lookup_cs, by = c("benchmarkset", "level"))

ci_table_wide_cs |>
  select(
    benchmarkset, feature, level, n,
    `Conservative (unsure = callset correct)|prob`,
    `Conservative (unsure = callset correct)|asymp.LCL`,
    `Sensitive (unsure = callset incorrect)|prob`,
    `Sensitive (unsure = callset incorrect)|asymp.LCL`
  ) |>
  gt(groupname_col = "benchmarkset", rowname_col = "level") |>
  cols_label(
    feature = "Feature", n = "N",
    `Conservative (unsure = callset correct)|prob` = "Est. (%)",
    `Conservative (unsure = callset correct)|asymp.LCL` = "LCB (%)",
    `Sensitive (unsure = callset incorrect)|prob` = "Est. (%)",
    `Sensitive (unsure = callset incorrect)|asymp.LCL` = "LCB (%)"
  ) |>
  cols_align(align = "center",
             columns = -c(feature, level)) |>
  cols_align(align = "left", columns = feature) |>
  fmt_percent(
    columns = c(
      `Conservative (unsure = callset correct)|prob`,
      `Conservative (unsure = callset correct)|asymp.LCL`,
      `Sensitive (unsure = callset incorrect)|prob`,
      `Sensitive (unsure = callset incorrect)|asymp.LCL`
    ),
    decimals = 1
  ) |>
  tab_spanner(
    label = "Conservative (unsure = callset correct)",
    columns = c(`Conservative (unsure = callset correct)|prob`,
                `Conservative (unsure = callset correct)|asymp.LCL`)
  ) |>
  tab_spanner(
    label = "Sensitive (unsure = callset incorrect)",
    columns = c(`Sensitive (unsure = callset incorrect)|prob`,
                `Sensitive (unsure = callset incorrect)|asymp.LCL`)
  ) |>
  tab_header(
    title = "Estimated Probability That Benchmark Identifies an Error in the Comparison Callset",
    subtitle = "Based on correct-in-callset curations; marginal estimates from binomial GLMs with one-sided 95% LCB"
  ) |>
  tab_footnote(
    footnote = "Marginal probabilities via emmeans with proportional weighting, averaging over other model terms.",
    locations = cells_column_spanners()
  ) |>
  tab_footnote(
    footnote = "HP >6bp stratum applies only to INDELs.",
    locations = cells_body(columns = feature, rows = level == "HP >6bp")
  ) |>
  tab_footnote(
    footnote = "Bias-reduced logistic regression (brglm2) used where quasi-complete separation was detected.",
    locations = cells_column_spanners(
      spanners = "Sensitive (unsure = callset incorrect)")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = everything(), rows = feature == "Overall")
  )
```

### Methods: External Evaluation Confidence Intervals

To estimate the probability that a discrepancy between the benchmark set and
a comparison callset represents an error in the comparison callset (i.e., the
benchmark is correct), we fit separate binomial generalized linear models
(GLMs) with a logit link for small variants and structural variants.

For each comparison callset, five discrepancies per stratum were randomly
selected for manual curation. Each discrepancy was classified as "correct in
benchmark" (yes), "not correct in benchmark" (no), or "unsure." A subset of
curations were independently re-reviewed by a second expert (JZ); where
available, the second reviewer's assessment was used as the final curation.
The primary analysis binarized the outcome conservatively: "yes" was treated
as a success (the discrepancy is an error in the comparison callset), while
"no" and "unsure" were combined as failures. A sensitivity analysis treated
"unsure" as successes (combining with "yes"), providing an upper range for
the estimated probabilities.

For small variants (N = `r nrow(smvar_model_df)`;
`r length(unique(smvar_model_df$callset))` callsets), the model included
three additive predictors: variant type (SNP vs. INDEL), genomic context
(Autosome, tandem repeat, homopolymer >6 bp, or low mappability region),
and whether the discrepancy was unique to the query callset or the benchmark.
For structural variants (N = `r nrow(stvar_model_df)`;
`r length(unique(stvar_model_df$callset))` callsets), predictors were tandem
repeat annotation (TR vs. non-TR), SV category (DEL vs. INS), and unique-to
status (query vs. benchmark). All predictors were entered as main effects
without interactions. Because different sets of variants were curated for
each callset, each observation represents an independent evaluation at a
distinct genomic position, and callset was not included as a model term.

Because the goal is to establish a lower bound on how reliably the benchmark
identifies errors in comparison callsets, one-sided 95% lower confidence
bounds were reported rather than two-sided intervals. Marginal probabilities
and confidence bounds on the probability scale were obtained using the
`emmeans` R package with proportional weighting to average over the other
model terms. Proportional weighting ensures that factor levels are weighted
by their observed frequencies in the data, which is relevant because the
homopolymer >6 bp context applies only to INDELs. In the sensitivity
analysis, the small variant model was fit using bias-reduced logistic
regression (Firth penalization via the `brglm2` R package) because treating
"unsure" as correct left only 5 failures across 350 observations, causing
quasi-complete separation in the low mappability stratum.

**Assumptions.** The binomial GLM assumes that curations are independent
Bernoulli trials, that the log-odds of the outcome are a linear additive
function of the predictors, and that the effect of each predictor is
constant across levels of the other predictors (no interactions).
Independence is reasonable given that each observation is a unique genomic
position evaluated by a single curator. The additive assumption implies,
for example, that the difference in log-odds between TR and autosomal
contexts is the same for SNPs and INDELs. Given the sample sizes, there was
insufficient power to test interactions. The one-sided Wald confidence bounds
rely on asymptotic normality of the maximum likelihood estimates, which is
generally adequate for these sample sizes but may be slightly liberal for
cells with very high observed proportions.

**Callset-based complementary analysis.** A parallel set of binomial GLMs
was fit using the "correct in callset" curation as the outcome variable. In
this analysis, a discrepancy is classified as a success (benchmark identified
an error) when the comparison callset is judged "not correct" (`no`). This
perspective captures cases where both the benchmark and comparison callset
may be imperfect at a site, but the benchmark nonetheless correctly flagged
an error in the comparison. The conservative analysis treats "unsure" as the
comparison being correct (fewer successes); the sensitivity analysis treats
"unsure" as the comparison being wrong (more successes). The same model
structure, predictors, and emmeans extraction were used. Bias-reduced
logistic regression was applied where quasi-complete separation was detected.

**Limitations.** The XY non-PAR strata had no discrepancies for small
variants and were excluded from the small variant model. The treatment of
"unsure" curations has a notable effect on the results, particularly for
structural variants where "unsure" curations were more common; the
sensitivity analysis brackets this uncertainty. The models do not include
callset-specific effects, so the estimated probabilities reflect an average
across all comparison callsets evaluated. The curated discrepancies were
selected from specific callsets and may not be representative of all possible
discrepancies. Finally, the additive model does not capture potential
interactions between predictors (e.g., whether TR effects differ for INS
vs. DEL in structural variants); the data would support such extensions but
at the cost of wider confidence intervals.



