---
title: "v5.0q Small and Structural Variant External Evaluation"
date: '`r Sys.Date()`'
author: "N.D. Olson"
format: html
editor: source
embed-resources: true
---

__TODO__: linear model to get confidence intervals for curations

Update curations with JZ notes https://docs.google.com/spreadsheets/d/1evd6q8jqF51IM_uX6XrTlc-3f7dDUvurlGI-_vcV6t0/edit?usp=sharing

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Data Loading

### External Evaluation Data

The external evaluation data with JZ curations is loaded from local TSV files
(previously sourced from Google Sheets).

```{r message=FALSE}
## Loading collaborator curations
eval_csvs <- list.files(
  path = here("data/external-evaluations"),
  pattern = "Miqa.csv",
  full.names = TRUE,
  recursive = TRUE,
  include.dirs = TRUE
) %>%
  {
    set_names(., str_extract(., "(?<=evaluations/).*(?= On)"))
  }

## Subsetting small and structural variant tables based on number of rows
## - small variants expect 71, structural variants 41
## a few files have less as less than 5 variants per strata to subset
nlines <- eval_csvs %>% map_int(~ length(read_lines(.x)))

cols_to_keep <- c("callset","correct.in.callset", "correct.in.benchmark", "chrom", "chromStart", "var_type","strata")
smvar_evals <- eval_csvs[nlines > 50] %>%
  map_dfr(read_csv, name_repair = "universal", show_col_types = FALSE, .id = "callset") %>%
  select(all_of(cols_to_keep))

stvar_evals <- eval_csvs[nlines < 50] %>%
  map_dfr(read_csv,
    name_repair = "universal",
    show_col_types = FALSE,
    .id = "callset")


## Load pre-combined evaluation data with JZ curations from local TSV files
smvar_evals_jz <- read_tsv(
  here("data/external-evaluations/Q100-ext-evals-2025-04-11-smvars.tsv"),
  show_col_types = FALSE,
  name_repair = "universal"
)

stvar_evals_jz <- read_tsv(
  here("data/external-evaluations/Q100-ext-evals-2025-04-11-stvars.tsv"),
  show_col_types = FALSE,
  name_repair = "universal"
)

## Manually Revising callset names for figures
stvar_callsets <- c(
  "Baylor-Hifi" = "sniffles2\nHiFi",
  "Baylor-Ont" = "sniffles2\nONT",
  "Comenius-Svdss2" = "svdss2",
  "Illumina-Dragen" = "DRAGEN\nIll",
  "Iter-Dragen" = "DRAGEN\nITER",
  "Pacbio-Sawfish" = "sawfish",
  "Ucla-Delly-Newvcf" = "Delly",
  "Ucla-Manta-Newvcf" = "Manta",
  "Ucla-Ensamble" = "Ensamble"
)

smvar_callsets <- c(
  "Illumina-Dragen" = "DRAGEN",
  "Jhu-Imputefirst" = "Impute",
  "Pacbio-Deepvariant" = "PB-DV",
  "Roche-Neusomatic" = "roche",
  "Ultima-Ug" = "ultima"
)


## Cleaned up tables with updated curations
## Note: TSV files already contain JZ curations, no join needed
smvar_curations <- full_join(smvar_evals, smvar_evals_jz) %>%
  rename(unique_to = label) %>%
  select(
    callset,
    GRCh38_chr, GRCh38_start, GRCh38_end,
    correct.in.benchmark, JZcorrect.in.benchmark,
    var_type, strata, unique_to
  ) %>%
  mutate(
    curation = if_else(!is.na(JZcorrect.in.benchmark),
      JZcorrect.in.benchmark,
      correct.in.benchmark
    ),
    curation = factor(curation,
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  mutate(callset = smvar_callsets[callset])

stvar_curations <- full_join(stvar_evals, stvar_evals_jz) %>%
  rename(var_type = SVTYPE) %>%
  mutate(
    unique_to = if_else(unique_to == "benchmark", "bench", "query"),
    strata = str_glue("{TR_anno}\n{sv_cat}\n{unique_to}")
  ) %>%
  select(
    callset,
    GRCh38_chr, GRCh38_start, GRCh38_end,
    correct.in.benchmark, JZcorrect.in.benchmark,
    var_type, strata, unique_to
  ) %>%
  mutate(
    curation = if_else(!is.na(JZcorrect.in.benchmark),
      JZcorrect.in.benchmark,
      correct.in.benchmark
    ),
    curation = factor(curation,
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  mutate(callset = stvar_callsets[callset])
```



## Strata Label Mapping

Map S-codes to readable labels for the figures.

```{r strata_labels}
smvar_strata_labels <- c(
  "S01" = "SNV: TR",
  "S02" = "SNV: Low Map",
  "S03" = "SNV: XY nonPAR",
  "S04" = "SNV: Autosome",
  "S05" = "INDEL: HP >6bp",
  "S06" = "INDEL: TR",
  "S07" = "INDEL: Low Map",
  "S08" = "INDEL: XY nonPAR",
  "S09" = "INDEL: Autosome",
  "S10" = "SNV: TR",
  "S11" = "SNV: Low Map",
  "S12" = "SNV: XY nonPAR",
  "S13" = "SNV: Autosome",
  "S14" = "INDEL: HP >6bp",
  "S15" = "INDEL: TR",
  "S16" = "INDEL: Low Map",
  "S17" = "INDEL: XY nonPAR",
  "S18" = "INDEL: Autosome"
)

## Add readable labels to smvar curations
smvar_curations <- smvar_curations %>%
  mutate(
    unique_to = if_else(strata %in% paste0("S0", 1:9), "Query", "Benchmark"),
    strata_label = smvar_strata_labels[strata]
  )
```

## Small Variants External Evaluations

```{r}
total_smvs <- smvar_evals %>% nrow()
smvs_to_review <- smvar_evals %>%
  filter(correct.in.benchmark != "yes") %>%
  nrow()
```

A total of `r total_smvs` small variants were evaluated, of which
`r smvs_to_review` were categorized as not correct or unsure in the benchmark.

```{r smvar_by_callset, fig.width = 7, fig.height = 3.5}
(smvar_eval_callset_plt <- smvar_curations %>%
  count(callset, curation) %>%
  ggplot(aes(x = callset, y = n, fill = curation)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Callset", y = "# of Variants", fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

```{r smvar_by_strata, fig.width = 7, fig.height = 4}
(smvar_eval_strata_plt <- smvar_curations %>%
  count(strata_label, unique_to, curation) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  facet_grid(
    rows = vars(unique_to),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = NULL, fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(
    legend.position = "right",
    strip.text.y = element_text(angle = 0)
  ))
```

### Stratification Flow Diagram (Supplemental)

```{r smvar_flow, fig.width = 8, fig.height = 8}
library(ggflowchart)

edges <- tribble(
  ~from, ~to,
  "Start", "SNV_Branch",
  "Start", "INDEL_Branch",
  "SNV_Branch", "SNV_IsTR",
  "SNV_IsTR", "S01_S10",
  "SNV_IsTR", "SNV_IsDiff",
  "SNV_IsDiff", "S02_S11",
  "SNV_IsDiff", "SNV_IsXY",
  "SNV_IsXY", "S03_S12",
  "SNV_IsXY", "S04_S13",
  "INDEL_Branch", "IND_IsHomo",
  "IND_IsHomo", "S05_S14",
  "IND_IsHomo", "IND_IsTR",
  "IND_IsTR", "S06_S15",
  "IND_IsTR", "IND_IsDiff",
  "IND_IsDiff", "S07_S16",
  "IND_IsDiff", "IND_IsXY",
  "IND_IsXY", "S08_S17",
  "IND_IsXY", "S09_S18"
)

nodes <- tribble(
  ~name, ~label, ~type,
  "Start", "Small Variants", "Root",
  "SNV_Branch", "SNVs", "Branch",
  "INDEL_Branch", "INDELs", "Branch",
  "SNV_IsTR", "in Tandem\nRepeat", "Decision",
  "SNV_IsDiff", "in Low Map\nRegion", "Decision",
  "SNV_IsXY", "XY\nnonPAR", "Decision",
  "IND_IsHomo", "in HP\n(>6bp)", "Decision",
  "IND_IsTR", "in Tandem\nRepeat", "Decision",
  "IND_IsDiff", "in Low Map\nRegion", "Decision",
  "IND_IsXY", "XY\nnonPAR", "Decision",
  "S01_S10", "S01/S10\n(TR)", "Leaf",
  "S02_S11", "S02/S11\n(Low Map)", "Leaf",
  "S03_S12", "S03/S12\n(XY)", "Leaf",
  "S04_S13", "S04/S13\n(Auto)", "Leaf",
  "S05_S14", "S05/S14\n(HP)", "Leaf",
  "S06_S15", "S06/S15\n(TR)", "Leaf",
  "S07_S16", "S07/S16\n(Low Map)", "Leaf",
  "S08_S17", "S08/S17\n(XY)", "Leaf",
  "S09_S18", "S09/S18\n(Auto)", "Leaf"
)

(smvar_flow <- ggflowchart(edges, node_data = nodes, fill = type, text_size = 2.5) +
  scale_fill_manual(values = c(
    "Root" = "#333333", "Branch" = "#777777",
    "Decision" = "#56B4E9", "Leaf" = "#E69F00"
  )) +
  labs(
    title = "Small Variant Stratification Logic",
    caption = "Left: Yes | Right/Down: No. S01-S09 unique to query; S10-S18 unique to benchmark."
  ) +
  theme(legend.title = element_blank()))

ggsave(
  filename = here("manuscript", "figs", "smvar_strat_flow.png"),
  plot = smvar_flow,
  width = 8, height = 8, dpi = 300
)
```


### Structural Variants External Evaluations (UCLA Ensemble Only)

```{r stvar_ensemble_filter}
## Filter to UCLA Ensemble callset only
stvar_curations_ensemble <- stvar_curations %>%
  filter(callset == "Ensamble")
```

```{r stvar_by_strata, fig.width = 7, fig.height = 4}
## Create readable strata labels for SV
(stvar_eval_strata_plt <- stvar_curations_ensemble %>%
  mutate(
    strata_label = str_replace_all(strata, "\n", " "),
    strata_label = str_replace(strata_label, "nonTR", "Non-TR"),
    strata_label = str_replace(strata_label, "TR", "TR"),
    curation = factor(
      coalesce(as.character(curation), correct.in.benchmark),
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  count(strata_label, curation) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "# of Variants", y = NULL,
    fill = "Correct in\nBenchmark",
    title = "UCLA Ensemble SV Callset"
  ) +
  theme_manuscript() +
  theme(legend.position = "right"))

ggsave(
  filename = here("manuscript", "figs", "stvar_eval_strata_supplemental.png"),
  plot = stvar_eval_strata_plt,
  width = 7, height = 4, dpi = 300
)
```


## Combined Evaluation Figure

```{r combined_eval, fig.width = 8, fig.height = 8}
(combined_eval_plt <- (
  smvar_eval_callset_plt + smvar_eval_strata_plt +
    stvar_eval_strata_plt + plot_spacer()
) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect", ncol = 2, nrow = 2) &
  theme(legend.position = "bottom"))

ggsave(
  filename = here("manuscript", "figs", "combined_eval.png"),
  plot = combined_eval_plt,
  width = 8, height = 8, dpi = 300
)
```

