---
title: "v5.0q Small and Structural Variant External Evaluation"
date: '`r Sys.Date()`'
author: "N.D. Olson"
format: html
editor: source
embed-resources: true
---

Update curations with JZ notes https://docs.google.com/spreadsheets/d/1evd6q8jqF51IM_uX6XrTlc-3f7dDUvurlGI-_vcV6t0/edit?usp=sharing

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Data Loading

### External Evaluation Data

The external evaluation data with JZ curations is loaded from local TSV files
(previously sourced from Google Sheets).

```{r message=FALSE}
## Loading collaborator curations
eval_csvs <- list.files(
  path = here("data/external-evaluations"),
  pattern = "Miqa.csv",
  full.names = TRUE,
  recursive = TRUE,
  include.dirs = TRUE
) %>%
  {
    set_names(., str_extract(., "(?<=evaluations/).*(?= On)"))
  }

## Subsetting small and structural variant tables based on number of rows
## - small variants expect 71, structural variants 41
## a few files have less as less than 5 variants per strata to subset
nlines <- eval_csvs %>% map_int(~ length(read_lines(.x)))

cols_to_keep <- c("callset","correct.in.callset", "correct.in.benchmark", "chrom", "chromStart", "var_type","strata")
smvar_evals <- eval_csvs[nlines > 50] %>%
  map_dfr(read_csv, name_repair = "universal", show_col_types = FALSE, .id = "callset") %>%
  select(all_of(cols_to_keep))

stvar_evals <- eval_csvs[nlines < 50] %>%
  map_dfr(read_csv,
    name_repair = "universal",
    show_col_types = FALSE,
    .id = "callset")


## Load pre-combined evaluation data with JZ curations from local TSV files
smvar_evals_jz <- read_tsv(
  here("data/external-evaluations/Q100-ext-evals-2025-04-11-smvars.tsv"),
  show_col_types = FALSE,
  name_repair = "universal"
)

stvar_evals_jz <- read_tsv(
  here("data/external-evaluations/Q100-ext-evals-2025-04-11-stvars.tsv"),
  show_col_types = FALSE,
  name_repair = "universal"
)

## Manually Revising callset names for figures
stvar_callsets <- c(
  "Baylor-Hifi" = "sniffles2\nHiFi",
  "Baylor-Ont" = "sniffles2\nONT",
  "Comenius-Svdss2" = "svdss2",
  "Illumina-Dragen" = "DRAGEN\nIll",
  "Iter-Dragen" = "DRAGEN\nITER",
  "Pacbio-Sawfish" = "sawfish",
  "Ucla-Delly-Newvcf" = "Delly",
  "Ucla-Manta-Newvcf" = "Manta",
  "Ucla-Ensamble" = "Ensamble"
)

smvar_callsets <- c(
  "Illumina-Dragen" = "DRAGEN",
  "Jhu-Imputefirst" = "Impute",
  "Pacbio-Deepvariant" = "PB-DV",
  "Roche-Neusomatic" = "roche",
  "Ultima-Ug" = "ultima"
)


## Cleaned up tables with updated curations
## Note: TSV files already contain JZ curations, no join needed
smvar_curations <- full_join(smvar_evals, smvar_evals_jz) %>%
  rename(unique_to = label) %>%
  select(
    callset,
    GRCh38_chr, GRCh38_start, GRCh38_end,
    correct.in.benchmark, JZcorrect.in.benchmark,
    correct.in.callset, JZcorrect.in.callset,
    var_type, strata, unique_to
  ) %>%
  mutate(
    curation_benchmark = if_else(!is.na(JZcorrect.in.benchmark),
      JZcorrect.in.benchmark,
      correct.in.benchmark
    ),
    curation_benchmark = factor(curation_benchmark,
      levels = c("no", "unsure", "yes")
    ),
    curation_callset = if_else(!is.na(JZcorrect.in.callset),
      JZcorrect.in.callset,
      correct.in.callset
    ),
    curation_callset = factor(curation_callset,
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  mutate(callset = smvar_callsets[callset])

stvar_curations <- full_join(stvar_evals, stvar_evals_jz) %>%
  rename(var_type = SVTYPE) %>%
  mutate(
    unique_to = if_else(unique_to == "benchmark", "bench", "query"),
    strata = str_glue("{TR_anno}\n{sv_cat}\n{unique_to}")
  ) %>%
  select(
    callset,
    GRCh38_chr, GRCh38_start, GRCh38_end,
    correct.in.benchmark, JZcorrect.in.benchmark,
    correct.in.callset, JZcorrect.in.callset,
    var_type, strata, unique_to, TR_anno, sv_cat, unique_to
  ) %>%
  mutate(
    curation_benchmark = if_else(!is.na(JZcorrect.in.benchmark),
      JZcorrect.in.benchmark,
      correct.in.benchmark
    ),
    curation_benchmark = factor(curation_benchmark,
      levels = c("no", "unsure", "yes")
    ),
    curation_callset = if_else(!is.na(JZcorrect.in.callset),
      JZcorrect.in.callset,
      correct.in.callset
    ),
    curation_callset = factor(curation_callset,
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  mutate(callset = stvar_callsets[callset]) %>% 
  filter(callset != "Manta", callset != "Delly")
```



## Strata Label Mapping

Map S-codes to readable labels for the figures.

```{r strata_labels}
smvar_strata_labels <- c(
  "S01" = "SNV: TR",
  "S02" = "SNV: Low Map",
  "S03" = "SNV: XY nonPAR",
  "S04" = "SNV: Autosome",
  "S05" = "INDEL: HP >6bp",
  "S06" = "INDEL: TR",
  "S07" = "INDEL: Low Map",
  "S08" = "INDEL: XY nonPAR",
  "S09" = "INDEL: Autosome",
  "S10" = "SNV: TR",
  "S11" = "SNV: Low Map",
  "S12" = "SNV: XY nonPAR",
  "S13" = "SNV: Autosome",
  "S14" = "INDEL: HP >6bp",
  "S15" = "INDEL: TR",
  "S16" = "INDEL: Low Map",
  "S17" = "INDEL: XY nonPAR",
  "S18" = "INDEL: Autosome"
)

## Add readable labels to smvar curations
smvar_curations <- smvar_curations %>%
  mutate(
    unique_to = if_else(strata %in% paste0("S0", 1:9), "Query", "Benchmark"),
    strata_label = smvar_strata_labels[strata]
  )
```

## Small Variants External Evaluations

```{r}
total_smvs <- smvar_evals %>% nrow()
smvs_to_review <- smvar_evals %>%
  filter(correct.in.benchmark != "yes") %>%
  nrow()
```

A total of `r total_smvs` small variants were evaluated, of which
`r smvs_to_review` were categorized as not correct or unsure in the benchmark.

```{r smvar_by_callset, fig.width = 7, fig.height = 3.5}
(smvar_eval_callset_plt <- smvar_curations %>%
  count(callset, curation_benchmark) %>%
  ggplot(aes(y = callset, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(y = "Callset", x = "# of Variants", fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r smvar_by_callset_cs, fig.width = 7, fig.height = 3.5}
(smvar_eval_callset_cs_plt <- smvar_curations %>%
  count(callset, curation_callset) %>%
  ggplot(aes(y = callset, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(y = "Callset", x = "# of Variants", fill = "Correct in\nCallset") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

```{r smvar_by_strata, fig.width = 7, fig.height = 4}
(smvar_eval_strata_plt <- smvar_curations %>%
  count(strata_label, unique_to, curation_benchmark) %>%
  mutate(strata_label = paste0(strata_label, " (", unique_to, ")")) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = "Strata", fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r smvar_by_strata_cs, fig.width = 7, fig.height = 4}
(smvar_eval_strata_cs_plt <- smvar_curations %>%
  count(strata_label, unique_to, curation_callset) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  facet_grid(
    rows = vars(unique_to),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = NULL, fill = "Correct in\nCallset") +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0)
  ))
```

### Stratification Flow Diagram (Supplemental)

```{r smvar_flow, fig.width = 8, fig.height = 8}
library(ggflowchart)

edges <- tribble(
  ~from, ~to,
  "Start", "SNV_Branch",
  "Start", "INDEL_Branch",
  "SNV_Branch", "SNV_IsTR",
  "SNV_IsTR", "S01_S10",
  "SNV_IsTR", "SNV_IsDiff",
  "SNV_IsDiff", "S02_S11",
  "SNV_IsDiff", "SNV_IsXY",
  "SNV_IsXY", "S03_S12",
  "SNV_IsXY", "S04_S13",
  "INDEL_Branch", "IND_IsHomo",
  "IND_IsHomo", "S05_S14",
  "IND_IsHomo", "IND_IsTR",
  "IND_IsTR", "S06_S15",
  "IND_IsTR", "IND_IsDiff",
  "IND_IsDiff", "S07_S16",
  "IND_IsDiff", "IND_IsXY",
  "IND_IsXY", "S08_S17",
  "IND_IsXY", "S09_S18"
)

nodes <- tribble(
  ~name, ~label, ~type,
  "Start", "Small Variants", "Root",
  "SNV_Branch", "SNVs", "Branch",
  "INDEL_Branch", "INDELs", "Branch",
  "SNV_IsTR", "in Tandem\nRepeat", "Decision",
  "SNV_IsDiff", "in Low Map\nRegion", "Decision",
  "SNV_IsXY", "XY\nnonPAR", "Decision",
  "IND_IsHomo", "in HP\n(>6bp)", "Decision",
  "IND_IsTR", "in Tandem\nRepeat", "Decision",
  "IND_IsDiff", "in Low Map\nRegion", "Decision",
  "IND_IsXY", "XY\nnonPAR", "Decision",
  "S01_S10", "S01/S10\n(TR)", "Leaf",
  "S02_S11", "S02/S11\n(Low Map)", "Leaf",
  "S03_S12", "S03/S12\n(XY)", "Leaf",
  "S04_S13", "S04/S13\n(Auto)", "Leaf",
  "S05_S14", "S05/S14\n(HP)", "Leaf",
  "S06_S15", "S06/S15\n(TR)", "Leaf",
  "S07_S16", "S07/S16\n(Low Map)", "Leaf",
  "S08_S17", "S08/S17\n(XY)", "Leaf",
  "S09_S18", "S09/S18\n(Auto)", "Leaf"
)

(smvar_flow <- ggflowchart(edges, node_data = nodes, fill = type, text_size = 2.5) +
  scale_fill_manual(values = c(
    "Root" = "#333333", "Branch" = "#777777",
    "Decision" = "#56B4E9", "Leaf" = "#E69F00"
  )) +
  labs(
    title = "Small Variant Stratification Logic",
    caption = "Left: Yes | Right/Down: No. S01-S09 unique to query; S10-S18 unique to benchmark."
  ) +
  theme(legend.title = element_blank()))

ggsave(
  filename = here("figures", "smvar_strat_flow.png"),
  plot = smvar_flow,
  width = 8, height = 8, dpi = 300
)
ggsave(
  filename = here("figures", "smvar_strat_flow.pdf"),
  plot = smvar_flow,
  width = 8, height = 8
)
```


### Structural Variants 

```{r stvar_by_strata, fig.width = 7, fig.height = 4}
## Create readable strata labels for SV
(stvar_eval_strata_plt <- stvar_curations %>%
  mutate(
    strata_label = str_replace_all(strata, "\n", " "),
    strata_label = str_replace(strata_label, "nonTR", "Non-TR"),
    strata_label = str_replace(strata_label, "TR", "TR"),
    curation_benchmark = factor(
      coalesce(as.character(curation_benchmark), correct.in.benchmark),
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  count(strata_label, curation_benchmark) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "# of Variants",
    y = "Strat",
    fill = "Correct in\nBenchmark",
  ) +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r stvar_by_strata_cs, fig.width = 7, fig.height = 4}
(stvar_eval_strata_cs_plt <- stvar_curations %>%
  mutate(
    strata_label = str_replace_all(strata, "\n", " "),
    strata_label = str_replace(strata_label, "nonTR", "Non-TR"),
    strata_label = str_replace(strata_label, "TR", "TR"),
    curation_callset = factor(
      coalesce(as.character(curation_callset), correct.in.callset),
      levels = c("no", "unsure", "yes")
    )
  ) %>%
  count(strata_label, unique_to, curation_callset) %>%
  ggplot(aes(y = strata_label, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  facet_grid(
    rows = vars(unique_to),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "# of Variants",
    y = NULL,
    fill = "Correct in\nCallset",
  ) +
  theme_manuscript() +
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0)
  ))
```

```{r stvar_by_callset, fig.width = 7, fig.height = 3.5}
(stvar_eval_callset_plt <- stvar_curations %>%
  count(callset, curation_benchmark) %>%
  ggplot(aes(y = callset, x = n, fill = curation_benchmark)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = "Callset", fill = "Correct in\nBenchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

```{r stvar_by_callset_cs, fig.width = 7, fig.height = 3.5}
(stvar_eval_callset_cs_plt <- stvar_curations %>%
  count(callset, curation_callset) %>%
  ggplot(aes(y = callset, x = n, fill = curation_callset)) +
  geom_col(position = "stack", color = "grey30", linewidth = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "# of Variants", y = "Callset", fill = "Correct in\nCallset") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

## Combined Evaluation Figure

Simple curation-count summary retained for reference. The canonical
`figures/combined_eval_strata.{png,pdf}` export is now produced in the
"Combined Strata Definition, Curation, and RIDE Figure" section below, which
needs the RIDE model estimates.

```{r combined_eval, fig.width = 6, fig.height = 3}
(combined_eval_plt <- (
  smvar_eval_strata_plt +
  stvar_eval_strata_plt) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect", ncol = 2, nrow = 1) &
  theme(legend.position = "bottom", plot.margin = margin(0, 2, 0, 0)))
```

```{r}
(combined_eval_callset_plt <- (smvar_eval_callset_plt +
  stvar_eval_callset_plt) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect", ncol = 2, nrow = 1) &
  theme(legend.position = "bottom", plot.margin = margin(0, 2, 0, 0)))

ggsave(
  filename = here("figures", "combined_eval_callset.png"),
  plot = combined_eval_callset_plt,
  width = 6,
  height = 3,
  dpi = 300
)
ggsave(
  filename = here("figures", "combined_eval_callset.pdf"),
  plot = combined_eval_callset_plt,
  width = 6, height = 3
)
```

## Binomial Model: Probability Discrepancy is Error in Comparison Callset

A binomial GLM estimates the probability that a discrepancy between the
benchmark and a comparison callset is an error in the comparison callset
(i.e., the benchmark is correct). The outcome is binarized from the curation:
`yes` = correct in benchmark vs. `no`/`unsure` = not confirmed correct.
Separate models are fit for small and structural variants. Because the goal
is to establish a lower bound on benchmark reliability, one-sided 95% lower
confidence bounds are reported.

A sensitivity analysis treats "unsure" curations as correct in the benchmark
(combining with "yes") rather than incorrect (combining with "no"), providing
an upper range on the estimated probabilities.


```{r ci_data_prep}
library(emmeans)
library(brglm2)

## Shared feature preparation
prepare_smvar_features <- function(df) {
  df |>
    mutate(
      genomic_context = str_extract(strata_label, "(?<=: ).*"),
      genomic_context = factor(genomic_context,
                               levels = c("Autosome", "TR", "HP >6bp",
                                          "Low Map", "XY nonPAR")),
      genomic_context = droplevels(genomic_context),
      var_type = factor(var_type, levels = c("SNP", "INDEL")),
      unique_to = factor(unique_to, levels = c("Query", "Benchmark"))
    )
}

prepare_stvar_features <- function(df) {
  df |>
    mutate(
      TR_anno = factor(TR_anno, levels = c("nonTR", "TR")),
      sv_cat = factor(sv_cat, levels = c("DEL", "INS")),
      unique_to = factor(unique_to, levels = c("query", "bench"))
    )
}

## Primary analysis: unsure = incorrect (conservative)
smvar_model_df <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark == "yes")) |>
  prepare_smvar_features()

stvar_model_df <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark == "yes")) |>
  prepare_stvar_features()

## Sensitivity analysis: unsure = correct
smvar_sens_df <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark %in% c("yes", "unsure"))) |>
  prepare_smvar_features()

stvar_sens_df <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_benchmark %in% c("yes", "unsure"))) |>
  prepare_stvar_features()
```

```{r ci_models}
## Primary models (standard GLM)
smvar_glm <- glm(correct_binary ~ var_type + genomic_context + unique_to,
                 family = binomial, data = smvar_model_df)
stvar_glm <- glm(correct_binary ~ TR_anno + sv_cat + unique_to,
                 family = binomial, data = stvar_model_df)

## Sensitivity models
## smvar sensitivity has quasi-complete separation (only 5 "no" curations
## with 0 in the Low Map context), so bias-reduced estimation is used.
smvar_sens_glm <- glm(correct_binary ~ var_type + genomic_context + unique_to,
                      family = binomial, data = smvar_sens_df,
                      method = "brglmFit")
stvar_sens_glm <- glm(correct_binary ~ TR_anno + sv_cat + unique_to,
                      family = binomial, data = stvar_sens_df)
```

```{r ci_emmeans}
## Helper to extract emmeans with one-sided 95% lower bound
extract_emmeans <- function(em, benchmarkset, feature_name) {
  as_tibble(summary(em, side = ">")) |>
    mutate(
      benchmarkset = benchmarkset,
      feature = feature_name,
      level = if ("1" %in% names(pick(everything()))) "Overall"
              else as.character(pick(1)[[1]])
    ) |>
    select(benchmarkset, feature, level, prob, SE, asymp.LCL)
}

## Compute emmeans for all four models
compute_all_emmeans <- function(smvar_fit, stvar_fit) {
  bind_rows(
    extract_emmeans(
      emmeans(smvar_fit, ~ 1, type = "response", weights = "proportional"),
      "Small Variant", "Overall"),
    extract_emmeans(
      emmeans(smvar_fit, ~ var_type, type = "response", weights = "proportional"),
      "Small Variant", "Variant Type"),
    extract_emmeans(
      emmeans(smvar_fit, ~ genomic_context, type = "response", weights = "proportional"),
      "Small Variant", "Genomic Context"),
    extract_emmeans(
      emmeans(smvar_fit, ~ unique_to, type = "response", weights = "proportional"),
      "Small Variant", "Unique To"),
    extract_emmeans(
      emmeans(stvar_fit, ~ 1, type = "response", weights = "proportional"),
      "Structural Variant", "Overall"),
    extract_emmeans(
      emmeans(stvar_fit, ~ TR_anno, type = "response", weights = "proportional"),
      "Structural Variant", "TR Annotation"),
    extract_emmeans(
      emmeans(stvar_fit, ~ sv_cat, type = "response", weights = "proportional"),
      "Structural Variant", "SV Category"),
    extract_emmeans(
      emmeans(stvar_fit, ~ unique_to, type = "response", weights = "proportional"),
      "Structural Variant", "Unique To")
  )
}

primary_results <- compute_all_emmeans(smvar_glm, stvar_glm) |>
  mutate(analysis = "Conservative (unsure = incorrect)")
sens_results <- compute_all_emmeans(smvar_sens_glm, stvar_sens_glm) |>
  mutate(analysis = "Sensitive (unsure = correct)")

ci_results <- bind_rows(primary_results, sens_results)
```

```{r ci_forest_plot, fig.width = 7, fig.height = 5.5}
#| fig-cap: >
#|   Estimated probability that a discrepancy between the Q100 v5.0q HG002
#|   benchmark and a comparison callset represents an error in the comparison
#|   callset, stratified by variant class and genomic context. Each row shows
#|   one stratum within a feature category (facet strip); points are marginal
#|   probability estimates from bias-reduced logistic regression models via
#|   emmeans with proportional weighting. Red circles indicate the conservative
#|   assumption (unsure-classification variants treated as callset errors);
#|   blue triangles indicate the sensitive assumption (unsure variants treated
#|   as callset correct). The grey segment spanning each row shows the range of
#|   estimates across the two assumptions, reflecting uncertainty introduced by
#|   the classification of unsure variants. Columns separate small variants
#|   (left) and structural variants (right), which are modeled independently.
#|   The Overall estimate is not conditioned on any stratum.
## Prepare display labels; Overall placed first so its facet appears at top
ci_plot_df <- ci_results |>
  mutate(
    label = if_else(feature == "Overall", "Overall", level),
    label = case_when(
      label == "nonTR" ~ "Non-TR",
      label == "bench" ~ "Benchmark",
      label == "query" ~ "Query",
      .default = label
    ),
    feature = factor(feature,
      levels = c("Overall", "Unique To", "Variant Type", "Genomic Context",
                 "TR Annotation", "SV Category")
    ),
    benchmarkset = factor(benchmarkset,
      levels = c("Small Variant", "Structural Variant")
    )
  )

(ci_forest_plt <- ggplot(ci_plot_df,
       aes(x = prob, y = reorder(label, prob))) +
  ## Grey segment connecting conservative and sensitive endpoints
  geom_line(
    aes(group = interaction(label, feature, benchmarkset)),
    color = "grey60",
    linewidth = 0.8
  ) +
  ## Endpoints colored and shaped by analysis assumption
  geom_point(aes(color = analysis, shape = analysis), size = 2.5) +
  facet_grid(
    rows = vars(feature),
    cols = vars(benchmarkset),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0.6, 1)) +
  scale_color_manual(
    values = c(
      "Conservative (unsure = incorrect)" = "#E41A1C",
      "Sensitive (unsure = correct)" = "#377EB8"
    )
  ) +
  scale_shape_manual(
    values = c(
      "Conservative (unsure = incorrect)" = 16,
      "Sensitive (unsure = correct)" = 17
    )
  ) +
  labs(
    x = "Probability Discrepancy is Error in Comparison Callset",
    y = NULL,
    color = "Unsure Classification",
    shape = "Unsure Classification"
  ) +
  theme_manuscript() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "bottom"
  ))
```


```{r}
ggsave(
  filename = here("figures", "ride_estimates.png"),
  plot = ci_forest_plt,
  width = 8, height = 8, dpi = 300
)
ggsave(
  filename = here("figures", "ride_estimates.pdf"),
  plot = ci_forest_plt,
  width = 8, height = 8
)
```

## Combined Strata Definition, Curation, and RIDE Figure

One figure per benchmark set block combining (i) an UpSet-style matrix that
defines each curation stratum by its factor levels, (ii) the curation counts
per stratum, and (iii) the marginal RIDE probability estimates placed directly
above the factor-level column each estimate averages over.

Small-variant genomic contexts were assigned sequentially (INDEL: HP >6bp,
then TR, then Low Map, then remaining autosomal; SNV: TR, then Low Map, then
remaining autosomal), so a stratum's context implies the variant is *not* in
any higher-precedence context (open circles). XY non-PAR strata (S03, S08,
S12, S17) were defined but not curated and therefore do not appear.

```{r strata_definition_tables}
smvar_context_order <- list(
  SNV = c("TR", "Low Map", "Autosome"),
  INDEL = c("HP >6bp", "TR", "Low Map", "Autosome")
)

## Status of one context column for a stratum: the assigned context is
## "in", higher-precedence contexts are "not in", others are not evaluated.
context_status <- function(var_type, assigned, column) {
  ord <- smvar_context_order[[var_type]]
  if (!column %in% ord) {
    return(NA_character_)
  }
  d <- match(column, ord) - match(assigned, ord)
  if (d == 0) "in" else if (d < 0) "not in" else NA_character_
}

smvar_def_cols <- tribble(
  ~column,     ~group,
  "Overall",   "Overall",
  "Query",     "Unique to",
  "Benchmark", "Unique to",
  "SNV",       "Variant type",
  "INDEL",     "Variant type",
  "HP >6bp",   "Genomic context",
  "TR",        "Genomic context",
  "Low Map",   "Genomic context",
  "Autosome",  "Genomic context"
)

smvar_strata_def <- smvar_curations %>%
  distinct(strata, unique_to, var_type, strata_label) %>%
  mutate(
    var_type = if_else(var_type == "SNP", "SNV", var_type),
    context = str_extract(strata_label, "(?<=: ).*"),
    row_id = strata
  ) %>%
  arrange(strata)

smvar_def_long <- smvar_strata_def %>%
  crossing(smvar_def_cols) %>%
  mutate(status = case_when(
    group == "Overall" ~ "in",
    group == "Unique to" ~ if_else(column == unique_to, "in", NA_character_),
    group == "Variant type" ~ if_else(column == var_type, "in", NA_character_),
    group == "Genomic context" ~
      pmap_chr(list(var_type, context, column), context_status)
  ))

stvar_def_cols <- tribble(
  ~column,     ~group,
  "Overall",   "Overall",
  "Query",     "Unique to",
  "Benchmark", "Unique to",
  "DEL",       "SV type",
  "INS",       "SV type",
  "TR",        "TR annotation",
  "Non-TR",    "TR annotation"
)

stvar_strata_def <- stvar_curations %>%
  distinct(strata, unique_to, sv_cat, TR_anno) %>%
  mutate(
    unique_to = if_else(unique_to == "bench", "Benchmark", "Query"),
    TR_anno = if_else(TR_anno == "nonTR", "Non-TR", TR_anno),
    row_id = paste(unique_to, sv_cat, TR_anno)
  ) %>%
  # Query block first, then DEL/INS, then TR/Non-TR (matches column order)
  arrange(desc(unique_to == "Query"), sv_cat, desc(TR_anno == "TR"))

stvar_def_long <- stvar_strata_def %>%
  crossing(stvar_def_cols) %>%
  mutate(status = case_when(
    group == "Overall" ~ "in",
    group == "Unique to" ~ if_else(column == unique_to, "in", NA_character_),
    group == "SV type" ~ if_else(column == sv_cat, "in", NA_character_),
    group == "TR annotation" ~ if_else(column == TR_anno, "in", NA_character_)
  ))

## Curation counts per stratum
smvar_bars_df <- smvar_curations %>%
  count(row_id = strata, curation_benchmark)

stvar_bars_df <- stvar_curations %>%
  mutate(row_id = paste(
    if_else(unique_to == "bench", "Benchmark", "Query"),
    sv_cat,
    if_else(TR_anno == "nonTR", "Non-TR", TR_anno)
  )) %>%
  count(row_id, curation_benchmark)

## Marginal RIDE estimates keyed to the matrix columns
ride_by_column <- function(df, benchmarkset, def_cols) {
  df %>%
    filter(benchmarkset == !!benchmarkset) %>%
    mutate(column = case_when(
      feature == "Overall" ~ "Overall",
      label == "SNP" ~ "SNV",
      .default = label
    )) %>%
    inner_join(def_cols, by = "column")
}

smvar_ride_df <- ride_by_column(ci_plot_df, "Small Variant", smvar_def_cols)
stvar_ride_df <- ride_by_column(ci_plot_df, "Structural Variant", stvar_def_cols)
```

```{r strata_block_builder}
unsure_cols <- c(
  "Conservative (unsure = incorrect)" = "#E41A1C",
  "Sensitive (unsure = correct)" = "#377EB8"
)
unsure_shapes <- c(
  "Conservative (unsure = incorrect)" = 16,
  "Sensitive (unsure = correct)" = 17
)
unsure_labels <- c(
  "Conservative (unsure = incorrect)" = "Conservative (unsure = no)",
  "Sensitive (unsure = correct)" = "Sensitive (unsure = yes)"
)

## Column labels wrapped so they fit horizontally above the matrix columns
strata_col_labels <- c(
  "Benchmark" = "Bench", "HP >6bp" = "HP\n>6bp",
  "Low Map" = "Low\nMap", "Autosome" = "Auto-\nsome"
)

## Build the three row/column-aligned panels for one benchmark set:
## RIDE header (above matrix columns), definition matrix, curation bars.
build_strata_block <- function(def_long, ride_df, bars_df, def_cols,
                               group_labels, row_levels, title,
                               show_row_labels = TRUE) {
  col_levels <- def_cols$column
  group_levels <- unique(def_cols$group)
  fct_cols <- function(df) {
    df %>%
      mutate(
        column = factor(column, col_levels),
        group = factor(group, group_levels)
      )
  }
  def_long <- fct_cols(def_long) %>%
    mutate(
      row = factor(row_id, rev(row_levels)),
      status = factor(status, levels = c("in", "not in"))
    )
  ride_df <- fct_cols(ride_df)
  bars_df <- bars_df %>% mutate(row = factor(row_id, rev(row_levels)))
  n_max <- max(tapply(bars_df$n, bars_df$row_id, sum))
  stripe_rows <- tibble(
    row = factor(row_levels[seq_along(row_levels) %% 2 == 0], rev(row_levels))
  )

  p_ride <- ggplot(ride_df, aes(x = column, y = prob)) +
    geom_line(aes(group = column), color = "grey60", linewidth = 0.8) +
    geom_point(aes(color = analysis, shape = analysis), size = 2) +
    facet_grid(
      cols = vars(group), scales = "free_x", space = "free_x",
      labeller = as_labeller(group_labels)
    ) +
    scale_y_continuous(
      labels = scales::percent, limits = c(0.7, 1), breaks = seq(0.7, 1, 0.1)
    ) +
    scale_color_manual(values = unsure_cols, labels = unsure_labels) +
    scale_shape_manual(values = unsure_shapes, labels = unsure_labels) +
    guides(
      color = guide_legend(order = 3, direction = "vertical", title.position = "top"),
      shape = guide_legend(order = 3, direction = "vertical", title.position = "top")
    ) +
    labs(
      # title = title, 
      x = NULL, 
      y = "P(query error)",
      color = "Unsure classification", shape = "Unsure classification"
    ) +
    theme_manuscript() +
    theme(
      plot.title = element_text(size = 9.5, margin = margin(l = 14, b = 2)),
      plot.title.position = "plot",
      plot.margin = margin(2, 2, 0, 2),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
      strip.text = element_text(size = 8, face = "bold", lineheight = 0.9),
      strip.background = element_blank(),
      panel.spacing.x = unit(4, "pt")
    )

  p_matrix <- ggplot(def_long, aes(x = column, y = row)) +
    geom_tile(data = ~ semi_join(.x, stripe_rows, by = "row"), fill = "grey92") +
    geom_point(data = ~ filter(.x, !is.na(status)), aes(shape = status), size = 1.9) +
    facet_grid(cols = vars(group), scales = "free_x", space = "free_x") +
    scale_x_discrete(position = "top", labels = strata_col_labels) +
    scale_y_discrete(limits = rev(row_levels)) +
    scale_shape_manual(
      values = c("in" = 19, "not in" = 1),
      labels = c("in" = "Defining", "not in" = "Excluded"),
      name = "Stratum definition", drop = FALSE
    ) +
    guides(shape = guide_legend(order = 1, direction = "horizontal", title.position = "top")) +
    labs(x = NULL, y = NULL) +
    theme_manuscript() +
    theme(
      plot.margin = margin(0, 2, 2, 2),
      strip.text = element_blank(),
      strip.background = element_blank(),
      axis.text.x.top = element_text(
        size = 7.5, lineheight = 0.85, vjust = 0, margin = margin(b = 2)
      ),
      axis.text.y = if (show_row_labels) element_text(size = 7.5) else element_blank(),
      axis.ticks = element_blank(),
      panel.border = element_blank(),
      panel.spacing.x = unit(4, "pt")
    )

  p_bars <- ggplot(bars_df, aes(x = n, y = row, fill = curation_benchmark)) +
    geom_tile(
      data = stripe_rows, aes(x = n_max / 2, y = row),
      width = n_max, height = 1, fill = "grey92", inherit.aes = FALSE
    ) +
    geom_col(position = "stack", color = "grey30", linewidth = 0.2, width = 0.7) +
    scale_fill_brewer(palette = "Set2", drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
    scale_y_discrete(limits = rev(row_levels)) +
    guides(fill = guide_legend(order = 2, direction = "horizontal", title.position = "top")) +
    labs(x = "# of curated variants", y = NULL, fill = "Correct in benchmark") +
    theme_manuscript() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

  list(ride = p_ride, matrix = p_matrix, bars = p_bars)
}
```

```{r combined_eval_strata, fig.width = 7.09, fig.height = 7.7}
#| fig-cap: >
#|   External evaluation of the v5.0q (A) small variant and (B) structural
#|   variant benchmark sets by manual curation of discrepancies with external
#|   callsets. Matrices define the strata used to sample discrepancies for
#|   curation: each row is a stratum (small variant S-codes correspond to the
#|   Methods strata table) and filled circles mark its defining features.
#|   Small variant genomic contexts were assigned sequentially (INDELs:
#|   homopolymer >6 bp, then tandem repeat, then difficult-to-map, then
#|   remaining autosomal; SNVs: tandem repeat, then difficult-to-map, then
#|   remaining autosomal), so open circles mark higher-precedence contexts the
#|   variant does not overlap; XY non-PAR strata were defined but not curated.
#|   Bars give the number of curated variants per stratum classified as
#|   correct (yes), incorrect (no), or unsure in the benchmark. Points above
#|   each factor-level column are the estimated probabilities that a
#|   discrepancy is an error in the comparison callset, from binomial GLMs
#|   (marginal estimates via emmeans with proportional weighting): red circles
#|   treat unsure curations as incorrect (conservative), blue triangles as
#|   correct (sensitive), and the grey segment spans the two. Low Map denotes
#|   the composite difficult-to-map annotation (low mappability 100/250 bp,
#|   segmental duplications, RepeatMasker LINE/LTR/satellite).
smvar_block <- build_strata_block(
  smvar_def_long, smvar_ride_df, smvar_bars_df, smvar_def_cols,
  group_labels = c(
    "Overall" = "", "Unique to" = "Unique\nto",
    "Variant type" = "Variant\ntype", "Genomic context" = "Genomic\ncontext"
  ),
  row_levels = smvar_strata_def$row_id,
  title = "Small variants",
  show_row_labels = FALSE
)

stvar_block <- build_strata_block(
  stvar_def_long, stvar_ride_df, stvar_bars_df, stvar_def_cols,
  group_labels = c(
    "Overall" = "", "Unique to" = "Unique\nto",
    "SV type" = "SV\ntype", "TR annotation" = "TR\nannotation"
  ),
  row_levels = stvar_strata_def$row_id,
  title = "Structural variants",
  show_row_labels = FALSE
)
## SV strata have no excluded contexts; the small variant matrix carries the legend
stvar_block$matrix <- stvar_block$matrix + guides(shape = "none")

## Each matrix column is one design column so dot spacing matches across
## blocks; bars share the final column. Legends fill the empty SV header area.
strata_design <- "
AAAAAAAAA#
BBBBBBBBBC
DDDDDDDDD#
EEEEEEEEEF
GGGGGGGGGG
"

(combined_eval_strata_plt <- smvar_block$ride + smvar_block$matrix + smvar_block$bars +
  stvar_block$ride + stvar_block$matrix + stvar_block$bars + guide_area() + 
  plot_layout(
    design = strata_design, widths = c(rep(1, 9), 5),
    heights = c(0.95, 3.0, 0.95, 1.75), guides = "collect"
  ) +
  plot_annotation(tag_levels = list(c("A", "", "", "B", "", ""))) &
  theme(
    legend.position = "right", legend.box = "horizontal",
    legend.justification = c(0, 0.5), legend.box.just = "left",
    legend.key.height = unit(0.3, "cm"), legend.key.width = unit(0.35, "cm"),
    legend.margin = margin(0, 0, 4, 0), legend.spacing.y = unit(1, "pt"),
  ))

ggsave(
  filename = here("figures", "combined_eval_strata.png"),
  plot = combined_eval_strata_plt,
  width = 7.09, height = 7.7, dpi = 450, bg = "white"
)
ggsave(
  filename = here("figures", "combined_eval_strata.pdf"),
  plot = combined_eval_strata_plt,
  width = 7.09, height = 7.7, bg = "white"
)
```

```{r ci_summary_table}
## Per-level sample sizes
smvar_n <- smvar_model_df |>
  summarise(
    Overall = n(), SNP = sum(var_type == "SNP"),
    INDEL = sum(var_type == "INDEL"),
    Autosome = sum(genomic_context == "Autosome"),
    TR = sum(genomic_context == "TR"),
    `HP >6bp` = sum(genomic_context == "HP >6bp"),
    `Low Map` = sum(genomic_context == "Low Map"),
    Query = sum(unique_to == "Query"),
    Benchmark = sum(unique_to == "Benchmark")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Small Variant")

stvar_n <- stvar_model_df |>
  summarise(
    Overall = n(), `Non-TR` = sum(TR_anno == "nonTR"),
    TR = sum(TR_anno == "TR"), DEL = sum(sv_cat == "DEL"),
    INS = sum(sv_cat == "INS"), Query = sum(unique_to == "query"),
    Benchmark = sum(unique_to == "bench")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Structural Variant")

n_lookup <- bind_rows(smvar_n, stvar_n)

## Build table with both analyses side by side
ci_table_wide <- ci_results |>
  mutate(
    level = case_when(
      feature == "Overall" ~ "",
      level == "nonTR" ~ "Non-TR",
      level == "bench" ~ "Benchmark",
      level == "query" ~ "Query",
      .default = level
    )
  ) |>
  select(benchmarkset, feature, level, analysis, prob, asymp.LCL) |>
  pivot_wider(
    names_from = analysis,
    values_from = c(prob, asymp.LCL),
    names_glue = "{analysis}|{.value}"
  ) |>
  left_join(n_lookup, by = c("benchmarkset", "level"))

ci_table_wide |>
  select(
    benchmarkset, level, feature, n,
    `Conservative (unsure = incorrect)|prob`,
    `Conservative (unsure = incorrect)|asymp.LCL`,
    `Sensitive (unsure = correct)|prob`,
    `Sensitive (unsure = correct)|asymp.LCL`
  ) |>
  mutate(across(
    c(
      `Conservative (unsure = incorrect)|prob`,
      `Conservative (unsure = incorrect)|asymp.LCL`,
      `Sensitive (unsure = correct)|prob`,
      `Sensitive (unsure = correct)|asymp.LCL`
    ),
    ~ sprintf("%.1f%%", .x * 100)
  )) |>
  as_grouped_flextable(groupname_col = "benchmarkset") |>
  flextable::add_header_row(
    top = TRUE,
    values = c(
      "", "", "",
      "Conservative (unsure = incorrect)",
      "Sensitive (unsure = correct)"
    ),
    colwidths = c(1, 1, 1, 2, 2)
  ) |>
  cols_label_flextable(
    level = "Level", feature = "Feature", n = "N",
    `Conservative (unsure = incorrect)|prob` = "Est. (%)",
    `Conservative (unsure = incorrect)|asymp.LCL` = "LCB (%)",
    `Sensitive (unsure = correct)|prob` = "Est. (%)",
    `Sensitive (unsure = correct)|asymp.LCL` = "LCB (%)"
  ) |>
  flextable::align(j = c("level", "feature"), align = "left", part = "body") |>
  flextable::align(j = c(3, 4, 5, 6, 7), align = "center", part = "body") |>
  flextable::bold(i = ~ feature == "Overall") |>
  flextable::footnote(
    i = 1, j = c(4, 6), part = "header",
    value = flextable::as_paragraph(
      "Marginal probabilities via emmeans with proportional weighting, averaging over other model terms."
    ),
    ref_symbols = "a"
  ) |>
  flextable::footnote(
    i = ~ level == "HP >6bp", j = "feature", part = "body",
    value = flextable::as_paragraph("HP >6bp stratum applies only to INDELs."),
    ref_symbols = "b"
  ) |>
  flextable::footnote(
    i = 1, j = 6, part = "header",
    value = flextable::as_paragraph(
      "Small variant sensitivity model fit with bias-reduced logistic regression (brglm2) due to quasi-complete separation."
    ),
    ref_symbols = "c"
  ) |>
  theme_flextable_manuscript() |>
  flextable::set_caption(
    "Estimated Probability That a Discrepancy is an Error in the Comparison Callset. Marginal estimates from binomial GLMs with one-sided 95% lower confidence bounds (LCB)."
  )
```

## Binomial Model: Benchmark Identifies Errors in Comparison Callset (Callset-Based)

This analysis uses the `correct.in.callset` curation to estimate the
probability that a discrepancy reflects an error in the comparison callset.
When the comparison callset is judged "not correct" (`no`), the benchmark has
identified a real error — even if the benchmark's own call at that site is
also imperfect. This captures the broader utility of the benchmark for
flagging comparison callset errors.

The outcome is inverted relative to the curation values: `no` = the
comparison is wrong (success), `yes` = the comparison is correct (failure).
The conservative analysis treats `unsure` as the comparison being correct
(fewer successes); the sensitivity analysis treats `unsure` as the comparison
being wrong (more successes).

```{r cs_data_prep}
## Primary analysis: unsure = callset correct (conservative)
smvar_model_df_cs <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_callset == "no")) |>
  prepare_smvar_features()

stvar_model_df_cs <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_callset == "no")) |>
  prepare_stvar_features()

## Sensitivity analysis: unsure = callset incorrect
smvar_sens_df_cs <- smvar_curations |>
  mutate(correct_binary = as.integer(curation_callset %in% c("no", "unsure"))) |>
  prepare_smvar_features()

stvar_sens_df_cs <- stvar_curations |>
  mutate(correct_binary = as.integer(curation_callset %in% c("no", "unsure"))) |>
  prepare_stvar_features()
```

```{r cs_models}
## Primary models
smvar_glm_cs <- glm(correct_binary ~ var_type + genomic_context + unique_to,
                    family = binomial, data = smvar_model_df_cs)
stvar_glm_cs <- glm(correct_binary ~ TR_anno + sv_cat + unique_to,
                    family = binomial, data = stvar_model_df_cs)

## Sensitivity models — use brglmFit if quasi-complete separation is detected
smvar_sens_glm_cs <- tryCatch(
  glm(correct_binary ~ var_type + genomic_context + unique_to,
      family = binomial, data = smvar_sens_df_cs),
  warning = function(w) {
    glm(correct_binary ~ var_type + genomic_context + unique_to,
        family = binomial, data = smvar_sens_df_cs,
        method = "brglmFit")
  }
)
stvar_sens_glm_cs <- tryCatch(
  glm(correct_binary ~ TR_anno + sv_cat + unique_to,
      family = binomial, data = stvar_sens_df_cs),
  warning = function(w) {
    glm(correct_binary ~ TR_anno + sv_cat + unique_to,
        family = binomial, data = stvar_sens_df_cs,
        method = "brglmFit")
  }
)
```

```{r cs_emmeans}
primary_results_cs <- compute_all_emmeans(smvar_glm_cs, stvar_glm_cs) |>
  mutate(analysis = "Conservative (unsure = callset correct)")
sens_results_cs <- compute_all_emmeans(smvar_sens_glm_cs, stvar_sens_glm_cs) |>
  mutate(analysis = "Sensitive (unsure = callset incorrect)")

ci_results_cs <- bind_rows(primary_results_cs, sens_results_cs)
```

```{r cs_forest_plot, fig.width = 7, fig.height = 5.5}
#| fig-cap: >
#|   Estimated probability that the Q100 v5.0q HG002 benchmark correctly
#|   identifies an error in the comparison callset, stratified by variant class
#|   and genomic context. Each row shows one stratum within a feature category
#|   (facet strip); points are marginal probability estimates from bias-reduced
#|   logistic regression models via emmeans with proportional weighting. Red
#|   circles indicate the conservative assumption (unsure-classification variants
#|   treated as callset correct, reducing apparent error rate); blue triangles
#|   indicate the sensitive assumption (unsure variants treated as callset
#|   errors, increasing apparent error rate). The grey segment spanning each row
#|   shows the range of estimates across the two assumptions. Columns separate
#|   small variants (left) and structural variants (right), which are modeled
#|   independently. The Overall estimate is not conditioned on any stratum.
## Prepare display labels; Overall placed first so its facet appears at top
ci_plot_df_cs <- ci_results_cs |>
  mutate(
    label = if_else(feature == "Overall", "Overall", level),
    label = case_when(
      label == "nonTR" ~ "Non-TR",
      label == "bench" ~ "Benchmark",
      label == "query" ~ "Query",
      .default = label
    ),
    feature = factor(feature,
      levels = c("Overall", "Unique To", "Variant Type", "Genomic Context",
                 "TR Annotation", "SV Category")
    ),
    benchmarkset = factor(benchmarkset,
      levels = c("Small Variant", "Structural Variant")
    )
  )

(ci_forest_cs_plt <- ggplot(ci_plot_df_cs,
       aes(x = prob, y = reorder(label, prob))) +
  ## Grey segment connecting conservative and sensitive endpoints
  geom_line(
    aes(group = interaction(label, feature, benchmarkset)),
    color = "grey60",
    linewidth = 0.8
  ) +
  ## Endpoints colored and shaped by analysis assumption
  geom_point(aes(color = analysis, shape = analysis), size = 2.5) +
  facet_grid(
    rows = vars(feature),
    cols = vars(benchmarkset),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0.6, 1)) +
  scale_color_manual(
    values = c(
      "Conservative (unsure = callset correct)" = "#E41A1C",
      "Sensitive (unsure = callset incorrect)" = "#377EB8"
    )
  ) +
  scale_shape_manual(
    values = c(
      "Conservative (unsure = callset correct)" = 16,
      "Sensitive (unsure = callset incorrect)" = 17
    )
  ) +
  labs(
    x = "Probability Benchmark Identifies Error in Comparison Callset",
    y = NULL,
    color = "Unsure Classification",
    shape = "Unsure Classification"
  ) +
  theme_manuscript() +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.position = "bottom"
  ))
```

```{r}
ggsave(
  filename = here("figures", "ride_estimates_callset.png"),
  plot = ci_forest_cs_plt,
  width = 8, height = 8, dpi = 300
)
ggsave(
  filename = here("figures", "ride_estimates_callset.pdf"),
  plot = ci_forest_cs_plt,
  width = 8, height = 8
)
```

```{r cs_summary_table}
## Per-level sample sizes (same as benchmark analysis)
smvar_n_cs <- smvar_model_df_cs |>
  summarise(
    Overall = n(), SNP = sum(var_type == "SNP"),
    INDEL = sum(var_type == "INDEL"),
    Autosome = sum(genomic_context == "Autosome"),
    TR = sum(genomic_context == "TR"),
    `HP >6bp` = sum(genomic_context == "HP >6bp"),
    `Low Map` = sum(genomic_context == "Low Map"),
    Query = sum(unique_to == "Query"),
    Benchmark = sum(unique_to == "Benchmark")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Small Variant")

stvar_n_cs <- stvar_model_df_cs |>
  summarise(
    Overall = n(), `Non-TR` = sum(TR_anno == "nonTR"),
    TR = sum(TR_anno == "TR"), DEL = sum(sv_cat == "DEL"),
    INS = sum(sv_cat == "INS"), Query = sum(unique_to == "query"),
    Benchmark = sum(unique_to == "bench")
  ) |>
  pivot_longer(everything(), names_to = "level", values_to = "n") |>
  mutate(level = if_else(level == "Overall", "", level),
         benchmarkset = "Structural Variant")

n_lookup_cs <- bind_rows(smvar_n_cs, stvar_n_cs)

ci_table_wide_cs <- ci_results_cs |>
  mutate(
    level = case_when(
      feature == "Overall" ~ "",
      level == "nonTR" ~ "Non-TR",
      level == "bench" ~ "Benchmark",
      level == "query" ~ "Query",
      .default = level
    )
  ) |>
  select(benchmarkset, feature, level, analysis, prob, asymp.LCL) |>
  pivot_wider(
    names_from = analysis,
    values_from = c(prob, asymp.LCL),
    names_glue = "{analysis}|{.value}"
  ) |>
  left_join(n_lookup_cs, by = c("benchmarkset", "level"))

ci_table_wide_cs |>
  select(
    benchmarkset, level, feature, n,
    `Conservative (unsure = callset correct)|prob`,
    `Conservative (unsure = callset correct)|asymp.LCL`,
    `Sensitive (unsure = callset incorrect)|prob`,
    `Sensitive (unsure = callset incorrect)|asymp.LCL`
  ) |>
  mutate(across(
    c(
      `Conservative (unsure = callset correct)|prob`,
      `Conservative (unsure = callset correct)|asymp.LCL`,
      `Sensitive (unsure = callset incorrect)|prob`,
      `Sensitive (unsure = callset incorrect)|asymp.LCL`
    ),
    ~ sprintf("%.1f%%", .x * 100)
  )) |>
  as_grouped_flextable(groupname_col = "benchmarkset") |>
  flextable::add_header_row(
    top = TRUE,
    values = c(
      "", "", "",
      "Conservative (unsure = callset correct)",
      "Sensitive (unsure = callset incorrect)"
    ),
    colwidths = c(1, 1, 1, 2, 2)
  ) |>
  cols_label_flextable(
    level = "Level", feature = "Feature", n = "N",
    `Conservative (unsure = callset correct)|prob` = "Est. (%)",
    `Conservative (unsure = callset correct)|asymp.LCL` = "LCB (%)",
    `Sensitive (unsure = callset incorrect)|prob` = "Est. (%)",
    `Sensitive (unsure = callset incorrect)|asymp.LCL` = "LCB (%)"
  ) |>
  flextable::align(j = c("level", "feature"), align = "left", part = "body") |>
  flextable::align(j = c(3, 4, 5, 6, 7), align = "center", part = "body") |>
  flextable::bold(i = ~ feature == "Overall") |>
  flextable::footnote(
    i = 1, j = c(4, 6), part = "header",
    value = flextable::as_paragraph(
      "Marginal probabilities via emmeans with proportional weighting, averaging over other model terms."
    ),
    ref_symbols = "a"
  ) |>
  flextable::footnote(
    i = ~ level == "HP >6bp", j = "feature", part = "body",
    value = flextable::as_paragraph("HP >6bp stratum applies only to INDELs."),
    ref_symbols = "b"
  ) |>
  flextable::footnote(
    i = 1, j = 6, part = "header",
    value = flextable::as_paragraph(
      "Bias-reduced logistic regression (brglm2) used where quasi-complete separation was detected."
    ),
    ref_symbols = "c"
  ) |>
  theme_flextable_manuscript() |>
  flextable::set_caption(
    "Estimated Probability That Benchmark Identifies an Error in the Comparison Callset. Based on correct-in-callset curations; marginal estimates from binomial GLMs with one-sided 95% LCB."
  )
```

### Methods: External Evaluation Confidence Intervals

To estimate the probability that a discrepancy between the benchmark set and
a comparison callset represents an error in the comparison callset (i.e., the
benchmark is correct), we fit separate binomial generalized linear models
(GLMs) with a logit link for small variants and structural variants.

For each comparison callset, five discrepancies per stratum were randomly
selected for manual curation. Each discrepancy was classified as "correct in
benchmark" (yes), "not correct in benchmark" (no), or "unsure." A subset of
curations were independently re-reviewed by a second expert (JZ); where
available, the second reviewer's assessment was used as the final curation.
The primary analysis binarized the outcome conservatively: "yes" was treated
as a success (the discrepancy is an error in the comparison callset), while
"no" and "unsure" were combined as failures. A sensitivity analysis treated
"unsure" as successes (combining with "yes"), providing an upper range for
the estimated probabilities.

For small variants (N = `r nrow(smvar_model_df)`;
`r length(unique(smvar_model_df$callset))` callsets), the model included
three additive predictors: variant type (SNP vs. INDEL), genomic context
(Autosome, tandem repeat, homopolymer >6 bp, or low mappability region),
and whether the discrepancy was unique to the query callset or the benchmark.
For structural variants (N = `r nrow(stvar_model_df)`;
`r length(unique(stvar_model_df$callset))` callsets), predictors were tandem
repeat annotation (TR vs. non-TR), SV category (DEL vs. INS), and unique-to
status (query vs. benchmark). All predictors were entered as main effects
without interactions. Because different sets of variants were curated for
each callset, each observation represents an independent evaluation at a
distinct genomic position, and callset was not included as a model term.

Because the goal is to establish a lower bound on how reliably the benchmark
identifies errors in comparison callsets, one-sided 95% lower confidence
bounds were reported rather than two-sided intervals. Marginal probabilities
and confidence bounds on the probability scale were obtained using the
`emmeans` R package with proportional weighting to average over the other
model terms. Proportional weighting ensures that factor levels are weighted
by their observed frequencies in the data, which is relevant because the
homopolymer >6 bp context applies only to INDELs. In the sensitivity
analysis, the small variant model was fit using bias-reduced logistic
regression (Firth penalization via the `brglm2` R package) because treating
"unsure" as correct left only 5 failures across 350 observations, causing
quasi-complete separation in the low mappability stratum.

**Assumptions.** The binomial GLM assumes that curations are independent
Bernoulli trials, that the log-odds of the outcome are a linear additive
function of the predictors, and that the effect of each predictor is
constant across levels of the other predictors (no interactions).
Independence is reasonable given that each observation is a unique genomic
position evaluated by a single curator. The additive assumption implies,
for example, that the difference in log-odds between TR and autosomal
contexts is the same for SNPs and INDELs. Given the sample sizes, there was
insufficient power to test interactions. The one-sided Wald confidence bounds
rely on asymptotic normality of the maximum likelihood estimates, which is
generally adequate for these sample sizes but may be slightly liberal for
cells with very high observed proportions.

**Callset-based complementary analysis.** A parallel set of binomial GLMs
was fit using the "correct in callset" curation as the outcome variable. In
this analysis, a discrepancy is classified as a success (benchmark identified
an error) when the comparison callset is judged "not correct" (`no`). This
perspective captures cases where both the benchmark and comparison callset
may be imperfect at a site, but the benchmark nonetheless correctly flagged
an error in the comparison. The conservative analysis treats "unsure" as the
comparison being correct (fewer successes); the sensitivity analysis treats
"unsure" as the comparison being wrong (more successes). The same model
structure, predictors, and emmeans extraction were used. Bias-reduced
logistic regression was applied where quasi-complete separation was detected.

**Limitations.** The XY non-PAR strata had no discrepancies for small
variants and were excluded from the small variant model. The treatment of
"unsure" curations has a notable effect on the results, particularly for
structural variants where "unsure" curations were more common; the
sensitivity analysis brackets this uncertainty. The models do not include
callset-specific effects, so the estimated probabilities reflect an average
across all comparison callsets evaluated. The curated discrepancies were
selected from specific callsets and may not be representative of all possible
discrepancies. Finally, the additive model does not capture potential
interactions between predictors (e.g., whether TR effects differ for INS
vs. DEL in structural variants); the data would support such extensions but
at the cost of wider confidence intervals.


