---
title: "Genomic Context Coverage and Variant Density Analysis"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(
  load_plot_themes = TRUE,
  load_sessioninfo = TRUE
)
```

## Overview

This notebook analyzes the overlap between genomic context regions (difficult
regions: homopolymers, tandem repeats, segmental duplications, low mappability)
and the GIAB benchmark sets. It addresses:

1. How much of each genomic context is covered by the benchmark?
2. How much of the benchmark falls within each genomic context?
3. How do variant counts and densities differ across genomic contexts?
4. How do these patterns compare across benchmark versions, references, and variant types?

The primary data source is `genomic_context_combined_metrics.csv` produced by
the `var_counts.smk` pipeline, which combines BED overlap metrics with
per-context variant counts.

## Data Loading

```{r load_data}
# Load aggregated genomic context metrics (coverage + variant counts)
genomic_context_metrics_df <- load_genomic_context_metrics()

# Load reference sizes for normalization
ref_sizes_df <- load_reference_sizes()
total_ref_asm_bp <- ref_sizes_df %>%
  group_by(ref) %>%
  summarise(total_asm_bp = sum(asm_bp), .groups = "drop")
```

```{r}
#| echo: false
glimpse(genomic_context_metrics_df)
```

## Genomic Context Coverage Summary

This table summarizes the overlap between each genomic context region and the
benchmark regions. `pct_of_context` indicates the fraction of the difficult
region covered by the benchmark, while `pct_of_bench` indicates how much of
the benchmark lies within that difficult region.

```{r coverage_summary_table}
genomic_context_metrics_df %>%
  filter(bench_version == "v5.0q") %>%
  mutate(
    ref = std_references(ref),
    context_label = get_context_labels()[as.character(context_name)]
  ) %>%
  select(
    context_label, ref, bench_type,
    context_bp, intersect_bp,
    pct_of_context, pct_of_bench
  ) %>%
  arrange(context_label, ref, bench_type) %>%
  as_grouped_flextable(groupname_col = "context_label") %>%
  fmt_integer_flextable(columns = c("context_bp", "intersect_bp")) %>%
  fmt_percent_flextable(columns = c("pct_of_context", "pct_of_bench"), decimals = 1) %>%
  cols_label_flextable(
    ref = "Reference",
    bench_type = "Bench Type",
    context_bp = "Context Size (bp)",
    intersect_bp = "Overlap (bp)",
    pct_of_context = "% Context Covered",
    pct_of_bench = "% Bench in Context"
  ) %>%
  theme_flextable_manuscript()
```

## Benchmark Coverage of Genomic Contexts

### Percent of genomic context covered by benchmark

Shows how completely the benchmark regions cover each type of difficult region.
Higher values indicate more comprehensive benchmarking of that region type.

Use of alpha so small variant and SV are on the same plot. Ideally would like to have the small variant bar in front of SV.

```{r pct_context_covered_plot, fig.width = 8, fig.height = 5}
(pct_context_plt <- genomic_context_metrics_df %>%
  mutate(
    ref = std_references(ref),
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  ) %>%
  # filter(context_name != "TR10kb") %>%
  ggplot(aes(
    x = context_name,
    y = pct_of_context,
    fill = bench_version
  )) +
  geom_col(aes(alpha = bench_type),
    position = "dodge", color = "grey30", linewidth = 0.2
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1)
  ) +
  scale_benchmark_version(aesthetic = "fill") +
  scale_genomic_context(aesthetic = "color", guide = "none") +
  facet_grid(. ~ ref) +
  labs(
    x = "Genomic Context",
    y = "Context Included",
    fill = "Benchmark Version"
  ) +
  theme_manuscript() +
  theme(
    legend.position = "bottom"
  ))
```

### Percent of benchmark within each genomic context

Shows the fraction of the total benchmark that falls within each difficult
region type. This indicates how much of the benchmark is devoted to each
category of difficulty.

```{r pct_bench_in_context_plot, fig.width = 8, fig.height = 5}
(pct_bench_plt <- genomic_context_metrics_df %>%
  mutate(
    ref = std_references(ref),
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  ) %>%
  ggplot(aes(
    x = context_name,
    y = pct_of_bench,
    fill = bench_version,
    alpha = bench_type
  )) +
  geom_col(position = "dodge", color = "grey30", linewidth = 0.2) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_benchmark_version(aesthetic = "fill") +
  facet_grid(. ~ ref) +
  labs(
    x = "Genomic Context",
    y = "Context in Bench",
    fill = "Benchmark Version"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0),
    legend.position = "bottom"
  ))
```
```{r save_coverage_combined, fig.width = 8, fig.height = 8}
(coverage_combined_plt <- (pct_context_plt +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank())) / pct_bench_plt +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom"))

ggsave(
  filename = here::here("manuscript", "figs", "genomic_context_coverage.png"),
  plot = coverage_combined_plt,
  width = 8,
  height = 8,
  dpi = 300
)
```

## Coverage Change Across Benchmark Versions

Comparison of how the coverage of difficult regions has changed between
benchmark versions. Focuses on v5.0q versus legacy benchmarks.

```{r coverage_change_plot, fig.width = 8, fig.height = 5}
coverage_change_df <- genomic_context_metrics_df %>%
  filter(bench_version %in% c("v5.0q", "v4.2.1", "v0.6")) %>%
  select(bench_version, ref, bench_type, context_name, pct_of_context) %>%
  pivot_wider(
    names_from = bench_version,
    values_from = pct_of_context
  ) %>%
  mutate(
    change_vs_legacy = case_when(
      !is.na(v4.2.1) ~ v5.0q - v4.2.1,
      !is.na(v0.6) ~ v5.0q - v0.6,
      .default = NA_real_
    ),
    legacy_version = case_when(
      !is.na(v4.2.1) ~ "v4.2.1",
      !is.na(v0.6) ~ "v0.6",
      .default = NA_character_
    )
  ) %>%
  filter(!is.na(change_vs_legacy))

(coverage_change_plt <- coverage_change_df %>%
  mutate(
    ref = std_references(ref),
    bench_type_label = get_bench_type_labels()[as.character(bench_type)]
  ) %>%
  ggplot(aes(
    x = context_name,
    y = change_vs_legacy,
    fill = context_name
  )) +
  geom_col(color = "grey30", linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1)
  ) +
  scale_genomic_context(aesthetic = "fill") +
  facet_wrap(bench_type_label ~ ref) +
  labs(
    x = "Genomic Context",
    y = "Change in Context Coverage (v5.0q - legacy)",
    fill = "Genomic Context"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -45, hjust = 0),
    legend.position = "bottom"
  ))

ggsave(
  filename = here::here("manuscript", "figs", "genomic_context_coverage_change.png"),
  plot = coverage_change_plt,
  width = 8,
  height = 5,
  dpi = 300
)
```

## Variant Counts and Density by Genomic Context

### Variant counts per context (v5.0q)

```{r variant_count_table}
genomic_context_metrics_df %>%
  filter(bench_version == "v5.0q") %>%
  mutate(
    ref = std_references(ref),
    context_label = get_context_labels()[as.character(context_name)]
  ) %>%
  select(
    context_label, ref, bench_type,
    total_variants, variant_density_per_mb
  ) %>%
  arrange(context_label, ref, bench_type) %>%
  mutate(across(c(total_variants), ~ replace_na(., 0L))) %>%
  as_grouped_flextable(groupname_col = "context_label") %>%
  fmt_integer_flextable(columns = "total_variants") %>%
  fmt_number_flextable(columns = "variant_density_per_mb", decimals = 1) %>%
  cols_label_flextable(
    ref = "Reference",
    bench_type = "Bench Type",
    total_variants = "Total Variants",
    variant_density_per_mb = "Density (var/Mb)"
  ) %>%
  theme_flextable_manuscript()
```

Detailed variant count breakdowns by variant type, size bin, and fold change
across benchmark versions are in `benchmark_difficult.qmd`.

## Comprehensive Metrics Table

Full table of all v5.0q genomic context metrics for reference.

```{r full_metrics_table}
genomic_context_metrics_df %>%
  filter(bench_version == "v5.0q") %>%
  mutate(
    ref = std_references(ref),
    context_label = get_context_labels()[as.character(context_name)],
    intersect_mb = round(intersect_bp / 1e6, 1),
    context_mb = round(context_bp / 1e6, 1)
  ) %>%
  select(
    context_label, ref, bench_type,
    context_mb, intersect_mb,
    pct_of_context, pct_of_bench,
    total_variants, variant_density_per_mb
  ) %>%
  arrange(context_label, ref, bench_type) %>%
  as_grouped_flextable(groupname_col = "context_label") %>%
  fmt_percent_flextable(columns = c("pct_of_context", "pct_of_bench"),
    decimals = 1, scale_values = TRUE) %>%
  fmt_integer_flextable(columns = "total_variants") %>%
  fmt_number_flextable(columns = "variant_density_per_mb", decimals = 1) %>%
  cols_label_flextable(
    ref = "Reference",
    bench_type = "Bench Type",
    context_mb = "Context (Mb)",
    intersect_mb = "Overlap (Mb)",
    pct_of_context = "% Ctx Covered",
    pct_of_bench = "% Bench in Ctx",
    total_variants = "Total Variants",
    variant_density_per_mb = "Density (var/Mb)"
  ) %>%
  theme_flextable_manuscript()
```

## Session Info

```{r}
#| echo: false
sessioninfo::session_info()
```
