---
title: "Benchmarking Use-Case: v4.2.1 vs v5.0q"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)

source(here::here("scripts/happy_giab.R"))  ## imports convert_to_phred()

callset_colors <- c(
  "ONT-Sniffles (Sniffles2)" = "#2ca02c",
  "ONT-Verkko (dipcall)" = "#1f77b4"
)

size_bin_levels <- c("SNP", "1-4bp", "5-14bp", "15-49bp")
sv_size_levels <- c("[50,100)", "[100,300)", "[300,1000)", "[1000,10000)", ">=10000")

## Phred-scale ggplot2 transformation ----------------------------------------
phred_trans <- scales::trans_new(
  name      = "phred",
  transform = convert_to_phred,
  inverse   = function(y) 1 - 10^(-y / 10)
)

## Break positions and labels (in proportion scale) for each plot type
phred_smvar_pcts   <- c(0.90, 0.95, 0.99, 0.999)
phred_smvar_labels <- c("90%", "95%", "99%", "99.9%")

phred_stvar_pcts   <- c(0.5, 0.7, 0.8, 0.9, 0.95)
phred_stvar_labels <- c("50%", "70%", "80%", "90%", "95%")

## Colour and shape mappings for context labels in comparison scatter plots
context_label_colors <- c(
  "Overall"                = "black",
  "Homopolymers"           = "#E41A1C",
  "Low Mappability"        = "#377EB8",
  "Segmental Duplications" = "#4DAF4A",
  "Tandem Repeats"         = "#FF7F00"
)
context_label_shapes <- c(
  "Overall"                = 21,
  "Homopolymers"           = 24,
  "Low Mappability"        = 22,
  "Segmental Duplications" = 23,
  "Tandem Repeats"         = 25
)

## annotation_phredticks and helpers ------------------------------------------
## Adapted from:
## https://github.com/usnistgov/giab-pFDA-2nd-challenge/blob/main/scripts/annotation-phredticks.R
##
## Calculates phred-scale tick positions for annotation_phredticks().
calc_phredticks <- function(
  minphred = 0, maxphred = minphred + 1,
  start = 0, shortend = 0.1, midend = 0.2, longend = 0.3
) {
  tick_df <- tibble(
    x_pct    = c(0, seq(10, 90, 10),
                 seq(91, 99, 1),
                 seq(99.1,   99.9,   0.1),
                 seq(99.91,  99.99,  0.01),
                 seq(99.991, 99.999, 0.001)),
    ticknums = c(0, rep(1:9, times = 5))
  ) |>
    mutate(
      x       = x_pct / 100,
      phred_x = convert_to_phred(x),
      start   = start,
      tickend = case_when(
        x == 0        ~ longend,
        ticknums == 9 ~ longend,
        ticknums == 5 ~ midend,
        TRUE          ~ shortend
      )
    )
  data.frame(value = tick_df$phred_x, start = tick_df$start, end = tick_df$tickend)
}

GeomPhredticks <- ggplot2::ggproto(
  "GeomPhredticks", ggplot2::Geom,
  extra_params = "",
  handle_na = function(data, params) data,

  draw_panel = function(
    data, panel_params, coord,
    base = 10, sides = "bl", outside = FALSE, scaled = TRUE,
    short    = grid::unit(0.1, "cm"),
    mid      = grid::unit(0.2, "cm"),
    long     = grid::unit(0.3, "cm"),
    colour   = "black", size = 0.5, linetype = 1, alpha = 1
  ) {
    ticks <- list()
    short <- grid::convertUnit(short, "cm", valueOnly = TRUE)
    mid   <- grid::convertUnit(mid,   "cm", valueOnly = TRUE)
    long  <- grid::convertUnit(long,  "cm", valueOnly = TRUE)
    flipped <- inherits(coord, "CoordFlip")
    x_name  <- if (flipped) "y" else "x"
    y_name  <- if (flipped) "x" else "y"

    if (grepl("[bt]", sides)) {
      xticks <- calc_phredticks(
        minphred = floor(panel_params$x.range[1]),
        maxphred = ceiling(panel_params$x.range[2]),
        start = 0, shortend = short, midend = mid, longend = long
      )
      names(xticks)[names(xticks) == "value"] <- x_name
      xticks <- coord$transform(xticks, panel_params)
      xticks <- xticks[xticks$x >= 0 & xticks$x <= 1, ]
      if (outside) xticks$end <- -xticks$end

      if (grepl("b", sides))
        ticks$x_b <- grid::segmentsGrob(
          x0 = grid::unit(xticks$x, "native"), x1 = grid::unit(xticks$x, "native"),
          y0 = grid::unit(xticks$start, "cm"), y1 = grid::unit(xticks$end, "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
      if (grepl("t", sides))
        ticks$x_t <- grid::segmentsGrob(
          x0 = grid::unit(xticks$x, "native"), x1 = grid::unit(xticks$x, "native"),
          y0 = grid::unit(1, "npc") - grid::unit(xticks$start, "cm"),
          y1 = grid::unit(1, "npc") - grid::unit(xticks$end,   "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
    }

    if (grepl("[lr]", sides)) {
      yticks <- calc_phredticks(
        minphred = floor(panel_params$y.range[1]),
        maxphred = ceiling(panel_params$y.range[2]),
        start = 0, shortend = short, midend = mid, longend = long
      )
      names(yticks)[names(yticks) == "value"] <- y_name
      yticks <- coord$transform(yticks, panel_params)
      yticks <- yticks[yticks$y >= 0 & yticks$y <= 1, ]
      if (outside) yticks$end <- -yticks$end

      if (grepl("l", sides))
        ticks$y_l <- grid::segmentsGrob(
          y0 = grid::unit(yticks$y, "native"), y1 = grid::unit(yticks$y, "native"),
          x0 = grid::unit(yticks$start, "cm"), x1 = grid::unit(yticks$end, "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
      if (grepl("r", sides))
        ticks$y_r <- grid::segmentsGrob(
          y0 = grid::unit(yticks$y, "native"), y1 = grid::unit(yticks$y, "native"),
          x0 = grid::unit(1, "npc") - grid::unit(yticks$start, "cm"),
          x1 = grid::unit(1, "npc") - grid::unit(yticks$end,   "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
    }

    grid::gTree(children = do.call(grid::gList, ticks))
  },

  default_aes = ggplot2::aes(colour = "black", size = 0.5, linetype = 1, alpha = 1)
)

annotation_phredticks <- function(
  base = 10, sides = "bl", outside = FALSE, scaled = TRUE,
  short    = grid::unit(0.1, "cm"),
  mid      = grid::unit(0.2, "cm"),
  long     = grid::unit(0.3, "cm"),
  colour = "black", size = 0.5, linetype = 1, alpha = 1, color = NULL, ...
) {
  if (!is.null(color)) colour <- color
  ggplot2::layer(
    data        = data.frame(x = 1),
    mapping     = NULL,
    stat        = ggplot2::StatIdentity,
    geom        = GeomPhredticks,
    position    = ggplot2::PositionIdentity,
    show.legend = FALSE,
    inherit.aes = FALSE,
    params = list(
      base     = base,     sides    = sides,    outside  = outside,
      scaled   = scaled,   short    = short,    mid      = mid,
      long     = long,     colour   = colour,   size     = size,
      linetype = linetype, alpha    = alpha,
      ...
    )
  )
}
```

## Small Variant Benchmarking: Roche Neusomatic

Comparison of Roche Neusomatic (pFDA Truth Challenge V2) callset against
v4.2.1 and v5.0q small variant benchmarks on GRCh38, using GIAB v3.6
stratifications. Evaluated with `rtg vcfeval`.

```{r smvar_data}
## Helper to compute recall/precision from TP/FP/FN counts
compute_metrics <- function(df) {
  df %>%
    mutate(
      recall = tp / (tp + fn),
      precision = tp / (tp + fp),
      f1 = 2 * (precision * recall) / (precision + recall)
    )
}

## Load vcfeval ROC metrics (from extract_smvar_metrics.py)
## 8-column ROC files (Overall, SNP, NON_SNP): have fn, precision, sensitivity
## 4-column ROC files (HP, TR, SD, MAP): have tp_baseline and fp only
smvar_roc_df <- read_csv(
  here("results", "use_case", "smvar", "vcfeval_roc_metrics.csv"),
  show_col_types = FALSE
)

## Overall and per-type metrics (from 8-column ROC files)
smvar_full_df <- smvar_roc_df %>%
  filter(subset %in% c("Overall", "SNP", "NON_SNP")) %>%
  mutate(
    recall = sensitivity,
    f1 = f_measure,
    var_type_label = case_when(
      subset == "SNP" ~ "SNV",
      subset == "NON_SNP" ~ "INDEL",
      .default = NA_character_
    )
  )

## Per-context metrics (from 4-column ROC files — no FN/recall)
smvar_context_df <- smvar_roc_df %>%
  filter(subset %in% c("HP", "TR", "SD", "MAP")) %>%
  mutate(
    context_label = get_context_labels()[subset],
    context_label = factor(
      context_label,
      levels = c("Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    )
  )

## By size bin
smvar_size_df <- read_csv(
  here("results", "use_case", "smvar", "size_bin_counts.csv"),
  show_col_types = FALSE
) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  filter(size_bin %in% size_bin_levels) %>%
  mutate(size_bin = factor(size_bin, levels = size_bin_levels))
```

### Benchmark Comparison: v4.2.1 vs v5.0q by Variant Type

Overall and per-type metrics from vcfeval's 8-column ROC output (full metrics
including FN, precision, recall).

```{r smvar_type_scatter, fig.width = 10, fig.height = 7}
smvar_type_compare_df <- smvar_full_df |>
  filter(!is.na(var_type_label)) |>
  select(bench_version, var_type_label, recall, precision, f1) |>
  pivot_longer(c(recall, precision, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = bench_version, values_from = value) |>
  rename(v421 = `v4.2.1`, v50q = `v5.0q`) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

## Overall metrics for annotation
smvar_overall_compare_df <- smvar_full_df |>
  filter(subset == "Overall") |>
  select(bench_version, recall, precision, f1) |>
  pivot_longer(c(recall, precision, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = bench_version, values_from = value) |>
  rename(v421 = `v4.2.1`, v50q = `v5.0q`) |>
  mutate(
    metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")),
    var_type_label = "Overall"
  )

smvar_all_compare_df <- bind_rows(smvar_type_compare_df, smvar_overall_compare_df) |>
  mutate(var_type_label = factor(var_type_label, levels = c("Overall", "SNV", "INDEL")))

smvar_type_xy_range <- range(c(smvar_all_compare_df$v421, smvar_all_compare_df$v50q), na.rm = TRUE)

(smvar_type_plt <- smvar_all_compare_df |>
  ggplot(aes(x = v421, y = v50q, fill = var_type_label, shape = var_type_label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(color = "grey30", size = 3.5, stroke = 0.8) +
  facet_wrap(~metric) +
  scale_x_continuous(
    trans = phred_trans,
    breaks = phred_smvar_pcts, labels = phred_smvar_labels,
    limits = smvar_type_xy_range
  ) +
  scale_y_continuous(
    trans = phred_trans,
    breaks = phred_smvar_pcts, labels = phred_smvar_labels,
    limits = smvar_type_xy_range
  ) +
  annotation_phredticks(sides = "bl") +
  scale_fill_manual(values = c("Overall" = "black", "SNV" = "#1f77b4", "INDEL" = "#ff7f0e")) +
  scale_shape_manual(values = c("Overall" = 21, "SNV" = 24, "INDEL" = 22)) +
  labs(
    x = "v4.2.1",
    y = "v5.0q",
    fill = "Variant Type",
    shape = "Variant Type"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### Per-Context TP and FP (4-column ROC data)

Per-context metrics from vcfeval's 4-column ROC output. These provide
TP (baseline-rescaled) and FP directly from vcfeval's matching engine.
FN and recall are not available from these files.

```{r smvar_context_table}
smvar_context_df |>
  select(bench_version, context_label, tp_baseline, fp) |>
  arrange(bench_version, context_label) |>
  gt::gt() |>
  gt::fmt_number(columns = c(tp_baseline, fp), decimals = 1) |>
  gt::cols_label(
    bench_version = "Benchmark",
    context_label = "Context",
    tp_baseline = "TP (baseline)",
    fp = "FP"
  ) |>
  theme_gt_manuscript()
```

### By Variant Size

```{r smvar_size_plot, fig.width = 7, fig.height = 5}
smvar_size_long <- smvar_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = str_to_title(metric),
    size_bin_num = as.numeric(size_bin)
  )

(smvar_size_plt <- smvar_size_long %>%
  ggplot(aes(x = size_bin_num, y = value, color = bench_version, group = bench_version)) +
  geom_line() +
  geom_point(aes(fill = bench_version), shape = 21, color = "grey30", size = 2.5) +
  facet_wrap(~metric) +
  scale_x_continuous(
    breaks = seq_along(size_bin_levels),
    labels = size_bin_levels
  ) +
  scale_y_continuous(
    trans   = phred_trans,
    breaks  = phred_smvar_pcts,
    labels  = phred_smvar_labels,
    limits  = c(0.85, 0.9999),
    oob     = scales::squish
  ) +
  scale_benchmark_version(aesthetic = "color", guide = "none") +
  scale_benchmark_version(aesthetic = "fill") +
  labs(x = "Variant Size", y = NULL, fill = "Benchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### Combined Small Variant Figure

```{r combined_smvar, fig.width = 8, fig.height = 10}
design <- c(area(l = 1, r = 8, t = 1, b = 8), area(l = 1, r = 8, t = 8, b = 10))
(combined_smvar_plt <-
  # (smvar_scatter_plt + theme(legend.position = "bottom")) /
  (smvar_type_plt) /
  (smvar_size_plt) &
  plot_annotation(tag_levels = "A") &
  plot_layout(design = design)
 )


ggsave(
  filename = here("manuscript", "figs", "use_case_smvar.png"),
  plot = combined_smvar_plt,
  width = 8,
  height = 10,
  dpi = 300
)
```

### Figure Legend

**Figure use_case_smvar: Small variant benchmarking of the Roche NeuSomatic callset against v4.2.1 and v5.0q benchmark sets.**
(A) Benchmark comparison scatter plots for the Roche NeuSomatic HG002 callset (Illumina + PacBio + ONT input data) evaluated against the v4.2.1 and v5.0q GRCh38 small variant benchmarks using `rtg vcfeval`. Each panel shows one performance metric (recall, precision, F1); each point represents a variant type (overall, SNV, INDEL) with metrics computed from vcfeval's ROC output. Axes are phred-scaled; the dashed diagonal marks equal performance between the two benchmarks.
(B) Recall and precision by variant size bin (SNP, 1–4 bp, 5–14 bp, 15–49 bp) on a phred-scaled y-axis.

### Results Text

To demonstrate the practical impact of the expanded v5.0q benchmark regions on variant caller evaluation, we benchmarked the Roche NeuSomatic HG002 callset from the precisionFDA Truth Challenge V2 against both the v4.2.1 and v5.0q GRCh38 small variant benchmark sets using `rtg vcfeval` with `--roc-subset snp,non-snp` for per-type stratification.

Against the v4.2.1 benchmark, the NeuSomatic callset achieved near-perfect performance overall (F1 = 0.9997, recall = 99.96%, precision = 99.98%), with only 612 false positives and 1,685 false negatives across 3.89 million true positives. Both SNVs (F1 = 0.9998) and INDELs (F1 = 0.9988) showed high concordance, reflecting the limited scope of the v4.2.1 benchmark which largely excluded the most challenging repetitive regions.

Against the v5.0q benchmark, the same callset showed substantially lower performance (F1 = 0.9827, recall = 97.99%, precision = 98.55%), with 64,931 false positives and 90,686 false negatives across 4.43 million true positives. Performance differences were particularly pronounced for INDELs (F1 = 0.9329, recall = 92.56%, precision = 94.03%) compared to SNVs (F1 = 0.9947, recall = 99.30%, precision = 99.65%). By variant size, the largest performance decrease was observed for INDELs in the 5–14 bp and 15–49 bp size ranges, where the v5.0q benchmark includes substantially more variants in repetitive regions.


## Structural Variant Benchmarking: ONT-Sniffles vs ONT-Verkko

Comparison of ONT-Sniffles (Sniffles2, mapping-based) and ONT-Verkko (dipcall, assembly-based)
SV callsets against the v5.0q GRCh38 structural variant benchmark.
Evaluated with `truvari bench` and `truvari refine` using GIAB v5.0q recommended parameters (`--pick ac --passonly -r 2000 -C 5000 --refine`).

```{r stvar_data}
stvar_df <- read_csv(
  here("results", "use_case", "stvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    callset_label = case_when(
      callset == "ont-sniffles" ~ "ONT-Sniffles (Sniffles2)",
      callset == "ont-verkko" ~ "ONT-Verkko (dipcall)"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV type-level metrics
svtype_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    callset_label = case_when(
      callset == "ont-sniffles" ~ "ONT-Sniffles (Sniffles2)",
      callset == "ont-verkko" ~ "ONT-Verkko (dipcall)"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV size-binned metrics
sv_size_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_size_counts.csv"),
  show_col_types = FALSE
) %>%
  filter(svtype %in% c("INS", "DEL")) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  mutate(
    size_bin = factor(size_bin, levels = sv_size_levels),
    callset_label = case_when(
      callset == "ont-sniffles" ~ "ONT-Sniffles (Sniffles2)",
      callset == "ont-verkko" ~ "ONT-Verkko (dipcall)"
    )
  )
```

### Callset Comparison Scatter: ONT-Sniffles vs ONT-Verkko by Context

```{r stvar_scatter, fig.width = 10, fig.height = 4}
stvar_compare_df <- stvar_df |>
  select(callset, context_label, precision, recall, f1) |>
  pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = callset, values_from = value) |>
  rename(ont_sniffles = `ont-sniffles`, ont_verkko = `ont-verkko`) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

stvar_xy_range <- range(c(stvar_compare_df$ont_sniffles, stvar_compare_df$ont_verkko), na.rm = TRUE)

(stvar_scatter_plt <- stvar_compare_df |>
  ggplot(aes(x = ont_sniffles, y = ont_verkko, fill = context_label, shape = context_label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(color = "grey30", size = 3.5, stroke = 0.8) +
  facet_wrap(~metric) +
  scale_x_continuous(
    trans = phred_trans,
    breaks = phred_stvar_pcts, labels = phred_stvar_labels,
    limits = stvar_xy_range
  ) +
  scale_y_continuous(
    trans = phred_trans,
    breaks = phred_stvar_pcts, labels = phred_stvar_labels,
    limits = stvar_xy_range
  ) +
  annotation_phredticks(sides = "bl") +
  scale_fill_manual(values = context_label_colors) +
  scale_shape_manual(values = context_label_shapes) +
  labs(
    x = "ONT-Sniffles (Sniffles2)",
    y = "ONT-Verkko (dipcall)",
    fill = "Genomic Context",
    shape = "Genomic Context"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### By SV Type and Size

```{r stvar_size_plot, fig.width = 8, fig.height = 5}
sv_size_long <- sv_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(metric = str_to_title(metric))

(stvar_size_plt <- sv_size_long %>%
  ggplot(aes(x = size_bin, y = value, fill = callset_label)) +
  geom_col(
    position = position_dodge(width = 0.8), width = 0.7,
    color = "grey30", linewidth = 0.3
  ) +
  facet_grid(svtype ~ metric) +
  scale_y_continuous(
    trans   = phred_trans,
    breaks  = phred_stvar_pcts,
    labels  = phred_stvar_labels,
    limits  = c(0, 0.99)
  ) +
  scale_fill_manual(values = callset_colors) +
  labs(x = "SV Size", y = NULL, fill = "Callset") +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -30, hjust = 0),
    legend.position = "bottom"
  ))
```

### Callset Comparison Scatter: ONT-Sniffles vs ONT-Verkko by SV Type and Size

```{r stvar_size_scatter, fig.width = 10, fig.height = 7}
sv_size_compare_df <- sv_size_df |>
  select(callset, svtype, size_bin, precision, recall, f1) |>
  pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = callset, values_from = value) |>
  rename(ont_sniffles = `ont-sniffles`, ont_verkko = `ont-verkko`) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

stvar_size_xy_vals <- c(sv_size_compare_df$ont_sniffles, sv_size_compare_df$ont_verkko)
stvar_size_xy_range <- range(stvar_size_xy_vals[!is.na(stvar_size_xy_vals) & stvar_size_xy_vals < 1])

(stvar_size_scatter_plt <- sv_size_compare_df |>
  ggplot(aes(x = ont_sniffles, y = ont_verkko, fill = svtype, size = size_bin)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(shape = 21, color = "grey30", stroke = 0.6) +
  facet_wrap(~metric) +
  # scale_x_continuous(
  #   trans = phred_trans,
  #   breaks = phred_stvar_pcts, labels = phred_stvar_labels,
  #   limits = stvar_size_xy_range
  # ) +
  # scale_y_continuous(
  #   trans = phred_trans,
  #   breaks = phred_stvar_pcts, labels = phred_stvar_labels,
  #   limits = stvar_size_xy_range
  # ) +
  # annotation_phredticks(sides = "bl") +
  scale_fill_manual(values = c("DEL" = "#D95F02", "INS" = "#1B9E77")) +
  # scale_size_manual(values = c("50-99bp" = 2, "100-299bp" = 3, "300-999bp" = 4, "1-10kb" = 5, ">=10kb" = 6)) +
  labs(
    x = "ONT-Sniffles (Sniffles2)",
    y = "ONT-Verkko (dipcall)",
    fill = "SV Type",
    size = "SV Size"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### Combined SV Figure

```{r combined_stvar, fig.width = 8, fig.height = 8}
(combined_stvar_plt <-
  (stvar_scatter_plt + theme(legend.position = "bottom")) /
  (stvar_size_scatter_plt + theme(legend.position = "bottom")) +
  plot_annotation(tag_levels = "A") &
  theme(margins = unit(c(1, 1, 1, 1), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_stvar.png"),
  plot = combined_stvar_plt,
  width = 8,
  height = 8,
  dpi = 300
)
```

### Figure Legend

**Figure use_case_stvar: Structural variant benchmarking of ONT-Sniffles (Sniffles2, mapping-based) and ONT-Verkko (dipcall, assembly-based) callsets against the v5.0q GRCh38 SV benchmark.**
(A) Callset comparison scatter plots for overall and stratified performance (homopolymers, low mappability, segmental duplications, tandem repeats) using GIAB v3.6 stratifications and `truvari bench` with GIAB-recommended parameters (`--pick ac --passonly -r 2000 -C 5000`) followed by `truvari refine`. Each panel shows one performance metric (recall, precision, F1); x-axis is ONT-Sniffles (Sniffles2), y-axis is ONT-Verkko (dipcall). Axes are phred-scaled; the dashed diagonal marks equal performance between callsets.
(B) Recall and precision by SV type (DEL and INS) and size bin on a phred-scaled y-axis.
(C) Callset comparison scatter by SV type and size bin, faceted by SV type and performance metric. Axes are phred-scaled.

### Results Text

To demonstrate the practical impact of SV calling methodology on benchmarking performance, we compared mapping-based (ONT-Sniffles2) and assembly-based (ONT-Verkko dipcall) structural variant callsets against the v5.0q GRCh38 benchmark using `truvari bench` with GIAB-recommended parameters (`--pick ac --passonly -r 2000 -C 5000`) followed by `truvari refine` for MSA-based re-evaluation. ALT="*" spanning deletion records were filtered prior to benchmarking per GIAB v5.0q README guidance.

[Add performance comparison results after reviewing actual metrics]
