---
title: "Benchmarking Use-Case: v4.2.1 vs v5.0q"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)

callset_colors <- c(
  "Baylor ONT (Sniffles2)" = "#2ca02c",
  "Illumina DRAGEN" = "#1f77b4"
)

size_bin_levels <- c("SNP", "1-4bp", "5-14bp", "15-49bp")
sv_size_levels <- c("50-99bp", "100-299bp", "300-999bp", "1-10kb", ">=10kb")
```

## Small Variant Benchmarking: Roche Neusomatic

Comparison of Roche Neusomatic (pFDA Truth Challenge V2) callset against
v4.2.1 and v5.0q small variant benchmarks on GRCh38, using GIAB v3.6
stratifications. Evaluated with `rtg vcfeval`.

```{r smvar_data}
## Helper to compute recall/precision from TP/FP/FN counts
compute_metrics <- function(df) {
  df %>%
    mutate(
      recall = tp / (tp + fn),
      precision = tp / (tp + fp),
      f1 = 2 * (precision * recall) / (precision + recall)
    )
}

smvar_df <- read_csv(
  here("results", "use_case", "smvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## By variant type and context
smvar_type_df <- read_csv(
  here("results", "use_case", "smvar", "stratified_by_type.csv"),
  show_col_types = FALSE
) %>%
  compute_metrics() %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    var_type_label = if_else(var_type == "SNP", "SNV", "INDEL")
  )

## By size bin
smvar_size_df <- read_csv(
  here("results", "use_case", "smvar", "size_bin_counts.csv"),
  show_col_types = FALSE
) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  filter(size_bin %in% size_bin_levels) %>%
  mutate(size_bin = factor(size_bin, levels = size_bin_levels))
```

### Precision-Recall Scatter: v4.2.1 vs v5.0q by Context

```{r smvar_scatter, fig.width = 7, fig.height = 5}
(smvar_scatter_plt <- smvar_df %>%
  ggplot(aes(x = recall, y = precision, color = bench_version)) +
  geom_point(aes(shape = context_label), size = 3.5, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = context_label),
    size = 2.5, max.overlaps = 15, seed = 42,
    show.legend = FALSE
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0.90, 1.002)) +
  scale_y_continuous(labels = scales::percent, limits = c(0.90, 1.002)) +
  scale_benchmark_version(aesthetic = "color") +
  scale_shape_manual(values = c(16, 17, 15, 18, 8)) +
  labs(
    x = "Recall",
    y = "Precision",
    color = "Benchmark",
    shape = "Genomic Context"
  ) +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### By Variant Type and Context

```{r smvar_type_scatter, fig.width = 8, fig.height = 5}
(smvar_type_plt <- smvar_type_df %>%
  ggplot(aes(x = recall, y = precision, color = bench_version)) +
  geom_point(aes(shape = context_label), size = 3, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = context_label),
    size = 2, max.overlaps = 20, seed = 42,
    show.legend = FALSE
  ) +
  facet_wrap(~var_type_label) +
  scale_x_continuous(labels = scales::percent, limits = c(0.85, 1.005)) +
  scale_y_continuous(labels = scales::percent, limits = c(0.85, 1.005)) +
  scale_benchmark_version(aesthetic = "color") +
  scale_shape_manual(values = c(16, 17, 15, 18, 8)) +
  labs(
    x = "Recall",
    y = "Precision",
    color = "Benchmark",
    shape = "Genomic Context"
  ) +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### By Variant Size

```{r smvar_size_plot, fig.width = 7, fig.height = 5}
smvar_size_long <- smvar_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(metric = str_to_title(metric))

(smvar_size_plt <- smvar_size_long %>%
  ggplot(aes(x = size_bin, y = value, fill = bench_version)) +
  geom_col(
    position = position_dodge(width = 0.8), width = 0.7,
    color = "grey30", linewidth = 0.3
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", value * 100)),
    position = position_dodge(width = 0.8),
    vjust = -0.3, size = 2.5
  ) +
  facet_wrap(~metric) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0.85, 1.01),
    oob = scales::squish,
    breaks = seq(0.85, 1.0, 0.05)
  ) +
  scale_benchmark_version(aesthetic = "fill") +
  labs(x = "Variant Size", y = NULL, fill = "Benchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### Combined Small Variant Figure

```{r combined_smvar, fig.width = 10, fig.height = 10}
(combined_smvar_plt <-
  (smvar_scatter_plt + theme(legend.position = "bottom")) /
  smvar_type_plt /
  smvar_size_plt +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(5, 5, 5, 5), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_smvar.png"),
  plot = combined_smvar_plt,
  width = 10,
  height = 10,
  dpi = 300
)
```

## Structural Variant Benchmarking: Baylor ONT vs Illumina DRAGEN

Comparison of Baylor ONT (Sniffles2) and Illumina DRAGEN SV callsets
against the v5.0q GRCh38 structural variant benchmark.
Evaluated with `truvari bench`.

```{r stvar_data}
stvar_df <- read_csv(
  here("results", "use_case", "stvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV type-level metrics
svtype_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV size-binned metrics
sv_size_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_size_counts.csv"),
  show_col_types = FALSE
) %>%
  filter(svtype %in% c("INS", "DEL")) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  mutate(
    size_bin = factor(size_bin, levels = sv_size_levels),
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    )
  )
```

### Precision-Recall Scatter: ONT vs DRAGEN by Context

```{r stvar_scatter, fig.width = 7, fig.height = 5}
(stvar_scatter_plt <- stvar_df %>%
  ggplot(aes(x = recall, y = precision, color = callset_label)) +
  geom_point(aes(shape = context_label), size = 3.5, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = context_label),
    size = 2.5, max.overlaps = 15, seed = 42,
    show.legend = FALSE
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = callset_colors) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8)) +
  labs(
    x = "Recall",
    y = "Precision",
    color = "Callset",
    shape = "Genomic Context"
  ) +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### By SV Type and Size

```{r stvar_size_plot, fig.width = 8, fig.height = 5}
sv_size_long <- sv_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(metric = str_to_title(metric))

(stvar_size_plt <- sv_size_long %>%
  ggplot(aes(x = size_bin, y = value, fill = callset_label)) +
  geom_col(
    position = position_dodge(width = 0.8), width = 0.7,
    color = "grey30", linewidth = 0.3
  ) +
  facet_grid(svtype ~ metric) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
  scale_fill_manual(values = callset_colors) +
  labs(x = "SV Size", y = NULL, fill = "Callset") +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -30, hjust = 0),
    legend.position = "bottom"
  ))
```

### Precision-Recall Scatter: By SV Type and Size

```{r stvar_size_scatter, fig.width = 7, fig.height = 5}
(stvar_size_scatter_plt <- sv_size_df %>%
  ggplot(aes(x = recall, y = precision, color = callset_label)) +
  geom_point(aes(shape = svtype), size = 3, stroke = 0.8) +
  geom_line(aes(group = interaction(callset_label, svtype)),
            linewidth = 0.4, alpha = 0.5) +
  ggrepel::geom_text_repel(
    aes(label = size_bin),
    size = 2, max.overlaps = 20, seed = 42,
    show.legend = FALSE
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = callset_colors) +
  scale_shape_manual(values = c("DEL" = 17, "INS" = 16)) +
  labs(
    x = "Recall",
    y = "Precision",
    color = "Callset",
    shape = "SV Type"
  ) +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### Combined SV Figure

```{r combined_stvar, fig.width = 10, fig.height = 12}
(combined_stvar_plt <-
  (stvar_scatter_plt + theme(legend.position = "bottom")) /
  stvar_size_plt /
  (stvar_size_scatter_plt + theme(legend.position = "bottom")) +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(5, 5, 5, 5), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_stvar.png"),
  plot = combined_stvar_plt,
  width = 10,
  height = 12,
  dpi = 300
)
```
