---
title: "Benchmarking Use-Case: v4.2.1 vs v5.0q"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)

source(here::here("scripts/happy_giab.R"))  ## imports convert_to_phred()

callset_colors <- c(
  "Baylor ONT (Sniffles2)" = "#2ca02c",
  "Illumina DRAGEN" = "#1f77b4"
)

size_bin_levels <- c("SNP", "1-4bp", "5-14bp", "15-49bp")
sv_size_levels <- c("50-99bp", "100-299bp", "300-999bp", "1-10kb", ">=10kb")

## Phred-scale ggplot2 transformation ----------------------------------------
phred_trans <- scales::trans_new(
  name      = "phred",
  transform = convert_to_phred,
  inverse   = function(y) 1 - 10^(-y / 10)
)

## Break positions and labels (in proportion scale) for each plot type
phred_smvar_pcts   <- c(0.90, 0.95, 0.99, 0.999)
phred_smvar_labels <- c("90%", "95%", "99%", "99.9%")

phred_stvar_pcts   <- c(0.5, 0.7, 0.8, 0.9, 0.95)
phred_stvar_labels <- c("50%", "70%", "80%", "90%", "95%")

## Colour and shape mappings for context labels in comparison scatter plots
context_label_colors <- c(
  "Overall"                = "black",
  "Homopolymers"           = "#E41A1C",
  "Low Mappability"        = "#377EB8",
  "Segmental Duplications" = "#4DAF4A",
  "Tandem Repeats"         = "#FF7F00"
)
context_label_shapes <- c(
  "Overall"                = 16,
  "Homopolymers"           = 17,
  "Low Mappability"        = 15,
  "Segmental Duplications" = 18,
  "Tandem Repeats"         = 8
)

## annotation_phredticks and helpers ------------------------------------------
## Adapted from:
## https://github.com/usnistgov/giab-pFDA-2nd-challenge/blob/main/scripts/annotation-phredticks.R
##
## Calculates phred-scale tick positions for annotation_phredticks().
calc_phredticks <- function(
  minphred = 0, maxphred = minphred + 1,
  start = 0, shortend = 0.1, midend = 0.2, longend = 0.3
) {
  tick_df <- tibble(
    x_pct    = c(0, seq(10, 90, 10),
                 seq(91, 99, 1),
                 seq(99.1,   99.9,   0.1),
                 seq(99.91,  99.99,  0.01),
                 seq(99.991, 99.999, 0.001)),
    ticknums = c(0, rep(1:9, times = 5))
  ) |>
    mutate(
      x       = x_pct / 100,
      phred_x = convert_to_phred(x),
      start   = start,
      tickend = case_when(
        x == 0        ~ longend,
        ticknums == 9 ~ longend,
        ticknums == 5 ~ midend,
        TRUE          ~ shortend
      )
    )
  data.frame(value = tick_df$phred_x, start = tick_df$start, end = tick_df$tickend)
}

GeomPhredticks <- ggplot2::ggproto(
  "GeomPhredticks", ggplot2::Geom,
  extra_params = "",
  handle_na = function(data, params) data,

  draw_panel = function(
    data, panel_params, coord,
    base = 10, sides = "bl", outside = FALSE, scaled = TRUE,
    short    = grid::unit(0.1, "cm"),
    mid      = grid::unit(0.2, "cm"),
    long     = grid::unit(0.3, "cm"),
    colour   = "black", size = 0.5, linetype = 1, alpha = 1
  ) {
    ticks <- list()
    short <- grid::convertUnit(short, "cm", valueOnly = TRUE)
    mid   <- grid::convertUnit(mid,   "cm", valueOnly = TRUE)
    long  <- grid::convertUnit(long,  "cm", valueOnly = TRUE)
    flipped <- inherits(coord, "CoordFlip")
    x_name  <- if (flipped) "y" else "x"
    y_name  <- if (flipped) "x" else "y"

    if (grepl("[bt]", sides)) {
      xticks <- calc_phredticks(
        minphred = floor(panel_params$x.range[1]),
        maxphred = ceiling(panel_params$x.range[2]),
        start = 0, shortend = short, midend = mid, longend = long
      )
      names(xticks)[names(xticks) == "value"] <- x_name
      xticks <- coord$transform(xticks, panel_params)
      xticks <- xticks[xticks$x >= 0 & xticks$x <= 1, ]
      if (outside) xticks$end <- -xticks$end

      if (grepl("b", sides))
        ticks$x_b <- grid::segmentsGrob(
          x0 = grid::unit(xticks$x, "native"), x1 = grid::unit(xticks$x, "native"),
          y0 = grid::unit(xticks$start, "cm"), y1 = grid::unit(xticks$end, "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
      if (grepl("t", sides))
        ticks$x_t <- grid::segmentsGrob(
          x0 = grid::unit(xticks$x, "native"), x1 = grid::unit(xticks$x, "native"),
          y0 = grid::unit(1, "npc") - grid::unit(xticks$start, "cm"),
          y1 = grid::unit(1, "npc") - grid::unit(xticks$end,   "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
    }

    if (grepl("[lr]", sides)) {
      yticks <- calc_phredticks(
        minphred = floor(panel_params$y.range[1]),
        maxphred = ceiling(panel_params$y.range[2]),
        start = 0, shortend = short, midend = mid, longend = long
      )
      names(yticks)[names(yticks) == "value"] <- y_name
      yticks <- coord$transform(yticks, panel_params)
      yticks <- yticks[yticks$y >= 0 & yticks$y <= 1, ]
      if (outside) yticks$end <- -yticks$end

      if (grepl("l", sides))
        ticks$y_l <- grid::segmentsGrob(
          y0 = grid::unit(yticks$y, "native"), y1 = grid::unit(yticks$y, "native"),
          x0 = grid::unit(yticks$start, "cm"), x1 = grid::unit(yticks$end, "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
      if (grepl("r", sides))
        ticks$y_r <- grid::segmentsGrob(
          y0 = grid::unit(yticks$y, "native"), y1 = grid::unit(yticks$y, "native"),
          x0 = grid::unit(1, "npc") - grid::unit(yticks$start, "cm"),
          x1 = grid::unit(1, "npc") - grid::unit(yticks$end,   "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
    }

    grid::gTree(children = do.call(grid::gList, ticks))
  },

  default_aes = ggplot2::aes(colour = "black", size = 0.5, linetype = 1, alpha = 1)
)

annotation_phredticks <- function(
  base = 10, sides = "bl", outside = FALSE, scaled = TRUE,
  short    = grid::unit(0.1, "cm"),
  mid      = grid::unit(0.2, "cm"),
  long     = grid::unit(0.3, "cm"),
  colour = "black", size = 0.5, linetype = 1, alpha = 1, color = NULL, ...
) {
  if (!is.null(color)) colour <- color
  ggplot2::layer(
    data        = data.frame(x = 1),
    mapping     = NULL,
    stat        = ggplot2::StatIdentity,
    geom        = GeomPhredticks,
    position    = ggplot2::PositionIdentity,
    show.legend = FALSE,
    inherit.aes = FALSE,
    params = list(
      base     = base,     sides    = sides,    outside  = outside,
      scaled   = scaled,   short    = short,    mid      = mid,
      long     = long,     colour   = colour,   size     = size,
      linetype = linetype, alpha    = alpha,
      ...
    )
  )
}
```

## Small Variant Benchmarking: Roche Neusomatic

Comparison of Roche Neusomatic (pFDA Truth Challenge V2) callset against
v4.2.1 and v5.0q small variant benchmarks on GRCh38, using GIAB v3.6
stratifications. Evaluated with `rtg vcfeval`.

```{r smvar_data}
## Helper to compute recall/precision from TP/FP/FN counts
compute_metrics <- function(df) {
  df %>%
    mutate(
      recall = tp / (tp + fn),
      precision = tp / (tp + fp),
      f1 = 2 * (precision * recall) / (precision + recall)
    )
}

smvar_df <- read_csv(
  here("results", "use_case", "smvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## By variant type and context
smvar_type_df <- read_csv(
  here("results", "use_case", "smvar", "stratified_by_type.csv"),
  show_col_types = FALSE
) %>%
  compute_metrics() %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    var_type_label = if_else(var_type == "SNP", "SNV", "INDEL")
  )

## By size bin
smvar_size_df <- read_csv(
  here("results", "use_case", "smvar", "size_bin_counts.csv"),
  show_col_types = FALSE
) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  filter(size_bin %in% size_bin_levels) %>%
  mutate(size_bin = factor(size_bin, levels = size_bin_levels))
```

### Benchmark Comparison Scatter: v4.2.1 vs v5.0q by Context

```{r smvar_scatter, fig.width = 10, fig.height = 4}
smvar_compare_df <- smvar_df |>
  select(bench_version, context_label, precision, recall, f1) |>
  pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = bench_version, values_from = value) |>
  rename(v421 = `v4.2.1`, v50q = `v5.0q`) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

(smvar_scatter_plt <- smvar_compare_df |>
  ggplot(aes(x = v421, y = v50q, color = context_label, shape = context_label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(size = 3.5, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = context_label),
    size = 2.5, max.overlaps = 15, seed = 42,
    show.legend = FALSE
  ) +
  facet_wrap(~metric) +
  scale_x_continuous(
    trans = phred_trans,
    breaks = phred_smvar_pcts, labels = phred_smvar_labels
  ) +
  scale_y_continuous(
    trans = phred_trans,
    breaks = phred_smvar_pcts, labels = phred_smvar_labels
  ) +
  annotation_phredticks(sides = "bl") +
  scale_color_manual(values = context_label_colors) +
  scale_shape_manual(values = context_label_shapes) +
  labs(
    x = "v4.2.1",
    y = "v5.0q",
    color = "Genomic Context",
    shape = "Genomic Context"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### By Variant Type and Context

```{r smvar_type_scatter, fig.width = 10, fig.height = 7}
smvar_type_compare_df <- smvar_type_df |>
  select(bench_version, var_type_label, context_label, precision, recall, f1) |>
  pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = bench_version, values_from = value) |>
  rename(v421 = `v4.2.1`, v50q = `v5.0q`) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

(smvar_type_plt <- smvar_type_compare_df |>
  ggplot(aes(x = v421, y = v50q, color = context_label, shape = context_label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(size = 3, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = context_label),
    size = 2, max.overlaps = 20, seed = 42,
    show.legend = FALSE
  ) +
  facet_grid(var_type_label ~ metric) +
  scale_x_continuous(
    trans = phred_trans,
    breaks = phred_smvar_pcts, labels = phred_smvar_labels
  ) +
  scale_y_continuous(
    trans = phred_trans,
    breaks = phred_smvar_pcts, labels = phred_smvar_labels
  ) +
  annotation_phredticks(sides = "bl") +
  scale_color_manual(values = context_label_colors) +
  scale_shape_manual(values = context_label_shapes) +
  labs(
    x = "v4.2.1",
    y = "v5.0q",
    color = "Genomic Context",
    shape = "Genomic Context"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### By Variant Size

```{r smvar_size_plot, fig.width = 7, fig.height = 5}
smvar_size_long <- smvar_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(metric = str_to_title(metric))

(smvar_size_plt <- smvar_size_long %>%
  ggplot(aes(x = size_bin, y = value, fill = bench_version)) +
  geom_col(
    position = position_dodge(width = 0.8), width = 0.7,
    color = "grey30", linewidth = 0.3
  ) +
  facet_wrap(~metric) +
  scale_y_continuous(
    trans   = phred_trans,
    breaks  = phred_smvar_pcts,
    labels  = phred_smvar_labels,
    limits  = c(0.85, 0.9999),
    oob     = scales::squish
  ) +
  scale_benchmark_version(aesthetic = "fill") +
  labs(x = "Variant Size", y = NULL, fill = "Benchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### Combined Small Variant Figure

```{r combined_smvar, fig.width = 10, fig.height = 16}
(combined_smvar_plt <-
  (smvar_scatter_plt + theme(legend.position = "bottom")) /
  smvar_type_plt /
  smvar_size_plt +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(5, 5, 5, 5), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_smvar.png"),
  plot = combined_smvar_plt,
  width = 10,
  height = 16,
  dpi = 300
)
```

### Figure Legend

**Figure use_case_smvar: Small variant benchmarking of the Roche NeuSomatic callset against v4.2.1 and v5.0q benchmark sets.**
(A) Benchmark comparison scatter plots for the Roche NeuSomatic HG002 callset (Illumina + PacBio + ONT input data) evaluated against the v4.2.1 and v5.0q GRCh38 small variant benchmarks using `rtg vcfeval`. Each panel shows one performance metric (recall, precision, F1); each point is a genomic context (overall, homopolymers, low mappability, segmental duplications, tandem repeats) using GIAB v3.6 stratifications. Axes are phred-scaled; the dashed diagonal marks equal performance between the two benchmarks.
(B) Same comparison faceted by variant type (SNV and INDEL) and performance metric.
(C) Recall and precision by variant size bin (SNP, 1–4 bp, 5–14 bp, 15–49 bp) on a phred-scaled y-axis.

### Results Text

To demonstrate the practical impact of the expanded v5.0q benchmark regions on variant caller evaluation, we benchmarked the Roche NeuSomatic HG002 callset from the precisionFDA Truth Challenge V2 against both the v4.2.1 and v5.0q GRCh38 small variant benchmark sets using `rtg vcfeval` with GIAB v3.6 stratifications.

Against the v4.2.1 benchmark, the NeuSomatic callset achieved near-perfect performance overall (F1 = 0.9997, recall = 99.96%, precision = 99.98%), with only 612 false positives and 1,685 false negatives across 3.89 million true positives. Performance was consistently high across all genomic contexts, with recall and precision exceeding 99.6% even in segmental duplications and low mappability regions. This high performance reflects the limited scope of the v4.2.1 benchmark, which largely excluded the most challenging repetitive regions.

Against the v5.0q benchmark, the same callset showed substantially lower performance (F1 = 0.9827, recall = 97.99%, precision = 98.55%), with 64,928 false positives and 90,686 false negatives across 4.43 million true positives. The decrease was most pronounced in homopolymers (recall = 92.2%, precision = 94.5%) and tandem repeats (recall = 91.1%, precision = 92.9%), followed by segmental duplications (recall = 91.6%, precision = 95.1%) and low mappability regions (recall = 94.6%, precision = 96.7%). These results demonstrate that the expanded v5.0q benchmark regions reveal variant calling errors in difficult genomic contexts that were not captured by the v4.2.1 benchmark.

Performance differences were particularly pronounced for INDELs compared to SNVs. Against v5.0q, INDEL recall dropped to 93.3% in homopolymers and 90.4% in tandem repeats, compared to >99.3% for both variant types against v4.2.1. By variant size, the largest performance decrease was observed for INDELs in the 5–14 bp and 15–49 bp size ranges, where the v5.0q benchmark includes substantially more variants in repetitive regions.


## Structural Variant Benchmarking: Baylor ONT vs Illumina DRAGEN

Comparison of Baylor ONT (Sniffles2) and Illumina DRAGEN SV callsets
against the v5.0q GRCh38 structural variant benchmark.
Evaluated with `truvari bench` and `truvari refine` using GIAB v5.0q recommended parameters (`--pick ac --passonly -r 2000 -C 5000 --refine`).

```{r stvar_data}
stvar_df <- read_csv(
  here("results", "use_case", "stvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV type-level metrics
svtype_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV size-binned metrics
sv_size_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_size_counts.csv"),
  show_col_types = FALSE
) %>%
  filter(svtype %in% c("INS", "DEL")) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  mutate(
    size_bin = factor(size_bin, levels = sv_size_levels),
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    )
  )
```

### Callset Comparison Scatter: ONT vs DRAGEN by Context

```{r stvar_scatter, fig.width = 10, fig.height = 4}
stvar_compare_df <- stvar_df |>
  select(callset, context_label, precision, recall, f1) |>
  pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = callset, values_from = value) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

(stvar_scatter_plt <- stvar_compare_df |>
  ggplot(aes(x = baylor_ont, y = dragen, color = context_label, shape = context_label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(size = 3.5, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = context_label),
    size = 2.5, max.overlaps = 15, seed = 42,
    show.legend = FALSE
  ) +
  facet_wrap(~metric) +
  scale_x_continuous(
    trans = phred_trans,
    breaks = phred_stvar_pcts, labels = phred_stvar_labels
  ) +
  scale_y_continuous(
    trans = phred_trans,
    breaks = phred_stvar_pcts, labels = phred_stvar_labels
  ) +
  annotation_phredticks(sides = "bl") +
  scale_color_manual(values = context_label_colors) +
  scale_shape_manual(values = context_label_shapes) +
  labs(
    x = "Baylor ONT (Sniffles2)",
    y = "Illumina DRAGEN",
    color = "Genomic Context",
    shape = "Genomic Context"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### By SV Type and Size

```{r stvar_size_plot, fig.width = 8, fig.height = 5}
sv_size_long <- sv_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(metric = str_to_title(metric))

(stvar_size_plt <- sv_size_long %>%
  ggplot(aes(x = size_bin, y = value, fill = callset_label)) +
  geom_col(
    position = position_dodge(width = 0.8), width = 0.7,
    color = "grey30", linewidth = 0.3
  ) +
  facet_grid(svtype ~ metric) +
  scale_y_continuous(
    trans   = phred_trans,
    breaks  = phred_stvar_pcts,
    labels  = phred_stvar_labels,
    limits  = c(0, 0.99)
  ) +
  scale_fill_manual(values = callset_colors) +
  labs(x = "SV Size", y = NULL, fill = "Callset") +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -30, hjust = 0),
    legend.position = "bottom"
  ))
```

### Callset Comparison Scatter: By SV Type and Size

```{r stvar_size_scatter, fig.width = 10, fig.height = 7}
sv_size_compare_df <- sv_size_df |>
  select(callset, svtype, size_bin, precision, recall, f1) |>
  pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") |>
  pivot_wider(names_from = callset, values_from = value) |>
  mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

(stvar_size_scatter_plt <- sv_size_compare_df |>
  ggplot(aes(x = baylor_ont, y = dragen, color = svtype, shape = size_bin)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(size = 3, stroke = 0.8) +
  ggrepel::geom_text_repel(
    aes(label = size_bin),
    size = 2, max.overlaps = 20, seed = 42,
    show.legend = FALSE
  ) +
  facet_grid(svtype ~ metric) +
  scale_x_continuous(
    trans = phred_trans,
    breaks = phred_stvar_pcts, labels = phred_stvar_labels
  ) +
  scale_y_continuous(
    trans = phred_trans,
    breaks = phred_stvar_pcts, labels = phred_stvar_labels
  ) +
  annotation_phredticks(sides = "bl") +
  scale_color_manual(values = c("DEL" = "#D95F02", "INS" = "#1B9E77")) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8)) +
  labs(
    x = "Baylor ONT (Sniffles2)",
    y = "Illumina DRAGEN",
    color = "SV Type",
    shape = "SV Size"
  ) +
  coord_equal(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "right"))
```

### Combined SV Figure

```{r combined_stvar, fig.width = 10, fig.height = 16}
(combined_stvar_plt <-
  (stvar_scatter_plt + theme(legend.position = "bottom")) /
  stvar_size_plt /
  (stvar_size_scatter_plt + theme(legend.position = "right")) +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(5, 5, 5, 5), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_stvar.png"),
  plot = combined_stvar_plt,
  width = 10,
  height = 16,
  dpi = 300
)
```

### Figure Legend

**Figure use_case_stvar: Structural variant benchmarking of Baylor ONT (Sniffles2) and Illumina DRAGEN callsets against the v5.0q GRCh38 SV benchmark.**
(A) Callset comparison scatter plots for overall and stratified performance (homopolymers, low mappability, segmental duplications, tandem repeats) using GIAB v3.6 stratifications and `truvari bench` with GIAB-recommended parameters (`--pick ac --passonly -r 2000 -C 5000`) followed by `truvari refine`. Each panel shows one performance metric (recall, precision, F1); x-axis is Baylor ONT (Sniffles2), y-axis is Illumina DRAGEN. Axes are phred-scaled; the dashed diagonal marks equal performance between callsets.
(B) Recall and precision by SV type (DEL and INS) and size bin (50–99 bp, 100–299 bp, 300–999 bp, 1–10 kb, ≥10 kb) on a phred-scaled y-axis.
(C) Callset comparison scatter by SV type and size bin, faceted by SV type and performance metric. Axes are phred-scaled.

### Results Text

To illustrate how the v5.0q SV benchmark enables differentiation of variant calling approaches, we compared the Baylor ONT (Sniffles2) and Illumina DRAGEN SV callsets against the v5.0q GRCh38 structural variant benchmark using `truvari bench` with GIAB-recommended parameters (`--pick ac --passonly -r 2000 -C 5000`) followed by `truvari refine` for MSA-based re-evaluation of candidate regions. ALT="*" spanning deletion records were filtered prior to benchmarking per GIAB v5.0q README guidance.

The Baylor ONT callset achieved higher overall recall (74.5% vs. 64.2%) and precision (93.6% vs. 78.6%) compared to DRAGEN. This performance advantage was consistent across all genomic contexts. Both callsets showed highest recall in homopolymers (ONT: 87.6%, DRAGEN: 77.9%) and lowest recall in segmental duplications (ONT: 45.0%, DRAGEN: 35.4%), where sequence similarity between duplicated loci presents challenges for both short and long read alignment.

The precision advantage of ONT over DRAGEN was most pronounced in tandem repeats (92.8% vs. 77.9%) and segmental duplications (80.4% vs. 65.9%). DRAGEN produced substantially more false positives overall (4,909 vs. 1,431), particularly in tandem repeats (3,334 vs. 1,106), likely reflecting the inherent difficulty of resolving repeat-length variation from short reads.

By SV type, both callsets had higher recall for deletions than insertions (ONT: DEL 77.3% vs. INS 73.0%; DRAGEN: DEL 64.8% vs. INS 63.9%). Precision was similar for both SV types in the ONT callset (~93–94%), while DRAGEN showed higher precision for deletions (88.4%) than insertions (85.8%). Across size bins, the ONT callset maintained relatively consistent recall for deletions from 50 bp to 10 kb, while DRAGEN recall dropped more steeply for larger deletions. For insertions, both callsets showed decreased recall for variants ≥10 kb, though the ONT callset retained a clear advantage at all size ranges.

These results demonstrate that the expanded v5.0q SV benchmark, which includes challenging repetitive regions, reveals meaningful performance differences between sequencing technologies and variant callers that would not be apparent with a benchmark limited to easier genomic contexts.
