---
title: "Benchmarking Use-Case: v4.2.1 vs v5.0q"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)

source(here::here("scripts/happy_giab.R"))  ## imports convert_to_phred()

size_bin_levels <- c("SNP", "1-5bp", "6-15bp", "16+bp")
sv_size_levels <- c("[50,100)", "[100,300)", "[300,1000)", "[1000,10000)", ">=10000")

## Phred-scale ggplot2 transformation ----------------------------------------
phred_trans <- scales::trans_new(
  name      = "phred",
  transform = convert_to_phred,
  inverse   = function(y) 1 - 10^(-y / 10)
)

## Break positions and labels (in proportion scale) for each plot type
phred_smvar_pcts   <- c(0.90, 0.95, 0.99, 0.999)
phred_smvar_labels <- c("90%", "95%", "99%", "99.9%")

phred_stvar_pcts   <- c(0.5, 0.7, 0.8, 0.9, 0.95)
phred_stvar_labels <- c("50%", "70%", "80%", "90%", "95%")

## Colour and shape mappings for context labels in comparison scatter plots
context_label_colors <- c(
  "Overall"                = "black",
  "Homopolymers"           = "#E41A1C",
  "Low Mappability"        = "#377EB8",
  "Segmental Duplications" = "#4DAF4A",
  "Tandem Repeats"         = "#FF7F00"
)
context_label_shapes <- c(
  "Overall"                = 21,
  "Homopolymers"           = 24,
  "Low Mappability"        = 22,
  "Segmental Duplications" = 23,
  "Tandem Repeats"         = 25
)

## annotation_phredticks and helpers ------------------------------------------
## Adapted from:
## https://github.com/usnistgov/giab-pFDA-2nd-challenge/blob/main/scripts/annotation-phredticks.R
##
## Calculates phred-scale tick positions for annotation_phredticks().
calc_phredticks <- function(
  minphred = 0, maxphred = minphred + 1,
  start = 0, shortend = 0.1, midend = 0.2, longend = 0.3
) {
  tick_df <- tibble(
    x_pct    = c(0, seq(10, 90, 10),
                 seq(91, 99, 1),
                 seq(99.1,   99.9,   0.1),
                 seq(99.91,  99.99,  0.01),
                 seq(99.991, 99.999, 0.001)),
    ticknums = c(0, rep(1:9, times = 5))
  ) |>
    mutate(
      x       = x_pct / 100,
      phred_x = convert_to_phred(x),
      start   = start,
      tickend = case_when(
        x == 0        ~ longend,
        ticknums == 9 ~ longend,
        ticknums == 5 ~ midend,
        TRUE          ~ shortend
      )
    )
  data.frame(value = tick_df$phred_x, start = tick_df$start, end = tick_df$tickend)
}

GeomPhredticks <- ggplot2::ggproto(
  "GeomPhredticks", ggplot2::Geom,
  extra_params = "",
  handle_na = function(data, params) data,

  draw_panel = function(
    data, panel_params, coord,
    base = 10, sides = "bl", outside = FALSE, scaled = TRUE,
    short    = grid::unit(0.1, "cm"),
    mid      = grid::unit(0.2, "cm"),
    long     = grid::unit(0.3, "cm"),
    colour   = "black", size = 0.5, linetype = 1, alpha = 1
  ) {
    ticks <- list()
    short <- grid::convertUnit(short, "cm", valueOnly = TRUE)
    mid   <- grid::convertUnit(mid,   "cm", valueOnly = TRUE)
    long  <- grid::convertUnit(long,  "cm", valueOnly = TRUE)
    flipped <- inherits(coord, "CoordFlip")
    x_name  <- if (flipped) "y" else "x"
    y_name  <- if (flipped) "x" else "y"

    if (grepl("[bt]", sides)) {
      xticks <- calc_phredticks(
        minphred = floor(panel_params$x.range[1]),
        maxphred = ceiling(panel_params$x.range[2]),
        start = 0, shortend = short, midend = mid, longend = long
      )
      names(xticks)[names(xticks) == "value"] <- x_name
      xticks <- coord$transform(xticks, panel_params)
      xticks <- xticks[xticks$x >= 0 & xticks$x <= 1, ]
      if (outside) xticks$end <- -xticks$end

      if (grepl("b", sides))
        ticks$x_b <- grid::segmentsGrob(
          x0 = grid::unit(xticks$x, "native"), x1 = grid::unit(xticks$x, "native"),
          y0 = grid::unit(xticks$start, "cm"), y1 = grid::unit(xticks$end, "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
      if (grepl("t", sides))
        ticks$x_t <- grid::segmentsGrob(
          x0 = grid::unit(xticks$x, "native"), x1 = grid::unit(xticks$x, "native"),
          y0 = grid::unit(1, "npc") - grid::unit(xticks$start, "cm"),
          y1 = grid::unit(1, "npc") - grid::unit(xticks$end,   "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
    }

    if (grepl("[lr]", sides)) {
      yticks <- calc_phredticks(
        minphred = floor(panel_params$y.range[1]),
        maxphred = ceiling(panel_params$y.range[2]),
        start = 0, shortend = short, midend = mid, longend = long
      )
      names(yticks)[names(yticks) == "value"] <- y_name
      yticks <- coord$transform(yticks, panel_params)
      yticks <- yticks[yticks$y >= 0 & yticks$y <= 1, ]
      if (outside) yticks$end <- -yticks$end

      if (grepl("l", sides))
        ticks$y_l <- grid::segmentsGrob(
          y0 = grid::unit(yticks$y, "native"), y1 = grid::unit(yticks$y, "native"),
          x0 = grid::unit(yticks$start, "cm"), x1 = grid::unit(yticks$end, "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
      if (grepl("r", sides))
        ticks$y_r <- grid::segmentsGrob(
          y0 = grid::unit(yticks$y, "native"), y1 = grid::unit(yticks$y, "native"),
          x0 = grid::unit(1, "npc") - grid::unit(yticks$start, "cm"),
          x1 = grid::unit(1, "npc") - grid::unit(yticks$end,   "cm"),
          gp = grid::gpar(col = scales::alpha(colour, alpha),
                          lty = linetype, lwd = size * ggplot2::.pt)
        )
    }

    grid::gTree(children = do.call(grid::gList, ticks))
  },

  default_aes = ggplot2::aes(colour = "black", size = 0.5, linetype = 1, alpha = 1)
)

annotation_phredticks <- function(
  base = 10, sides = "bl", outside = FALSE, scaled = TRUE,
  short    = grid::unit(0.1, "cm"),
  mid      = grid::unit(0.2, "cm"),
  long     = grid::unit(0.3, "cm"),
  colour = "black", size = 0.5, linetype = 1, alpha = 1, color = NULL, ...
) {
  if (!is.null(color)) colour <- color
  ggplot2::layer(
    data        = data.frame(x = 1),
    mapping     = NULL,
    stat        = ggplot2::StatIdentity,
    geom        = GeomPhredticks,
    position    = ggplot2::PositionIdentity,
    show.legend = FALSE,
    inherit.aes = FALSE,
    params = list(
      base     = base,     sides    = sides,    outside  = outside,
      scaled   = scaled,   short    = short,    mid      = mid,
      long     = long,     colour   = colour,   size     = size,
      linetype = linetype, alpha    = alpha,
      ...
    )
  )
}
```

## Small Variant Benchmarking: Roche NeuSomatic

Comparison of Roche NeuSomatic (pFDA Truth Challenge V2) callset against
v4.2.1 and v5.0q small variant benchmarks on GRCh38, evaluated with hap.py.

```{r smvar_data}
## Context mapping: v4 and v5 use different stratification BED names
v4_context_map <- c(
  "HP"  = "GRCh38_AllHomopolymers_gt6bp_imperfectgt10bp_slop5.bed.gz",
  "TR"  = "GRCh38_AllTandemRepeatsandHomopolymers_slop5.bed.gz",
  "SD"  = "GRCh38_segdups.bed.gz",
  "MAP" = "GRCh38_lowmappabilityall.bed.gz"
)

v5_context_map <- c(
  "HP"  = "AllHomopolymers_ge7bp_imperfectge11bp_slop5",
  "TR"  = "AllTandemRepeatsandHomopolymers_slop5",
  "SD"  = "segdups",
  "MAP" = "lowmappabilityall"
)

## Read hap.py extended CSVs
v4_happy <- read_happy_extended(here("data", "IA789_HG002.extended.csv"))
v5_happy <- read_happy_extended(
  here(
    "data",
    "GRCh38_HG002_T~T2TQ100v1.1_Q~EEroche_TR~none_GRCh38_HG002-T2TQ100v1.1-dipz2k_smvar-excluded",
    "GRCh38_HG2Q100EE-roche-neusomatic_HG2-T2TQ100-V1.1_smvar_dipcall-z2k.extended.csv"
  )
)

## Helper to extract metrics from a happy df for given subsets
extract_happy_metrics <- function(happy_df, subset_filter, bench_version) {
  happy_df %>%
    filter(Filter == "PASS", Subtype == "*", Subset %in% subset_filter) %>%
    select(
      Type, Subset,
      METRIC.Recall, METRIC.Precision, METRIC.F1_Score,
      TRUTH.TP, TRUTH.FN, QUERY.FP
    ) %>%
    mutate(bench_version = bench_version)
}

## Overall + per-type metrics
smvar_overall_df <- bind_rows(
  extract_happy_metrics(v4_happy, "*", "v4.2.1"),
  extract_happy_metrics(v5_happy, "*", "v5.0q")
) %>%
  filter(Type %in% c("SNP", "INDEL")) %>%
  mutate(
    context = case_when(
      Type == "SNP" ~ "SNV",
      Type == "INDEL" ~ "INDEL"
    )
  )

## Per-context metrics
v4_context_df <- extract_happy_metrics(v4_happy, v4_context_map, "v4.2.1") %>%
  mutate(context = names(v4_context_map)[match(Subset, v4_context_map)])
v5_context_df <- extract_happy_metrics(v5_happy, v5_context_map, "v5.0q") %>%
  mutate(context = names(v5_context_map)[match(Subset, v5_context_map)])

## Combine overall + context with labels
smvar_all_df <- bind_rows(
  smvar_overall_df %>%
    select(bench_version, context, Type,
           recall = METRIC.Recall, precision = METRIC.Precision,
           f1 = METRIC.F1_Score, tp = TRUTH.TP, fn = TRUTH.FN, fp = QUERY.FP),
  bind_rows(v4_context_df, v5_context_df) %>%
    filter(Type == "INDEL") %>%
    mutate(context_label = get_context_labels()[context]) %>%
    select(bench_version, context, Type,
           recall = METRIC.Recall, precision = METRIC.Precision,
           f1 = METRIC.F1_Score, tp = TRUTH.TP, fn = TRUTH.FN, fp = QUERY.FP)
) %>%
  mutate(
    context_label = case_when(
      context %in% c("SNV", "INDEL") ~ context,
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    )
  )

## Also get overall "all variants" row
smvar_total_df <- bind_rows(
  v4_happy %>%
    filter(Filter == "PASS", Subtype == "*", Subset == "*", Type == "INDEL") %>%
    mutate(bench_version = "v4.2.1"),
  v5_happy %>%
    filter(Filter == "PASS", Subtype == "*", Subset == "*", Type == "INDEL") %>%
    mutate(bench_version = "v5.0q")
)
```

### Benchmark Comparison: v4.2.1 vs v5.0q

```{r smvar_scatter, fig.width = 10, fig.height = 7}
## Combine per-type + per-context for scatter
scatter_labels <- c("SNV", "INDEL", "Homopolymers", "Low Mappability",
                    "Segmental Duplications", "Tandem Repeats")

## For per-type: use both SNP and INDEL rows
## For per-context: use INDEL rows (Type == "INDEL")
smvar_scatter_df <- smvar_all_df %>%
  filter(!is.na(context_label), context_label %in% scatter_labels) %>%
  select(bench_version, context_label, recall, precision, f1) %>%
  pivot_longer(c(recall, precision, f1), names_to = "metric", values_to = "value") %>%
  pivot_wider(names_from = bench_version, values_from = value) %>%
  rename(v421 = `v4.2.1`, v50q = `v5.0q`) %>%
  mutate(
    metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")),
    context_label = factor(context_label, levels = scatter_labels)
  )

## Shape and color maps
scatter_colors <- c(
  "SNV"                    = "#1f77b4",
  "INDEL"                  = "#ff7f0e",
  "Homopolymers"           = "#E41A1C",
  "Low Mappability"        = "#377EB8",
  "Segmental Duplications" = "#4DAF4A",
  "Tandem Repeats"         = "#FF7F00"
)
scatter_shapes <- c(
  "SNV"                    = 24,
  "INDEL"                  = 22,
  "Homopolymers"           = 24,
  "Low Mappability"        = 22,
  "Segmental Duplications" = 23,
  "Tandem Repeats"         = 25
)

(smvar_scatter_plt <- smvar_scatter_df %>%
  ggplot(aes(x = v421, y = v50q, fill = context_label, shape = context_label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_point(color = "grey30", size = 3.5, stroke = 0.8) +
  facet_wrap(~metric) +
  scale_x_continuous(
    trans   = phred_trans,
    breaks  = phred_smvar_pcts,
    labels  = phred_smvar_labels,
    limits  = c(0.90, 0.9999),
    oob     = scales::squish
  ) +
  scale_y_continuous(
    trans   = phred_trans,
    breaks  = phred_smvar_pcts,
    labels  = phred_smvar_labels,
    limits  = c(0.90, 0.9999),
    oob     = scales::squish
  ) +
  scale_fill_manual(values = scatter_colors) +
  scale_shape_manual(values = scatter_shapes) +
  labs(
    x = "v4.2.1",
    y = "v5.0q",
    fill = NULL,
    shape = NULL
  ) +
  coord_cartesian(clip = "off") +
  theme_manuscript() +
  theme(legend.position = "bottom", aspect.ratio = 1))
```

### By Variant Size

```{r smvar_size_data}
## Extract size bin metrics from hap.py Subtype column
## Subtype values: I1_5/D1_5/C1_5 → "1-5bp", I6_15/D6_15/C6_15 → "6-15bp",
## I16_PLUS/D16_PLUS/C16_PLUS → "16+bp"
subtype_to_sizebin <- c(
  "I1_5" = "1-5bp", "D1_5" = "1-5bp", "C1_5" = "1-5bp",
  "I6_15" = "6-15bp", "D6_15" = "6-15bp", "C6_15" = "6-15bp",
  "I16_PLUS" = "16+bp", "D16_PLUS" = "16+bp", "C16_PLUS" = "16+bp"
)

extract_size_metrics <- function(happy_df, bench_version) {
  ## INDEL size bins
  indel_size <- happy_df %>%
    filter(Filter == "PASS", Subset == "*", Type == "INDEL",
           Subtype %in% names(subtype_to_sizebin)) %>%
    mutate(size_bin = subtype_to_sizebin[Subtype]) %>%
    group_by(size_bin) %>%
    summarise(
      tp = sum(TRUTH.TP), fn = sum(TRUTH.FN), fp = sum(QUERY.FP),
      .groups = "drop"
    ) %>%
    mutate(
      recall = tp / (tp + fn),
      precision = tp / (tp + fp)
    )

  ## SNP row
  snp_row <- happy_df %>%
    filter(Filter == "PASS", Subset == "*", Type == "SNP", Subtype == "*") %>%
    transmute(
      size_bin = "SNP",
      tp = TRUTH.TP, fn = TRUTH.FN, fp = QUERY.FP,
      recall = METRIC.Recall, precision = METRIC.Precision
    )

  bind_rows(snp_row, indel_size) %>%
    mutate(bench_version = bench_version)
}

smvar_size_df <- bind_rows(
  extract_size_metrics(v4_happy, "v4.2.1"),
  extract_size_metrics(v5_happy, "v5.0q")
) %>%
  mutate(size_bin = factor(size_bin, levels = size_bin_levels))
```

```{r smvar_size_plot, fig.width = 7, fig.height = 5}
smvar_size_long <- smvar_size_df %>%
  pivot_longer(cols = c(recall, precision), names_to = "metric", values_to = "value") %>%
  mutate(
    metric = str_to_title(metric),
    size_bin_num = as.numeric(size_bin)
  )

(smvar_size_plt <- smvar_size_long %>%
  ggplot(aes(x = size_bin_num, y = value, color = bench_version, group = bench_version)) +
  geom_line() +
  geom_point(aes(fill = bench_version), shape = 21, color = "grey30", size = 2.5) +
  facet_wrap(~metric) +
  scale_x_continuous(
    breaks = seq_along(size_bin_levels),
    labels = size_bin_levels
  ) +
  scale_y_continuous(
    trans   = phred_trans,
    breaks  = phred_smvar_pcts,
    labels  = phred_smvar_labels,
    limits  = c(0.85, 0.9999),
    oob     = scales::squish
  ) +
  scale_benchmark_version(aesthetic = "color", guide = "none") +
  scale_benchmark_version(aesthetic = "fill") +
  labs(x = "Variant Size", y = NULL, fill = "Benchmark") +
  theme_manuscript() +
  theme(legend.position = "bottom"))
```

### Combined Small Variant Figure

```{r combined_smvar, fig.width = 8, fig.height = 10}
design <- c(area(l = 1, r = 8, t = 1, b = 8), area(l = 1, r = 8, t = 8, b = 10))
(combined_smvar_plt <-
  (smvar_scatter_plt) /
  (smvar_size_plt) &
  plot_annotation(tag_levels = "A") &
  plot_layout(design = design)
 )


ggsave(
  filename = here("manuscript", "figs", "use_case_smvar.png"),
  plot = combined_smvar_plt,
  width = 8,
  height = 10,
  dpi = 300
)
```

### Figure Legend

**Figure use_case_smvar: Small variant benchmarking of the Roche NeuSomatic callset against v4.2.1 and v5.0q benchmark sets.**
(A) Benchmark comparison scatter plots for the Roche NeuSomatic HG002 callset (Illumina + PacBio + ONT input data) evaluated against the v4.2.1 and v5.0q GRCh38 small variant benchmarks using hap.py. Each panel shows one performance metric (recall, precision, F1); points represent variant types (SNV, INDEL) and per-context INDEL performance in genomic contexts (homopolymers, low mappability, segmental duplications, tandem repeats). Note that v4.2.1 and v5.0q use slightly different BED definitions for context stratifications. Axes are phred-scaled; the dashed diagonal marks equal performance between the two benchmarks.
(B) Recall and precision by variant size bin (SNP, 1–5 bp, 6–15 bp, 16+ bp) on a phred-scaled y-axis.

### Results Text

To demonstrate the practical impact of the expanded v5.0q benchmark regions on variant caller evaluation, we benchmarked the Roche NeuSomatic HG002 callset from the precisionFDA Truth Challenge V2 against both the v4.2.1 and v5.0q GRCh38 small variant benchmark sets using hap.py with GIAB stratifications.

Against the v4.2.1 benchmark, the NeuSomatic callset achieved near-perfect performance for both SNVs (F1 = 0.9998, recall = 99.97%, precision = 100.00%) and INDELs (F1 = 0.9989, recall = 99.87%, precision = 99.90%), with only 116 SNV and 549 INDEL false positives across 3.36 million SNV and 525K INDEL true positives. Per-context INDEL performance was similarly high, with recall exceeding 99.5% in all four genomic contexts (HP: 99.85%, TR: 99.83%, SD: 99.60%, MAP: 99.51%).

Against the v5.0q benchmark, the same callset showed substantially lower performance, reflecting the inclusion of more difficult genomic regions. SNV performance remained high but decreased (F1 = 0.9942, recall = 99.24%, precision = 99.61%), while INDEL performance dropped more substantially (F1 = 0.9456, recall = 94.16%, precision = 94.95%) with 50,522 false negatives and 46,678 false positives across 815K INDEL true positives. Per-context INDEL performance revealed the source of these differences: homopolymers (recall = 92.24%), tandem repeats (recall = 92.54%), segmental duplications (recall = 84.67%), and low mappability regions (recall = 81.42%) all showed markedly lower recall than against v4.2.1. By variant size, the largest performance decreases were observed for larger INDELs: 6–15 bp recall dropped from 99.7% (v4.2.1) to 87.3% (v5.0q), and 16+ bp recall dropped from 99.7% to 79.8%.


## Structural Variant Benchmarking: Sniffles1 vs Sniffles2

Comparison of Sniffles1 and Sniffles2 SV callsets from HiFi and ONT platforms
against the v5.0q GRCh38 structural variant benchmark.
Evaluated with `truvari bench` and `truvari refine` using GIAB v5.0q recommended parameters (`--pick ac --passonly -r 2000 -C 5000 --refine`).

```{r stvar_data}
## Helper to compute recall/precision from TP/FP/FN counts
compute_metrics <- function(df) {
  df %>%
    mutate(
      recall = tp / (tp + fn),
      precision = tp / (tp + fp),
      f1 = 2 * (precision * recall) / (precision + recall)
    )
}

stvar_df <- read_csv(
  here("results", "use_case", "stvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    callset_label = str_to_title(callset),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV type-level metrics
svtype_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    callset_label = str_to_title(callset),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

## SV size-binned metrics
sv_size_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_size_counts.csv"),
  show_col_types = FALSE
) %>%
  filter(svtype %in% c("INS", "DEL")) %>%
  pivot_wider(names_from = category, values_from = count, values_fill = 0L) %>%
  rename(tp = TP, fp = FP, fn = FN) %>%
  compute_metrics() %>%
  mutate(
    size_bin = factor(size_bin, levels = sv_size_levels),
    callset_label = str_to_title(callset)
  )

## Callset colors
callset_colors <- c("Sniffles1" = "#D95F02", "Sniffles2" = "#7570B3")
```

### ONT: Sniffles1 vs Sniffles2

```{r stvar_ont_scatter, fig.width = 10, fig.height = 4}
make_stvar_scatter <- function(df, platform_filter) {
  compare_df <- df %>%
    filter(platform == platform_filter) %>%
    select(callset, context_label, precision, recall, f1) %>%
    pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") %>%
    pivot_wider(names_from = callset, values_from = value) %>%
    mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

  xy_range <- range(c(compare_df$sniffles1, compare_df$sniffles2), na.rm = TRUE)

  compare_df %>%
    ggplot(aes(x = sniffles1, y = sniffles2, fill = context_label, shape = context_label)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
    geom_point(color = "grey30", size = 3.5, stroke = 0.8) +
    facet_wrap(~metric) +
    scale_x_continuous(
      trans = phred_trans,
      breaks = phred_stvar_pcts, labels = phred_stvar_labels,
      limits = xy_range
    ) +
    scale_y_continuous(
      trans = phred_trans,
      breaks = phred_stvar_pcts, labels = phred_stvar_labels,
      limits = xy_range
    ) +
    annotation_phredticks(sides = "bl") +
    scale_fill_manual(values = context_label_colors) +
    scale_shape_manual(values = context_label_shapes) +
    labs(
      x = "Sniffles1",
      y = "Sniffles2",
      fill = "Genomic Context",
      shape = "Genomic Context"
    ) +
    coord_cartesian(clip = "off") +
    theme_manuscript() +
    theme(legend.position = "right", aspect.ratio = 1)
}

(ont_scatter_plt <- make_stvar_scatter(stvar_df, "ONT"))
```

```{r stvar_ont_size_scatter, fig.width = 10, fig.height = 7}
make_stvar_size_scatter <- function(size_df, platform_filter) {
  compare_df <- size_df %>%
    filter(platform == platform_filter) %>%
    select(callset, svtype, size_bin, precision, recall, f1) %>%
    pivot_longer(c(precision, recall, f1), names_to = "metric", values_to = "value") %>%
    pivot_wider(names_from = callset, values_from = value) %>%
    mutate(metric = factor(str_to_title(metric), levels = c("Recall", "Precision", "F1")))

  compare_df %>%
    ggplot(aes(x = sniffles1, y = sniffles2, fill = svtype, size = size_bin)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
    geom_point(shape = 21, color = "grey30", stroke = 0.6) +
    facet_wrap(~metric) +
    scale_fill_manual(values = c("DEL" = "#D95F02", "INS" = "#1B9E77")) +
    labs(
      x = "Sniffles1",
      y = "Sniffles2",
      fill = "SV Type",
      size = "SV Size"
    ) +
    coord_equal(clip = "off") +
    theme_manuscript() +
    theme(legend.position = "right")
}

(ont_size_scatter_plt <- make_stvar_size_scatter(sv_size_df, "ONT"))
```

```{r combined_stvar_ont, fig.width = 8, fig.height = 8}
(combined_ont_plt <-
  (ont_scatter_plt + theme(legend.position = "bottom")) /
  (ont_size_scatter_plt + theme(legend.position = "bottom")) +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(1, 1, 1, 1), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_stvar_ont.png"),
  plot = combined_ont_plt,
  width = 8,
  height = 8,
  dpi = 300
)
```

### HiFi: Sniffles1 vs Sniffles2

```{r stvar_hifi, fig.width = 8, fig.height = 8}
hifi_scatter_plt <- make_stvar_scatter(stvar_df, "HiFi")
hifi_size_scatter_plt <- make_stvar_size_scatter(sv_size_df, "HiFi")

(combined_hifi_plt <-
  (hifi_scatter_plt + theme(legend.position = "bottom")) /
  (hifi_size_scatter_plt + theme(legend.position = "bottom")) +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(1, 1, 1, 1), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_stvar_hifi.png"),
  plot = combined_hifi_plt,
  width = 8,
  height = 8,
  dpi = 300
)
```

### Figure Legend

**Figure use_case_stvar_ont: ONT structural variant benchmarking comparing Sniffles1 and Sniffles2 callsets against the v5.0q GRCh38 SV benchmark.**
(A) Callset comparison scatter plots for overall and stratified performance (homopolymers, low mappability, segmental duplications, tandem repeats) using `truvari bench` with GIAB-recommended parameters (`--pick ac --passonly -r 2000 -C 5000`) followed by `truvari refine`. Each panel shows one performance metric (recall, precision, F1); x-axis is Sniffles1, y-axis is Sniffles2. Axes are phred-scaled; the dashed diagonal marks equal performance between callsets.
(B) Callset comparison scatter by SV type (DEL, INS) and size bin, faceted by performance metric.

**Figure use_case_stvar_hifi: HiFi structural variant benchmarking comparing Sniffles1 and Sniffles2 callsets against the v5.0q GRCh38 SV benchmark.** Same layout and parameters as the ONT figure.

### Results Text

To demonstrate the utility of the v5.0q SV benchmark for evaluating caller improvements, we compared Sniffles1 and Sniffles2 SV callsets from both ONT and HiFi platforms against the v5.0q GRCh38 benchmark using `truvari bench` with GIAB-recommended parameters (`--pick ac --passonly -r 2000 -C 5000`) followed by `truvari refine` for MSA-based re-evaluation.

For ONT data, Sniffles2 improved overall recall from 85.98% (Sniffles1) to 91.80% while maintaining comparable precision (99.34% vs 98.86%). Recall improvements were consistent across genomic contexts, with the largest absolute gains in tandem repeats (55.10% to 61.88%) and low mappability regions (46.44% to 52.27%). By SV type, Sniffles2 improved both deletion recall (65.99% to 67.88%) and insertion recall (59.57% to 70.13%), with insertions showing the larger gain.

For HiFi data, the improvement was even more pronounced. Sniffles2 increased overall recall from 71.39% to 91.36%, a 20 percentage-point gain, while precision remained high (98.29% vs 98.94%). Deletion recall improved from 48.76% to 68.96%, and insertion recall from 55.27% to 71.26%. Per-context recall gains were substantial: tandem repeats (45.62% to 63.80%), segmental duplications (24.19% to 38.69%), and low mappability (39.58% to 56.06%). The larger gains on HiFi suggest that Sniffles1 was less optimized for HiFi-specific characteristics (e.g., longer reads, different error profiles) compared to ONT.

Precision for per-context metrics was generally lower for Sniffles2 than Sniffles1, particularly in segmental duplications (HiFi: 87.77% to 72.96%; ONT: 82.70% to 75.41%), suggesting a recall-precision tradeoff in these challenging regions.


## Metrics Cross-Reference

Numbers used in the results text above, for verification against manuscript.

```{r metrics_crossref}
#| echo: false

## ── Small Variant Metrics ──
cat("=== Small Variant Metrics (hap.py, PASS, Subtype=*, Subset=*) ===\n\n")

smvar_metrics <- smvar_all_df %>%
  filter(context_label %in% c("SNV", "INDEL")) %>%
  select(bench_version, context_label, recall, precision, f1, tp, fn, fp)
cat("-- Overall by type --\n")
print(as.data.frame(smvar_metrics), row.names = FALSE)

cat("\n-- INDEL per-context --\n")
ctx_metrics <- smvar_all_df %>%
  filter(!context_label %in% c("SNV", "INDEL", "Overall"), !is.na(context_label)) %>%
  select(bench_version, context_label, recall, precision, f1, tp, fn, fp)
print(as.data.frame(ctx_metrics), row.names = FALSE)

cat("\n-- By size bin --\n")
print(as.data.frame(smvar_size_df %>% select(bench_version, size_bin, recall, precision, tp, fn, fp)),
      row.names = FALSE)

## ── SV Metrics ──
cat("\n\n=== Structural Variant Metrics (truvari bench+refine) ===\n\n")

cat("-- Overall + per-context (from stratified_metrics.csv) --\n")
stvar_print <- stvar_df %>%
  select(platform, callset, context_label, tp, fp, fn, recall, precision, f1)
print(as.data.frame(stvar_print), row.names = FALSE)

cat("\n-- By SV type (from svtype_metrics.csv) --\n")
svtype_print <- svtype_df %>%
  select(platform, callset, svtype, tp, fp, fn, recall, precision, f1)
print(as.data.frame(svtype_print), row.names = FALSE)
```
