---
title: "Benchmarking Use-Case: v4.2.1 vs v5.0q"
date: '`r Sys.Date()`'
author: "ND Olson"
format:
  html:
    theme: flatly
    page-layout: full
    toc: true
    self-contained: true
editor: source
code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false
source(here::here("analysis/_notebook_setup.R"))
analysis_setup(load_plot_themes = TRUE)
```

## Small Variant Benchmarking: Roche Neusomatic

Comparison of Roche Neusomatic (pFDA Truth Challenge V2) callset against
v4.2.1 and v5.0q small variant benchmarks on GRCh38, using GIAB v3.6
stratifications. Evaluated with `rtg vcfeval`.

```{r smvar_data}
smvar_df <- read_csv(
  here("results", "use_case", "smvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )
```

```{r smvar_plot, fig.width = 8, fig.height = 5}
smvar_long <- smvar_df %>%
  pivot_longer(
    cols = c(recall, precision),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = str_to_title(metric))

(smvar_plt <- smvar_long %>%
  ggplot(aes(
    x = context_label,
    y = value,
    fill = bench_version
  )) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7,
           color = "grey30", linewidth = 0.3) +
  geom_text(
    aes(label = sprintf("%.1f%%", value * 100)),
    position = position_dodge(width = 0.8),
    vjust = -0.3, size = 2.5
  ) +
  facet_wrap(~metric) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0.9, 1.005),
    oob = scales::squish,
    breaks = seq(0.90, 1.0, 0.02)
  ) +
  scale_benchmark_version(aesthetic = "fill") +
  labs(
    x = "Genomic Context",
    y = NULL,
    fill = "Benchmark Version"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -30, hjust = 0),
    legend.position = "bottom"
  ))

ggsave(
  filename = here("manuscript", "figs", "use_case_smvar.png"),
  plot = smvar_plt,
  width = 8,
  height = 5,
  dpi = 300
)
```

## Structural Variant Benchmarking: Baylor ONT vs Illumina DRAGEN

Comparison of Baylor ONT (Sniffles2) and Illumina DRAGEN SV callsets
against the v5.0q GRCh38 structural variant benchmark.
Evaluated with `truvari bench`.

```{r stvar_data}
stvar_df <- read_csv(
  here("results", "use_case", "stvar", "stratified_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    context_label = case_when(
      context == "Overall" ~ "Overall",
      .default = get_context_labels()[context]
    ),
    context_label = factor(
      context_label,
      levels = c("Overall", "Homopolymers", "Low Mappability",
                 "Segmental Duplications", "Tandem Repeats")
    ),
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )

svtype_df <- read_csv(
  here("results", "use_case", "stvar", "svtype_metrics.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    callset_label = case_when(
      callset == "baylor_ont" ~ "Baylor ONT (Sniffles2)",
      callset == "dragen" ~ "Illumina DRAGEN"
    ),
    f1 = 2 * (precision * recall) / (precision + recall)
  )
```

### By Genomic Context

```{r stvar_context_plot, fig.width = 8, fig.height = 5}
stvar_long <- stvar_df %>%
  pivot_longer(
    cols = c(recall, precision),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = str_to_title(metric))

(stvar_ctx_plt <- stvar_long %>%
  ggplot(aes(
    x = context_label,
    y = value,
    fill = callset_label
  )) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7,
           color = "grey30", linewidth = 0.3) +
  geom_text(
    aes(label = sprintf("%.1f%%", value * 100)),
    position = position_dodge(width = 0.8),
    vjust = -0.3, size = 2.5
  ) +
  facet_wrap(~metric) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1.05),
    breaks = seq(0, 1.0, 0.2)
  ) +
  scale_fill_manual(
    values = c("Baylor ONT (Sniffles2)" = "#2ca02c",
               "Illumina DRAGEN" = "#1f77b4")
  ) +
  labs(
    x = "Genomic Context",
    y = NULL,
    fill = "Callset"
  ) +
  theme_manuscript() +
  theme(
    axis.text.x = element_text(angle = -30, hjust = 0),
    legend.position = "bottom"
  ))
```

### By SV Type

```{r stvar_svtype_plot, fig.width = 6, fig.height = 4}
svtype_long <- svtype_df %>%
  pivot_longer(
    cols = c(recall, precision),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = str_to_title(metric))

(stvar_type_plt <- svtype_long %>%
  ggplot(aes(
    x = svtype,
    y = value,
    fill = callset_label
  )) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7,
           color = "grey30", linewidth = 0.3) +
  geom_text(
    aes(label = sprintf("%.1f%%", value * 100)),
    position = position_dodge(width = 0.8),
    vjust = -0.3, size = 2.5
  ) +
  facet_wrap(~metric) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1.05),
    breaks = seq(0, 1.0, 0.2)
  ) +
  scale_fill_manual(
    values = c("Baylor ONT (Sniffles2)" = "#2ca02c",
               "Illumina DRAGEN" = "#1f77b4")
  ) +
  labs(
    x = "SV Type",
    y = NULL,
    fill = "Callset"
  ) +
  theme_manuscript() +
  theme(legend.position = "bottom")
)
```

### Combined SV Figure

```{r combined_stvar, fig.width = 8, fig.height = 8}
(combined_stvar_plt <- stvar_ctx_plt / stvar_type_plt +
  plot_annotation(tag_levels = "A") &
  theme(plot.margin = unit(c(5, 5, 5, 5), "pt")))

ggsave(
  filename = here("manuscript", "figs", "use_case_stvar.png"),
  plot = combined_stvar_plt,
  width = 8,
  height = 8,
  dpi = 300
)
```
